---
<<<<<<< HEAD
title: "Challenge 1 submission"
author: "Cameron Needels"
desription: "Reading in data and creating a post"
date: "02/22/2023"
=======
title: "Challenge 1 Instructions"
author: "Meredith Rolfe"
description: "Reading in data and creating a post"
date: "08/15/2022"
>>>>>>> 0d022ed06e5927a734103bd13cbe794fa2b1c08a
format:
  html:
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
  - challenge_1
  - wildbirds
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)
library(summarytools)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Challenge Overview

Today's challenge is to

1)  read in a dataset, and

2)  describe the dataset using both words and any supporting information (e.g., tables, etc)

## Read in the Data

Read in one (or more) of the following data sets, using the correct R package and command.

-   railroad_2012_clean_county.csv ⭐
-   birds.csv ⭐⭐
-   FAOstat\*.csv ⭐⭐
-   wild_bird_data.xlsx ⭐⭐⭐
-   StateCounty2012.xls ⭐⭐⭐⭐

Find the `_data` folder, located inside the `posts` folder. Then you can read in the data, using either one of the `readr` standard tidy read commands, or a specialized package such as `readxl`.

```{r} 
library(readxl)
wild_bird_data <- read_excel("B:/Needels/Documents/DACCS 601/601_Spring_2023/posts/_data/wild_bird_data.xlsx",
                            skip = 2,
                            col_names = c("weight", "pop_size"))
wild_bird_data
```

```{r} 
dim(wild_bird_data)
```

```{r}
colnames(wild_bird_data)
```

```{r}
dfSummary(wild_bird_data)
```


## Data description

The data set consists of 2 columns, water weight in grams and the population size of wild birds.

The minimum weight is 5.5g , the median is 69.2g, and the maximum is 9639.8g.




The data summary shows the median weight amongst all the birds is 363.7 grams with a standard deviation of 983.5.
```{r}
#| label: summary

```

