---
title: "Challenge 6"
author: "Connor Landreth"
description: "Visualizing Time and Relationships"
date: "2023-04-03"
format:
  html:
    toc: true
    code-copy: true
    code-tools: true
categories:
  - challenge_6
  
---
```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)
library(ggplot2)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Challenge Overview

Today's challenge is to:

1)  read in a data set, and describe the data set using both words and any supporting information (e.g., tables, etc)
2)  tidy data (as needed, including sanity checks)
3)  mutate variables as needed (including sanity checks)
4)  create at least one graph including time (evolution)
   - try to make them "publication" ready (optional)
   - Explain why you choose the specific graph type
5)  Create at least one graph depicting part-whole or flow relationships
   - try to make them "publication" ready (optional)
   - Explain why you choose the specific graph type

[R Graph Gallery](https://r-graph-gallery.com/) is a good starting point for thinking about what information is conveyed in standard graph types, and includes example R code.

## Read in data

Read in one (or more) of the following datasets, using the correct R package and command.

  - debt  ⭐
  - fed_rate ⭐⭐
  - abc_poll ⭐⭐⭐
  - usa_hh ⭐⭐⭐
  - hotel_bookings ⭐⭐⭐⭐
  - AB_NYC ⭐⭐⭐⭐⭐

  

poll <- read.csv("abc_poll_2021.csv")

```{r}
head(poll)
View(poll)

```

### Briefly describe the data

## Tidy Data (as needed)

First off, below I will rename all the columns to actually be usable. Some of them such as ppmsacat aren't impossible to identify, but the layman would have no idea what that means. Additionally, at first glance, this data is 31 variables and 527 observations. I would be inclined to assume those behind the server seek to identify trends in political parties. (eg do republicans have less education? Do democrats have higher income?, Do independents live in the suburbs with big families or in cities alone?, etc..)
```{r}
Poll <- poll %>% 
  rename(ID = 'id', Degree = 'ppeduc5', Primary_Language  = 'xspanish', Page = 'ppage', Education_Level = 'ppeducat', Gender = 'ppgender', Household_Size = 'pphhsize', Ethnicity = 'ppethm', Income = 'ppinc7', Marital_Status = 'ppmarit5', Metro_Stat_Area ='ppmsacat', Region = 'ppreg4', Rental_Status = 'pprent', State = 'ppstaten', Retired ='PPWORKA', Employment_Status = 'ppemploy', Complete_Status = 'complete_status', Political_Affiliation = 'QPID', Age = 'ABCAGE', Interview = 'Contact')

View(Poll)


```

Next, I will remove some columns that won't be helpful for this analysis

```{r}
Poll$Page <- NULL
Poll$ID <- NULL
Poll$Complete_Status <- NULL
Poll$weights_pid <- NULL

# My analysis will look at the working class, so I will filter out any retired individuals

Poll_re <- Poll %>% 
  select(Retired, Employment_Status, Education_Level, Income, Household_Size, Ethnicity, Metro_Stat_Area, Region, Political_Affiliation, Age) %>%
  filter(Retired != 'Retired')
 
View(Poll_re)
  


```

I want to look at the role education and employment status might play in political affiliation. I am going to hypothesize non-retired, unemployed individuals are less educated, but will explore this below and see where their political affiliation falls.

```{r}

```


## Visualizing Part-Whole Relationships
