---
title: "Final Project Assignment#1: Cam Needels"
author: "Cam Needels"
description: "Project & Data Description"
date: "05/3/2023"
format:
  html:
    df-print: paged
    toc: true
    code-copy: true
    code-tools: true
    css: styles.css
categories:
  - final_Project_assignment_1
  - final_project_data_description
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)
library(readr)
library(ggplot2)
library(summarytools)
library(lubridate)
library(GGally)
library(dplyr)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Part 1. Introduction {#describe-the-data-sets}

1. This data set is trending YouTube videos from August 11th, 2020 until April 10th, 2023 in the United States. It's collected from the top trending YouTube video list provided by YouTube itself. This is created in order to find the top-trending videos that drive the most traffic to the website. Since YouTube has so much data from all over the world and because different videos trend in different countries, this data set is specifically the US data. Each case represents a video that is trending on a certain date (this means the same video can be present multiple times because it can trend for multiple dates). 

2. Which topics have the most amount of videos trending? Which variables strongly relate to higher view count? Which channels trended the most? Which videos were the most viewed in this time period?

## Part 2. Describe the data set(s) {#describe-the-data-sets-1}

As stated above this data set takes the trending YouTube videos from August 11th, 2020 until April 10th, 2023 in the US. It contains 195390 trending videos on different dates (with some duplicates). There are 36572 unique videos in this data set. This data set has 16 columns including the view count, title, publish date, channel name, category id (instead of the name it makes each category a specific number, this actually makes it easier for my statistical analysis section), youtube channel id, likes, dislikes, comment count, the date the video was trending, the tags on the YouTube video, the thumbnail link, if comments were disabled or not, if ratings were disabled or not, and description. This is a large amount of data and I'm considering getting rid of duplicates to make it more manageable but I haven't decided if that's the route I want to go. I'm interested in mostly the comment_count, dislikes, likes, and view_count as those are the most important variables for YouTube videos.

```{r}
YouTube <- read.csv("B:/Needels/Documents/DACCS 601/DACSS_601_New/posts/CamNeedels_FinalProjectData/US_youtube_trending_data.csv")
YouTube
```


```{r}
#the dimensions
dim(YouTube)
```


```{r}
#the amount of unique videos
length(unique(YouTube$title))
```


```{r}
descr(YouTube)
```


## 3. The Tentative Plan for Visualization {#the-tentative-plan-for-visualization}


1. I plan on using a heat map to show how closely related the view count is to the comment_count, dislikes, likes. I want to see which of those have the greatest correlation. We haven't learned how to make a heat map, but I'll do some research and figure it out because that seems to make the most sense. If I leave the variables as is and make essentially a battleship map of how much they relate on a scale from 0% to 100%, depending on the color. For all the other questions I will most likely use a bar graph to show the highest amount of views and take only the top ten sample for the trending topics, channels, and videos. Such as the video on the x axis and the view count on the y axis to figure out the trending videos. I will most likely need to lubridate the current dates that are inserted so they can be created separately and fortunately for me there are no missing data points. I also plan on using the uniqe function so I don't get repeats of the same video in the top 10 because I feel like it's redundant to have the same video twice in that analysis.


```{r}
YouTube <- mutate(YouTube, category = recode(categoryId, '1' = "Film/Animation", '2' = "Auto/Vehicles", '10' = "Music", '15' = "Pets/Animals", '17' = "Sports", '18' = "Short_Movies", '19' = "Travel/Events", '20' = "Gaming", '21' = "Videoblogging", '22' = "People/Blogs", '23' = "Comedy", '24' = "Entertainment", '25' = "News/Politics", '26' = "Howto/Style", '27' = "Education", '28' = "Science/Technology", '29' = "Nonprofits/Activism", '30' = "Movies", '31' = "Anime/Animation", '32' = "Action/Adventure", '33' = "Classics", '34' = "Comedy", '35' = "Documentary", '36' = "Drama", '37' = "Family", '38' = "Foreign", '39' = "Horror", '40' = "SciFi/Fantasy", '41' = "Thriller", '42' = "Shorts", '43' = "Shows", '44' = "Trailers"))
```

```{r}
YouTube%>%
  ggplot(aes(category, fill = category)) +
  geom_bar()+
  coord_flip()

```



```{r}
#the correlation matrix being created and covering all variables and making it in a pearson style via color
ggcorr(select(YouTube, -categoryId) , method = c("everything", "pearson"))
```

```{r}
YouTube%>%
  arrange(desc(view_count))%>%
  slice(1:30)%>%
  ggplot(aes(y = view_count, x = reorder(title, -view_count), fill = trending_date)) + 
  geom_bar(position = "stack", stat = "identity") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

```{r}
#line graph x = trending date, y = view_count, color = channel. group_by(channel)  mutate(maxview= Max(view_count)) then slice
YouTube%>%
  arrange(desc(view_count))%>%
  slice(1:30)%>%
  ggplot(aes(y = view_count, x = channelTitle, fill = trending_date)) + 
  geom_bar(position = "stack", stat = "identity") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```


## 4. Analysis

From the heat map I gathered that the amount of likes and have the highest correlation to each other.The comment count is also the most related variable to view count which also makes sense because likes and comments tell the YouTube algorithm to recommend certain videos more. The biggest lack of indicator for views is dislikes as well. One surprising thing is that likes and comment count have a strong correlation which I find facinating.

From both the video trends and channel trends graphs I found that BLACKPINK (a music group) had the most amount of views in the time period. It trended as a channel and it's video Pink Venom trended 17 days. It was crazy to see that when I let the values only have the top 10 videos, it was 9 of the top 10 trending videos for different days.

The bar graph with the different categories showed that Entertainment had the most amount of trending videos with Gaming in Second and Music in Third. I was going to anticipate Music as first because so many of the top trending videos was music. In hindsight it makes sense because music tends to be replayed giving it more replayable value. However making music takes a lot of time as opposed to gaming and entertainment videos (which can take a while) but are easier to create and provide lots of value. 
