---
title: "Challenge 1"
author: "Abhinav Reddy Yadatha"
description: "Reading in data and creating a post"
date: "02/26/2023"
format:
  html:
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
  - challenge_1
  - my name
  - dataset
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Challenge Overview

Today's challenge is to

1)  read in a dataset, and

2)  describe the dataset using both words and any supporting information (e.g., tables, etc)

## Read in the Data

Read in one (or more) of the following data sets, using the correct R package and command.

-   railroad_2012_clean_county.csv ⭐
-   birds.csv ⭐⭐
-   FAOstat\*.csv ⭐⭐
-   wild_bird_data.xlsx ⭐⭐⭐
-   StateCounty2012.xls ⭐⭐⭐⭐

Find the `_data` folder, located inside the `posts` folder. Then you can read in the data, using either one of the `readr` standard tidy read commands, or a specialized package such as `readxl`.

```{r}
dataframe <- read_csv('_data/birds.csv', show_col_types = FALSE)
head(dataframe)
```

Add any comments or documentation as needed. More challenging data sets may require additional code chunks and documentation.

## Describe the data

Using a combination of words and results of R commands, can you provide a high level description of the data? Describe as efficiently as possible where/how the data was (likely) gathered, indicate the cases and variables (both the interpretation and any details you deem useful to the reader to fully understand your chosen data).

Description : The dataset 'birds.csv' contains information about the population of wild birds like chicken, geese etc for a few countries anually from 1961 to 2018.

```{r}

#| Displaying the first few rows.
head(dataframe)

```

Displaying the summary of the dataset.

```{r}
#| label: summary

library(summarytools)
dfSummary(dataframe)
```

Checking the dimensions of the dataset:

```{r}
dim(dataframe)
#| 
```

It can be observed that there are 30977 rows and 14 columns

Displaying the column names of the dataset :

```{r}
colnames(dataframe)
#| The dataset has 14 coulmns describing various fields such as above.
```

Number of unique years :

```{r}


unique_years <- dataframe%>% select(Year)%>% n_distinct(.)
unique_years

```

It can be observed that the dataset has the data for 58 unique years i.e 1961-2018

Number of unique wildbirds :

```{r}

unique_birds <- dataframe%>% select(Item)%>% n_distinct(.)
unique_birds

```

It can be observed that the dataset contains information about 5 different types of birds.

Number of unique Areas / countries:

```{r}
unique_areas <- dataframe%>% select(Area)%>% n_distinct(.)
unique_areas

```

It can be observed that the dataset contains information about 248 different areas / countires
