---
title: "Final Project Assignment#1: Aditya Salveru"
author: "Aditya Salveru"
description: "Project & Data Description"
date: "05/10/2023"
format:
  html:
    df-print: paged
    toc: true
    code-copy: true
    code-tools: true
    css: styles.css
categories:
  - final_Project_assignment_1
  - Behavioral Risk Factor Surveillance System (BRFSS)
editor_options: 
  chunk_output_type: console
---

<!-- ## Important Formatting & Submission Notes: {#important-formatting-submission-notes} -->

<!-- 1.  **Use this file as the template to work on: start your own writing from Section "Part.1"** -->

<!-- 2.  **Please make the following changes to the above YAML header:** -->

<!--     -   Change the "title" to "Final Project Assignment#1: First Name Last Name"; -->

<!--     -   Change the "author" to your name; -->

<!--     -   Change the "date" to the current date in the "MM-DD-YYYY" format; -->

<!-- 3.  **Submission:** -->

<!--     -   Delete the unnecessary sections ("Overview", "Tasks", "Special Note", and "Evaluation"). -->
<!--     -   **In the posts folder of your local 601_Spring_2023 project, create a folder named "FirstNameLastName_FinalProjectData", and save your final project dataset(s) in this folder. DO NOT save the dataset(s) to the \_data folder which stores the dataset(s) for challenges.** -->
<!--     -   Render and submit the file to the blog post like a regular challenge. -->

<!-- ## Overview of the Final Project {#overview-of-the-final-project} -->

<!-- The goal is to tell a coherent and focused story with your data, which answers a question (or questions) that a researcher, or current or future employer, might want to have answered. The goal might be to understand a source of covariance, make a recommendation, or understand change over time. We don't expect you to reach a definitive conclusion in this analysis. Still, you are expected to tell a data-driven story using evidence to support the claims you are making on the basis of the exploratory analysis conducted over the past term. -->

<!-- In this final project, statistical analyses **are not required**, but any students who wish to include these may do so. However, **your primary analysis should center around visualization rather than inferential statistics.** Many scientists only compute statistics after a careful process of exploratory data analysis and data visualization. Statistics are a way to gauge your certainty in your results - **NOT A WAY TO DISCOVER MEANINGFUL DATA PATTERNS**. **Do not run a multiple regression with numerous predictors and report which predictors are significant!!** -->

<!-- ## Tasks of Assignment#1 {#Tasks-of-assignment1} -->

<!-- This assignment is the first component of your final project. Together with the later assignments, it make up a short paper/report. In this assignment, you should introduce a dataset(s) and how you plan to present this dataset(s). This assignment should include the following components: -->

<!-- 1.  **A clear description of the dataset(s) that you are using.** -->

<!-- 2.  **What "story" do you want to present to the audience? In other words, what "question(s)" do you like to answer with this dataset(s)?** -->

<!-- 3.  **The Plan for Further Analysis and Visualization.** -->

<!-- We will have a special class meeting on April 12 to review and discuss students' proposed datasets for the final project. **If you want your project being discussed in the class, please submit this assignment before April 12.** -->

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Part 1. Introduction {#describe-the-data-sets}

My projects aims to analyze the heart disease data and explore the causes and general trends in the individuals with and without the heart disease. Particularly I am interested in finding the factors that contribute to the heart disease the most and early signs that might lead to a heart disease. I would also like to analyze the healthy habits that would keep the heart diseases away.


### 1.  Dataset(s) Introduction:

I have chosedn the 2020 annual CDC survey data of 400k adults related to their health status collected by Behavioral Risk Factor Surveillance System (BRFSS). For people of the majority of racial groups in the US (African Americans, American Indians and Alaska Natives, and white people), heart disease is one of the leading causes of death, according to the CDC. High blood pressure, high cholesterol, and smoking are the three main risk factors for heart disease that at least half of all Americans (47%) have. Other key factors are diabetic status, obesity (high BMI), not obtaining enough physical activity or drinking too much alcohol. In the field of medicine, it is crucial to identify and combat the causes that have the biggest influence on heart disease.

The Behavioral Risk Factor Surveillance System (BRFSS) is a collaborative project between all the states in the
United States and participating US territories and the Centers for Disease Control and Prevention (CDC). The
BRFSS is administered and supported by CDC's Population Health Surveillance Branch, under the Division of
Population Health at CDC’s National Center for Chronic Disease Prevention and Health Promotion. The
BRFSS is a system of ongoing health-related telephone surveys designed to collect data on health-related risk
behaviors, chronic health conditions, health-care access, and use of preventive services from the
non institutionalized adult population (≥ 18 years) residing in the United States and participating areas.


The original dataset contains over 300 columns and over 400k individual's responses as rows. However, I have chosen a condensed version of the dataset available at Kaggle which has the attributes that we are really interested in and ignoring the rest of the columns.


### 2.  What questions do you like to answer with this dataset(s)?

- How is heart disease related to lifesyle habits such as sleep time, alcohol consumption, BMI etc.
    
- Is there any particular group of people(Race, region) who are at high risk compared to the other people
    

## Part 2. Describe the data set(s) {#describe-the-data-sets-1}

<!-- This part contains both a coding and a storytelling component. -->

<!-- In the coding component, you should: -->

1.  read the dataset;

    <!-- -   (optional) If you have multiple dataset(s) you want to work with, you should combine these datasets at this step. -->

    <!-- -   (optional) If your dataset is too big (for example, it contains too many variables/columns that may not be useful for your analysis), you may want to subset the data just to include the necessary variables/columns. -->

```{r}
data <- read.csv('./AdityaSalveru_FinalProjectData/heart_2020.csv')

data %>%
  colnames() %>%
  sort()
```



The following are the coulmns in the dataset:


- AgeCategory : Age category of the subject(divided into 14 categories)

- AlcoholDrinking : Is the subject a heavy drinker(Men > 14 drinks/Week, Women > 7 drinks/Week)

- Asthma: Does the subject have asthma

- BMI: Body mass index of the subject

- Diabetic: Is the subject diabetic

- DiffWalking: Does the subject feel difficulty walking

- GenHealth: What would the subject perceive his general health as?

- HeartDisease: Does the Subject has coronary heart disease (CHD) or myocardial infarction (MI).

- KidneyDisease: Does the subject have a kidney disease excluding stones, bladder infection.

- MentalHealth: How many days in the past 30 days is the subject mentally ill.

- PhysicalActivity: Has the subject 

- PhysicalHealth:How many days in the past 30 days is the subject physically ill.

- Race: Race/ethenicity of the subject

- Sex: Is the subject male or female

- SkinCancer:Has the subject ever had skin cancer

- SleepTime: On average how many hours does the subject sleep everyday.

- Smoking: Does the subject smoke( atleast 100 cigaretts in the past)

- Stroke: Has the subject ever had a stroke


```{r}
summary(data)
```

#### Distribution of Male and Female subjects for the data
```{r}
# Create a table with the number of males and females
sex_table <- data %>%
  select(Sex) %>%
  table()

# Calculate the percentage of males and females
sex_percent <- round(prop.table(sex_table)*100)

# Create a bar plot of the sex data with percentages
barplot(sex_table, main="Number of Males and Females",
        xlab="Sex", ylab="Frequency", col=c("pink", "lightblue"),
        legend=c("Female", "Male"), ylim=c(0, max(sex_table)*1.2))
text(1:2, sex_table + 0.2, paste0(sex_percent, "%"), col = "black", cex = 1.5)

```

#### Distribution of Subjects across age groups
```{r}
age_table <- data %>%
  select(AgeCategory) %>%
  table()

# Add percentages to the pie chart
percent <- round(100*age_table/sum(age_table), 2)
percent_str <- paste(percent, "%", sep="")
labels <- paste(names(age_table), " (", percent_str, ")", sep="")
pie(age_table, main="Age Category Distribution", labels=labels, col=rainbow(length(age_table)))


```


#### Distribution of Subjects across age groups

```{r}
ggplot(data, aes(x = Race)) +
  geom_bar(fill = "lightblue") +
  xlab("Race") +
  ylab("Count") +
  ggtitle("Count of Race") +
  geom_text(stat='count', aes(label=..count..), vjust=-0.5) 
```

#### Overview of BMI of the individuals as a Histogram

```{r}
# create a histogram of the BMI data
ggplot(data = data, aes(x = BMI)) +
  geom_histogram(binwidth = 5, fill = "darkgray", color = "white") +
  labs(title = "BMI Distribution", x = "BMI", y = "Count")

```


<!-- 2.  present the descriptive information of the dataset(s) using the functions in Challenges 1, 2, and 3; -->

<!--     -   for examples: dim(), length(unique()), head(); -->

<!--     ```{r} -->

<!--     ``` -->

<!-- 3.  conduct summary statistics of the dataset(s); especially show the basic statistics (min, max, mean, median, etc.) for the variables you are interested in. -->

<!-- ```{r} -->

<!-- ``` -->

<!-- In the storytelling component, you should describe the basic information of the dataset(s) and the variables in a way that corresponds to your descriptive and summary statistics in the above coding component. **DO NOT** simply report the number of rows. Instead, describe the dataset(s) fully by **specifying what each row and column mean.** In other words, your description should be comprehensive and detailed enough for readers to picture or envision the dataset(s) in their brains. -->

<!-- -   For example, suppose I use a dataset of all the athletes who participated in the Olympic Games. Here is how I describe the basic information of the data: "the case of this dataset is ab individual athlete, represented by each row in the dataset. The dataset includes individual (e.g., gender, age, height, weight, race) and event performance (e.g., final placement) information for all athletes (22,398) competing in all events (e.g., Male 400m Free, Female ...) in all Olympics Games since 1922 (24 Winter and 28 Summer Games. Athletes appearing in the dataset competed in anywhere from 1-11 distinct events (of 198 possible) during 1-5 distinct Olympic competitions, for a total of XXX, XXX athlete-event-Olympic-year observations. XXX Countries are represented, etc)." -->

<!-- -   Erico's hint: as I mentioned above, sometimes a dataset is too large, and it is difficult to present and explain all the variables/columns (especially if you run summary statistics for the whole dataset). In this case, you will have to make a decision to select the most important variables/columns to discuss. For example, the Olympic dataset I mentioned above as an example contains more than 50 columns. For clarity of data presentation, I may just focus on 6 items/columns of individual athletes (gender, age, weight, height, race, nationality) and the column of final placement that are most relevant to answer my specific research questions. By doing so, you can just present the tables of the summary statistics of these 7 variables/columns without showing too much information and confusing the readers. -->

<!-- -   A good example is to can check out [the Data Description section of the above student's final project](https://dacss.github.io/601_Fall_2022_final_posts/posts/final_project_Guanhua_Tan.html#briefly-describe-the-dataset.). As you can see, the student describes the dataset after he runs a few descriptive statistics. You can also see the weekly challenge solutions by Professor Rolfe for other examples of clear, concise data descriptions. -->

## 3. The Tentative Plan for Visualization {#the-tentative-plan-for-visualization}

<!-- 1.  Briefly describe what data analyses (**please the special note on statistics in the next section)** and visualizations you plan to conduct to answer the research questions you proposed above. -->

One of the main factor that I would to study the relationship with heart disease the BMI of the people. Therefore I will create a box plot to analyze the BMIs of the people with and without heart disease. This Box plot would give an idea on how the BMIs of the people are distributed across the two classes of the people(with and without heart disease). I chose this box plot because the BMIs are continous numbers.


I would also like to study how age effects the possiblilt of heart diseases. specifically we need to find out what what is the age distribution of the people with heart disease and are there any age groups that are more likely to have heart disease as compared to other groups. Therefore I will be using histograms of number of people with heart disease with different ages.

One more important relation that I would like to study is how likely are people who smoke are to get a Heart Disease as compared to the people who does not smoke. For this I would like to careate a stacked bar plot with fill as Heart Disease. This we we can compare the percentage of people with heart disease for both the bars(Yes and No) for smoking.


<!-- 2.  Explain why you choose to conduct these specific data analyses and visualizations. In other words, how do such types of statistics or graphs (see [the R Gallery](https://r-graph-gallery.com/)) help you answer specific questions? For example, how can a bivariate visualization reveal the relationship between two variables, or how does a linear graph of variables over time present the pattern of development? -->

<!-- 3.  If you plan to conduct specific data analyses and visualizations, describe how do you need to process and prepare the tidy data. -->

<!--     -   What do you need to do to mutate the datasets (convert date data, create a new variable, pivot the data format, etc.)? -->

<!--     -   How are you going to deal with the missing data/NAs and outliers? And why do you choose this way to deal with NAs? -->

<!-- 4.  (Optional) It is encouraged, **but optional**, to include a coding component of tidy data in this part. -->



```{r}

```

<!-- ## Special Note on the role of statistics {#special-note-on-the-role-of-statistics} -->

<!-- Statistical analyses **are not required**, but any students who wish to include these may do so. However, your primary analysis should center around visualization rather than inferential statistics. Many scientists only compute statistics after a careful process of exploratory data analysis and data visualization. Statistics are a way to gauge your certainty in your results - **NOT A WAY TO DISCOVER MEANINGFUL DATA PATTERNS**. Do not run a multiple regression with numerous predictors and report which predictors are significant!! -->

<!-- **Remember: The goal is to tell a story about the data**. For example, you might identify sports where winning athletes are younger or older than average and then try to see if you can find some sort of pattern that accounts for this difference. Or perhaps you might compare a country's performance to GDP and see if this changes over time. The goal is not to get a significant statistical result but to identify an interesting pattern in the data and then extract some sort of meaningful recommendation or information from it. -->

<!-- ## Evaluation {#evaluation} -->

<!-- You will be evaluated on both the quality of your source code and your written report, with a greater emphasis on the clarity and details of the description of your dataset(s) and your research questions. -->
