---
title: "Challenge 8 Instructions"
author: "Anirudh Lakkaraju"
description: "Joining Data"
date: "05/15/2023"
format:
  html:
    toc: true
    code-copy: true
    code-tools: true
categories:
  - challenge_8
  - SNL
  - Anirudh Lakkaraju
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)
library(ggplot2)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Challenge Overview

Today's challenge is to:

1)  read in multiple data sets, and describe the data set using both words and any supporting information (e.g., tables, etc)
2)  tidy data (as needed, including sanity checks)
3)  mutate variables as needed (including sanity checks)
4)  join two or more data sets and analyze some aspect of the joined data

(be sure to only include the category tags for the data you use!)

## Read in data

Read in one (or more) of the following datasets, using the correct R package and command.

-   military marriages ⭐⭐
-   faostat ⭐⭐
-   railroads ⭐⭐⭐
-   fed_rate ⭐⭐⭐
-   debt ⭐⭐⭐
-   us_hh ⭐⭐⭐⭐
-   snl ⭐⭐⭐⭐⭐

```{r}
seasons_df <- read.csv("_data/snl_seasons.csv")
actors_df <- read.csv("_data/snl_actors.csv")
casts_df <- read.csv("_data/snl_casts.csv")

# Displaying first few rows of the datasets
head(seasons_df) 
```

```{r}
head(actors_df)
```

```{r}
head(casts_df)
```

```{r}
# Dimensions of datasets
dim(seasons_df)
```

```{r}
dim(actors_df)
```

```{r}
dim(casts_df)
```

```{r}
summary(actors_df)
```

```{r}
summary(seasons_df)
```

```{r}
summary(casts_df)
```

### Briefly describe the data

\
The data frames contain organized and independent data. The actors Data Frame includes a comprehensive list of actors, guests, musical guests, and crew members from the show, with each entry representing an individual actor and providing details about their role type and gender.

The casts data frame focuses on actors who were part of a specific season's cast. Each entry represents an actor and includes information like their featured status, the dates of their first and last episodes, whether they served as an anchor on weekend update, and the number of episodes they participated in during that season.

The seasons data frame provides information about each season individually. Each entry corresponds to a specific season and includes data such as the airing year, the dates of the first and last episodes, and the total number of episodes.

## Tidy Data (as needed)

```{r}

actors_df <- actors_df %>%
  drop_na()
```

```{r}
seasons_df <- seasons_df %>%
  drop_na()
```

```{r}
casts_df <- casts_df %>%
  drop_na()
```

```{r}
colnames(actors_df)
```

```{r}
colnames(casts_df)
```

```{r}
colnames(seasons_df)
```

```{r}
actors_df <- actors_df %>%
  mutate(appearances = rowSums(select(., starts_with("ep_")), na.rm = TRUE))
```

## Join Data

```{r}
df_casts_actors <- casts_df %>%
  left_join(actors_df, by = "aid") %>%
  select(sid, type, gender, featured, appearances)

```

```{r}
df_data <- seasons_df %>%
  left_join(df_casts_actors, by = "sid")
```

```{r}
colnames(df_data)
```

```{r}
colnames(df_casts_actors)
```

```{r}
joined_data <- casts_df %>%
  select(-update_anchor, -season_fraction, -n_episodes) %>%
  left_join(seasons_df, by="sid")

joined_data
```

```{r}
colnames(joined_data)
```
