---
title: "Challenge 3"
author: "Anirudh Lakkaraju"
description: "Tidy Data: Pivoting"
date: "05/02/2023"
format:
  html:
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
  - challenge_3
  - Anirudh Lakkaraju
  - animal_weights
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

### Reading the data

```{r}
df <- read_csv("_data/animal_weight.csv")
```

```{r}
head(df)
```

### Briefly describe the data

Find the number of rows and cols of the dataset

```{r}
nrow(df)
```

```{r}
ncol(df)
```

```{r}
summary(df)
```

The dataset contains animal weight data, which indicates the number of various types of livestock (such as buffaloes, chickens, and turkeys) in different regions. The dataset has 9 rows and 17 columns, but the 17 columns make the data difficult to handle or analyze. To make the data more efficient, we can utilize the "pivot_longer()" function to restructure it. This will transform the 17 columns into three columns, including region, cattle type, and weight, resulting in 144 rows (9 rows x 16 variables) of data.

### Pivoting the data

```{r}
data_longer<-pivot_longer(df, col=-`IPCC Area`,
                                    names_to = "Livestock",
                                    values_to = "Weight")
print(data_longer)

```

Dimensions of restructured data

```{r}
dim(data_longer)
```

As expected, using pivot_longer(), we get a dataset with 144 rows and 3 cols.
