---
title: "Challenge 7"
author: "Anirudh Lakkaraju"
description: "Visualizing Multiple Dimensions"
date: "05/16/2023"
format:
  html:
    toc: true
    code-copy: true
    code-tools: true
categories:
  - challenge_7
  - air_bnb
  - Anirudh Lakkaraju
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)
library(ggplot2)
library(dplyr)
library(lubridate)
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Read in data

```{r}
df <- read.csv("_data/AB_NYC_2019.csv")
head(df)
```

```{r}
dim(df)
```

```{r}
summary(df)
```

### Briefly describe the data

The dataset contains information about approximately 49,000 rental properties available on Airbnb in New York City during 2019. Each entry represents a unique rental unit and includes 16 different variables, including details like the unit's ID, name, location, host ID and name, room type, price, minimum nights required for booking, number of reviews, date of the last review, average monthly reviews, calculated count of host listings on Airbnb, and availability.

```{r}
str(df)

```

```{}
```

## Tidy Data (as needed)

```{r}
df <- na.omit(df)
df <- df %>% mutate(high_price = ifelse(price > 500, "High", "Low"))
```

## Visualization with Multiple Dimensions

Univariate visualization using histogram

```{r}
library(ggplot2)
ggplot(df, aes(x=price)) + 
  geom_histogram(binwidth=25, color="blue", fill="blue") +
  labs(title="Distribution of Price", x="Price", y="Frequency")
```

\
Bi-variate Visualizations :

The initial scatterplot illustrates the correlation between price and reviews_per_month within the complete dataset. The plot is created using the ggplot function, with the aes function specifying the variables for the x and y axes. Data points are then added to the plot using geom_point, and the title and axis labels are defined using labs.

```{r}
ggplot(df, aes(x=price, y=reviews_per_month)) +
  geom_point(alpha=0.46, color="red") +
  labs(title="Monthly Price vs Reviews", x="Price", y="Reviews per Month")
```

```{r}
ggplot(df, aes(x=price, y=reviews_per_month)) +
  geom_point(alpha=0.5, color="red") +
  labs(title="Price vs Reviews per Month by City", x="Price", y="Reviews per Month") +
  facet_wrap(~ neighbourhood_group, scales="free")
```

\
I opted for a scatter plot to visually represent the correlation between price and reviews per month across different cities. This choice was made because a scatter plot effectively displays individual data points as well as the general trend in the data. By utilizing this plot, we can observe whether there is a connection between price and reviews per month, as well as identify any outliers or patterns within the data. Additionally, the use of distinct colors to represent each city aids in distinguishing the data points associated with each specific location.
