---
title: "Challenge 1"
author: "FNU Avinesh Krishnan"
description: "Reading in data and creating a post"
date: "05/13/2023"
format:
  html:
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
  - challenge_1
  - FNU Avinesh Krishnan
  - wild_bird_data
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Challenge Overview

Today's challenge is to

1)  read in a dataset, and

2)  describe the dataset using both words and any supporting information (e.g., tables, etc)

## Read in the Data

Read in one (or more) of the following data sets, using the correct R package and command.

-   railroad_2012_clean_county.csv ⭐
-   birds.csv ⭐⭐
-   FAOstat\*.csv ⭐⭐
-   wild_bird_data.xlsx ⭐⭐⭐
-   StateCounty2012.xls ⭐⭐⭐⭐

Find the `_data` folder, located inside the `posts` folder. Then you can read in the data, using either one of the `readr` standard tidy read commands, or a specialized package such as `readxl`.

```{r}
 library(readxl)
 wild_birds <- read_csv("~/Desktop/601_Spring_2023/posts/_data/FAOstat_egg_chicken.csv")
 view(wild_birds)
```
Read the data from the csv file.

Add any comments or documentation as needed. More challenging data sets may require additional code chunks and documentation.

## Describe the data

Using a combination of words and results of R commands, can you provide a high level description of the data? Describe as efficiently as possible where/how the data was (likely) gathered, indicate the cases and variables (both the interpretation and any details you deem useful to the reader to fully understand your chosen data).


The given data appears to be related to Livestock Primary in different areas and several years. It contains information about the production of eggs, laying, yield, and their respective values in different units. The data also includes flags and flag descriptions for each entry. The columns contain information such as Domain Code, Domain, Area Code, Area, Element Code, Element, Item Code, Item, Year Code, Year, Unit, Value, Flag, and Flag Description. The data appears to have been collected by the Food and Agriculture Organization (FAO).


```{r}
#| label: summary
 head(wild_birds)
```

Look at the dataset's first six rows to get a sense of the type of data that is present.

```{r}
 dim(wild_birds)
```

Got an idea of number of observations taken on different fields.

```{r}
 colnames(wild_birds)
```

the dataset's column names.

```{r}
 filter(wild_birds, `Area`=="Afghanistan" & `Element`=="Laying")
```

Knowing the data of Afghanistan and the the Yield which is Laying.


```{r}
 wild_birds%>%
  select(`Area`) %>%
  n_distinct(.)
```

There are 245 distinct areas.

```{r}
 wild_birds%>%
  select(`Area`) %>%
  distinct(.)
```
The above ones are the distict area names.


```{r}
 count_flag<-wild_birds%>%
  filter(`Flag Description`=="Official data") %>% 
  count()
count_flag
```

There are 7548 data points where the data has a flag as official data.

```{r}
 summary(wild_birds)
```
A brief summary of the dataset.




