---
title: "Challenge 6 Submission"
author: "Cam Needels"
description: "Visualizing Time and Relationships"
date: "04/05/2023"
format:
  html:
    toc: true
    code-copy: true
    code-tools: true
categories:
  - challenge_6
  - CamNeedels
  - debt

---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)
library(ggplot2)
library(readxl)
library(summarytools)
library(lubridate)
library(here)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

```{r}
FedFunds <- read_csv("B:/Needels/Documents/DACCS 601/DACSS_601_New/posts/_data/FedFundsRate.csv")
FedFunds
```

### Briefly describe the data

This data spans from 1954 to 2017 and it includes variables that deal with federal funds rate or the amount of interest banks loan money to one another. 8 variables total in this data set.

```{r}
FedFundsTidy <- FedFunds %>%
  pivot_longer( cols= "Federal Funds Target Rate":"Inflation Rate",
                names_to = 'Economic Status',
                values_to = "Rates")
FedFundsTidy
```

```{r}
FedFundsDate <- FedFundsTidy%>%
  mutate( date = str_c(Month, Day, Year , sep = "-"),
          date = mdy(date))
FedFundsDate
```
We pivoted the data longer because it is cluttered and 3 of the variables have NA data so why not just pivot it longer. We will make a new column called rates and we also made the dates usable in the data.


```{r}

```



### Time Dependent Visualization

I wanted to make a line graph to show the trends of the economic status, but I would have also liked to add color to show the status at each time period. 

```{r}
FedFundsDate%>%
  ggplot(aes(x=date, y = Rates,)) +
  geom_line(size =.3)
```

I couldn't figure out how to make my colors show up for 'Economic Status' so this is as far as I ended up going.


## Visualizing Part-Whole Relationships
