---
title: "Challenge 3 Submission"
author: "Cam Needels"
description: "Tidy Data: Pivoting Featuring Eggs"
date: "08/17/2022"
format:
  html:
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
  - challenge_3
  - eggs
  - Cam Needels
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)
library(summarytools)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

```{r}
library(readr)
eggs_data <- read.csv("B:/Needels/Documents/DACCS 601/601_Spring_2023/posts/_data/eggs_tidy.csv")
eggs_data
```

### Describing the Data

This data is about the number of eggs coming from various dozen sizes. The dates vary from January 2004 until December 2013. This data looks solid however the extra large dozen column is all by itself and it isn't very tidy. In order to make it tidy I decided to make it long by having each category of dozen be a different data point so that extra large dozen can fit in.

### Data prediction


```{r}
#existing rows/cases
nrow(eggs_data)
#existing columns/cases
ncol(eggs_data)
#expected rows/cases
nrow(eggs_data) * (ncol(eggs_data)-2)
# expected columns 
2+2
```

With 120 rows in the current data set and 6 columns. 2  columns describe the case of the data set which include the month and year. The remainder of the 4 describe the carton and size of eggs. 

After creating the carton_type column and the price_per_pound column there are 4 columns total. We should now expect to see 480 columns total because the old 4 columns merged into new rows multiplying the 120 by four times.

## Pivot the Data

```{r}
eggs_data <-pivot_longer(eggs_data, col = c(large_half_dozen, large_dozen, extra_large_half_dozen, extra_large_dozen),
                         names_to= "carton_size",
                         values_to= "price_per_pound")
eggs_data
```

The pivoted data looks great and has the expected dimensions of 480 rows x 4 columns!



```{r}


```
