---
title: "Challenge 8 Submission"
author: "Cam Needels"
description: "Joining Data"
date: "05/01/2023"
format:
  html:
    toc: true
    code-copy: true
    code-tools: true
categories:
  - challenge_8
  - military
  - CamNeedels
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)
library(ggplot2)
library(readxl)
library(readr)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```



### Briefly describe the data

All these data sets are the amount of a certain item that are produced from livestock, dairy cattle, and eggs.

## Tidy Data (as needed)

Data is already tidy, all I needed to do is join it together. It makes it easy that they all have the same variables so merging them was pretty simple!
```{r}
#assigning the eggs data set to an object
Eggs <- read_csv(("B:/Needels/Documents/DACCS 601/DACSS_601_New/posts/_data/FAOSTAT_egg_chicken.csv"))
dim(Eggs)
Eggs
```

```{r}
#assigning the livestock data set to an object
livestock<- read_csv("B:/Needels/Documents/DACCS 601/DACSS_601_New/posts/_data/FAOSTAT_livestock.csv")
#the dimensions of livestock
dim(livestock)
livestock
```

```{r}
#assinging the dairy cattle dataset to an object
dairy<- read_csv("B:/Needels/Documents/DACCS 601/DACSS_601_New/posts/_data/FAOSTAT_cattle_dairy.csv")
#the dimensions for dairy
dim(dairy)
dairy
```

## Join Data

To double check the dimensions I should expect with math that the sum of the rows in both the dairy cattle and livestock should be 36449 + 82116 = 118565

```{r}
#joining the dairy cattle and livestock datasets
cattlestock <- full_join(dairy,livestock)
#the dimension of the joining
dim(cattlestock)
cattlestock
```

We should expect 36449 + 38170 = 74619 for cattle_eggs

```{r}
#joining the dairy cattle and eggs datasets
cattle_eggs <- full_join(dairy, Eggs)
#the dimensions of the joining
dim(cattle_eggs)
cattle_eggs
```

We should expect 82116 + 38170 = 120286

```{r}
#joining the livestock and eggs datasets
stockeggs <- full_join(livestock, Eggs)
#the dimensions of the joining
dim(stockeggs)
stockeggs
```

