---
title: "Challenge 4 submission"
author: "Cam Needels"
description: "More data wrangling: pivoting"
date: "03/27/2023"
format:
  html:
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
  - challenge_4
  - eggs
  - poultry
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)
library(summarytools)
library(readr)
library(readxl)
library(lubridate)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

```{r}
#displaying the dataset
eggs_data<- read_csv("B:/Needels/Documents/DACCS 601/DACSS_601_New/posts/_data/poultry_tidy.csv")
eggs_data

```

```{r}
#this was done in order to figure out the unique data points for product
categories <- unique(eggs_data$Product)
categories
#this was done in order to figure out how many years this data set ranged from.
yearcat <-- unique(eggs_data$Year)
yearcat
```

### Briefly describe the data

It shows the year and month when different forms of chicken were purchased. Whether it's Whole, B/S Breast, Bone-in Breast, Whole Legs, or Thighs and the amount of price per dollar.These ranges from the year 2004 to 2013.

## Tidy Data (as needed)

The data is already tidy so I don't need to make changes. However I do need to figure out the variables in the product category.


## Identify variables that need to be mutated

Product and Month are not numeric or double so we have to convert them into numbers so we can analyze the data more in depth. We will do this by creating a date column by taking the year and month to make a date column. We will also recode the products so that they can be changed into numbers and be able to be used for data analysis. I will use dfsummary afterwards and here are the results.

```{r}
#convert Month -> numbers
eggs_mutate <- eggs_data %>%
  mutate(Month_num = recode(Month, "January" = 1, "February" = 2, "March" = 3, "April" = 4, "May" = 5, "June" = 6, "July" = 7, "August" = 8, "September" = 9, "October" = 10, "November" = 11, "December" = 12))

#assigning IDs to chicken types
eggs_mutate <- eggs_mutate %>%
  mutate(Chicken_ID = recode(Product, "B/S Breast" = 1, 
                             "Bone-in Breast" = 2, 
                             "Thighs" = 3, 
                             "Whole" = 4, 
                             "Whole Legs" = 5))
eggs_mutate
```



```{r}
dfSummary(eggs_mutate)
```
