---
title: "Challenge 6 Solutions"
author: "Meredith Rolfe & Sean Conway & Erico Yu"
description: "Visualizing Time and Relationships"
date: "04/05/2023"
format:
  html:
    toc: true
    code-copy: true
    code-tools: true
categories:
  - challenge_6
  - solution
  - hotel_bookings
  - air_bnb
  - fed_rate
  - debt
  - usa_hh
  - abc_poll
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)
library(ggplot2)
library(readxl)
library(lubridate)
library(here)

source(here("posts","umass_colors.R"))
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Challenge Overview

Today's challenge is to:

1)  create at least one graph including time (evolution)
   - try to make them "publication" ready (optional)
   - Explain why you choose the specific graph type
2)  Create at least one graph depicting part-whole or flow relationships
   - try to make them "publication" ready (optional)
   - Explain why you choose the specific graph type

:::{.panel-tabset}

## Debt ⭐

This data set runs from the first quarter of 2003 to the second quarter of 2021, and includes quarterly measures of the total amount of household debt associated with 6 different types of loans - mortgage,HE revolving, auto, credit card, student, and other - plus a total household debt including all 6 loan types. This is another fantastic macroeconomic data product from the New York Federal Reserve. See Challenge 4.

```{r}
debt_orig<-here("posts","_data","debt_in_trillions.xlsx") %>%
  read_excel()
debt<-debt_orig%>%
  mutate(date = parse_date_time(`Year and Quarter`, 
                           orders="yq"))
```

### Time Dependent Visualization

Lets look at how debt changes over time.

```{r}
#| layout-ncol: 2
#| column: page
#| fig-cap: 
#|   - "Change in Total Debt Over Time"
#|   - "Change in Total Debt, v2"

ggplot(debt, aes(x=date, y=Total)) +
  geom_point()

ggplot(debt, aes(x=date, y=Total)) +
  geom_point(size=.5) +
  geom_line()+
  scale_y_continuous(labels = scales::label_number(suffix = " Trillion"))
  
```
### Visualizing Part-Whole Relationships

One thing to note is that it isn't easy to include multiple lines on a single graph, that is because our data are not pivoted. Here is an example of how pivoting into tidy format makes things super easy. 

```{r}
#| layout-ncol: 2
#| column: page
#| fig-cap: 
#|   - "Change in Debt Over Time, by Debt Type (line)"
#|   - "Change in Debt Over Time, by Debt Type (stacked)"

umass_palette<-c("red", "green", "dark blue", "light blue", "orange", 
                 "yellow")%>%
                   map(., get_umass_color)%>%
                   unlist(.)

debt_long<-debt%>%
  pivot_longer(cols = Mortgage:Other,
               names_to = "Loan", 
               values_to = "total")%>%
  select(-Total)%>%
  mutate(Loan = as.factor(Loan))

ggplot(debt_long, aes(x=date, y=total, color=Loan)) +
  geom_point(size=.5) +
  geom_line() +
  theme(legend.position = "right") +
  scale_y_continuous(labels = scales::label_number(suffix = " Trillion")) +
  scale_colour_manual(values=umass_palette)

ggplot(debt_long, aes(x=date, y=total, fill=Loan)) +
  geom_bar(position="stack", stat="identity") +
  scale_y_continuous(labels = scales::label_number(suffix = " Trillion"))+
  theme(legend.position = "top") +
  guides(fill = guide_legend(nrow = 1)) +
  scale_fill_manual(labels =
                      str_replace(levels(debt_long$Loan), " ", "\n"),
                      values=umass_palette)
```

While the stacked chart might be easier to read in some respects, it is harder to follow individual trend lines. One solution is to reorder in order to preserve as much information as possible.

```{r}
debt_long<-debt_long%>%
  mutate(Loan = fct_relevel(Loan, "Mortgage", "HE Revolving",
                            "Auto Loan", "Student Loan",  
                            "Credit Card","Other"))

ggplot(debt_long, aes(x=date, y=total, fill=Loan)) +
  geom_bar(position="stack", stat="identity") +
  scale_y_continuous(labels = scales::label_number(suffix = " Trillion"))+
  theme(legend.position = "top") +
  guides(fill = guide_legend(nrow = 1)) +
  scale_fill_manual(labels=
                      str_replace(levels(debt_long$Loan), " ", "\n"),
                      values=umass_palette)
```
## Fed Rates ⭐⭐

This data set runs from July 1954 to March 2017, and includes daily macroeconomic indicators related to the effective federal funds rate - or the interest rate at which banks lend money to each other in order to meet mandated reserve requirements. There are 7 variables besides the date: 4 values related to the federal funds rate (*target*, *upper target*, *lower target*, and *effective*), 3 are related macroeconomic indicators (*inflation*, *GDP change*, and *unemployment rate*.)

```{r}
fed_rates_vars<-here("posts","_data","FedFundsRate.csv") %>% 
  read_csv(n_max = 1,
           col_names = NULL)%>%
  select(-c(X1:X3))%>%
  unlist(.)

names(fed_rates_vars) <-c("fed_target", "fed_target_upper",
                         "fed_target_lower", "fed_effective",
                         "gdp_ch", "unemploy", "inflation")
      
fed_rates_orig<-here("posts","_data","FedFundsRate.csv") %>% 
  read_csv(skip=1,
           col_names = c("Year", "Month", "Day", 
                         names(fed_rates_vars)))

fed_rates<-fed_rates_orig%>%
  mutate(date = make_date(Year, Month, Day))%>%
  select(-c(Year, Month, Day))

fed_rates <- fed_rates%>%
  pivot_longer(cols=-date, 
               names_to = "variable",
               values_to = "value")
```
Now we can try to visualize the data over time, with care paid to missing data.

```{r}
fed_rates%>%
  filter(str_starts(variable, "fed"))%>%
ggplot(., aes(x=date, y=value, color=variable))+
  geom_point(size=0)+
  geom_line()+
  scale_y_continuous(labels = scales::label_percent(scale = 1))

```

We can now see how closely the effective rate adheres to the target rate (and can see how the Fed changed the way it set it target rate around the time of the 2009 financial crash). Can we find out more by comparing the effective rate to one of the other macroeconomic indicators?

```{r}
fed_rates%>%
  filter(variable%in%c("fed_effective", "gdp_ch", 
                       "unemploy", "inflation"))%>%
ggplot(., aes(x=date, y=value, color=variable))+
  geom_point(size=0)+
  geom_line()+
  facet_grid(rows = vars(variable))
```

```{r}
year_unemploy <- fed_rates %>%
  pivot_wider(names_from = variable, values_from = value) %>%
  mutate(year=year(date)) %>%
  group_by(year) %>%
  summarise(median_rate=median(unemploy)/100) %>%
  ungroup()
year_unemploy
```

```{r}
year_unemploy %>%
  ggplot(aes(year,median_rate))+
  geom_line()
```

```{r}
year_unemploy %>%
  filter(year<=1981) %>%
  ggplot(aes(year,median_rate))+
  geom_line()+
  scale_y_continuous(labels=scales::percent_format(),limits=c(0,.1))+
  scale_x_continuous(breaks=seq(1955,1980,5))
  labs(x="year",y="median unemployment rate")
```

## usa_hh ⭐⭐⭐

```{r}
income_brackets <- c(i1 = "Under $15,000",
                     i2 = "$15,000 to $24,999",
                     i3 = "$25,000 to $34,999",
                     i4= "$35,000 to $49,999",
                     i5 = "$50,000 to $74,999",
                     i6 = "$75,000 to $99,999",
                     i7 = "$100,000 to $149,999",
                     i8 = "$150,000 to $199,999",
                     i9 = "$200,000 and over")

ushh_orig <- here("posts","_data","USA Households by Total Money Income, Race, and Hispanic Origin of Householder 1967 to 2019.xlsx") %>%
  read_excel(skip=5,
         n_max = 352,
         col_names = c("year", "hholds", "del",
                       str_c("income",1:9,sep="_i"),
                       "median_inc", "median_se", 
                       "mean_inc","mean_se"))%>%
  select(-del)
ushh_orig 
ushh_id<-ushh_orig%>%
  mutate(identity = case_when(
    str_detect(year, "[[:alpha:]]") ~ year,
    TRUE ~ NA_character_
  ))%>%
  fill(identity)%>%
  filter(!str_detect(year, "[[:alpha:]]"))

ushh_id<-ushh_id%>%
  separate(year, into=c("year", "delete"), sep=" ")%>%
  mutate(identity = str_remove(identity, " [0-9]+"),
         across(any_of(c("hholds", "mean_inc", "mean_se", "year")), 
                as.numeric))%>%
  select(-delete)

ushh_id
ushh <-ushh_id%>%
  mutate(gp_identity = case_when(
   identity %in% c("BLACK", "BLACK ALONE") ~ "gp_black",
    identity %in% c("ASIAN ALONE OR IN COMBINATION",
                  "ASIAN AND PACIFIC ISLANDER") ~ "gp_asian",
    identity %in% c("WHITE, NOT HISPANIC", 
                    "WHITE ALONE, NOT HISPANIC") ~ "gp_white",
    identity %in% c("HISPANIC (ANY RACE)") ~ "gp_hisp",
    identity %in% c("ALL RACES") ~ "gp_all"
  ))%>%
  filter(!is.na(gp_identity))%>%
  group_by(year, gp_identity)%>%
  summarise(across(c(starts_with("inc"),starts_with("me"),
                     "hholds"), 
                   ~median(.x, na.rm=TRUE)))%>% # sort of cheating - getting median of a median?
  ungroup()

```

```{r}
ushh %>%
  filter(gp_identity=="gp_all") %>%
  ggplot(aes(year,median_inc))+
  geom_line()
```

```{r}
ushh %>%
  filter(gp_identity!="gp_all") %>%
  mutate(gp_identity=str_remove(gp_identity,"gp_"))%>%
  ggplot(aes(year,median_inc,col=gp_identity))+
  geom_line()+
  scale_x_continuous(limits=c(min(ushh$year),max(ushh$year)),
                     breaks=seq(min(ushh$year),max(ushh$year),by=10))+
  scale_y_continuous(labels = scales::dollar_format())+
  scale_color_discrete(name="identity")+
  labs(x="year",y="median income")
```

## hotel_bookings ⭐⭐⭐⭐

```{r}
bookings_orig<- here("posts","_data","hotel_bookings.csv") %>%
  read_csv()
bookings<-bookings_orig%>%
  mutate(date_arrival = str_c(arrival_date_day_of_month,
                              arrival_date_month,
                              arrival_date_year, sep="/"),
         date_arrival = dmy(date_arrival))%>%
  select(-starts_with("arrival"))
```

```{r}
bookings_cumul <- bookings %>%
  group_by(date_arrival) %>%
  summarise(n=n()) %>%
  ungroup() %>%
  mutate(n_cumul=cumsum(n))
bookings_cumul
```

```{r}
ggplot(bookings_cumul, aes(date_arrival,n_cumul))+
  geom_line()
```

```{r}
bookings_month_n <- bookings %>%
  mutate(month=floor_date(date_arrival,unit="month")) %>%
  group_by(month) %>%
  summarise(n=n()) %>%
  ungroup()

bookings_month_n
```

```{r}
bookings_month_n %>%
  ggplot(aes(month,n))+
  geom_line()+
  scale_x_date(NULL, date_labels = "%b %y",breaks="2 months")+
  scale_y_continuous(limits=c(0,7000))+
  labs(x="date",y="number of bookings")+
  theme(axis.text.x=element_text(angle=90))
```

```{r}
bookings_month_hotel_n <- bookings %>%
  mutate(month=floor_date(date_arrival,unit="month")) %>%
  group_by(month, hotel) %>%
  summarise(n=n()) %>%
  ungroup()
bookings_month_hotel_n
```

```{r}
bookings_month_hotel_n %>%
  ggplot(aes(month,n,col=hotel))+
  geom_line()+
  scale_x_date(NULL, date_labels = "%b %y",breaks="2 months")+
  scale_y_continuous(limits=c(0,5000))+
  labs(x="date",y="number of bookings")+
  theme(axis.text.x=element_text(angle=90))
```

## air_bnb  ⭐⭐⭐⭐⭐
```{r}
airb <- here("posts","_data","AB_NYC_2019.csv") %>%
  read_csv()

#reorder the columns and put host_id and name in the front
example<-subset(airb, select = c(1:11)) #just keep the first ten columns
example<-example[,c(3,4, 9,10, 1:2, 5:8, 11)] #reorder columns to move the room_type and price ahead to the 3rd and 4th columns
```

This dataset contains geolocation data (longtitude and latitude of each property liste), which allows us to create a map of these properties. So let's plot the may by the types of proeprties and their price. Note the options that I add for labels, titles, and colors.

```{r}
ggplot(example, aes(longitude, latitude, size = price, color = room_type), group = neighbourhood_group) +
  geom_point() +
  labs (size = "Price of Property",
        title = "Map of Airbnb Properties in NYC by Price and Room Types",
    subtitle = "the airbnb dataset (AB_NYC_2019.csv)")
  coord_quickmap()

```
Note that this map does not draw the boundary of the New York City. To plot a map with boundaries, we will have to get the shape data and merge the shape data of different neighbours or neighbourhood_groups. One source you can check out that contains the shape data or shape files (including most major countries and the US States) is the ggplot2::map__data().


Let's draw a part-whole relationship by showing different neighbourhood_groups within the sum of bookings of a specific type of room.

```{r}
## Bivariate Visualization(s)
ggplot(airb, aes(x=room_type, fill=neighbourhood_group)) +  # setting x-axis as the room_type variable; #filling each room type by a second variable, the neighbourdhood group the property locates in
 
  geom_bar( ) + #option of barchart
  labs(
    x = "Type of Rooms",
    y = "Amount of Bookings",
    colour = "Neighborhood Group",
    title = "Bivariate Visualization: room_type + neighbourhood_group",
    subtitle = "the airbnb dataset (AB_NYC_2019.csv)"
  )
```


:::
