---
title: "Challenge 6 Paritosh"
author: "Paritosh G"
description: "Visualizing Time and Relationships"
date: "05/28/2023"
format:
  html:
    toc: true
    code-copy: true
    code-tools: true
categories:
  - challenge_6
  - hotel_bookings
  - air_bnb
  - fed_rate
  - debt
  - usa_households
  - abc_poll
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)
library(ggplot2)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Challenge Overview

Today's challenge is to:

1)  read in a data set, and describe the data set using both words and any supporting information (e.g., tables, etc)
2)  tidy data (as needed, including sanity checks)
3)  mutate variables as needed (including sanity checks)
4)  create at least one graph including time (evolution)

-   try to make them "publication" ready (optional)
-   Explain why you choose the specific graph type

5)  Create at least one graph depicting part-whole or flow relationships

-   try to make them "publication" ready (optional)
-   Explain why you choose the specific graph type

[R Graph Gallery](https://r-graph-gallery.com/) is a good starting point for thinking about what information is conveyed in standard graph types, and includes example R code.

(be sure to only include the category tags for the data you use!)

## Read in data

Read in one (or more) of the following datasets, using the correct R package and command.

-   debt ⭐
-   fed_rate ⭐⭐
-   abc_poll ⭐⭐⭐
-   usa_hh ⭐⭐⭐
-   hotel_bookings ⭐⭐⭐⭐
-   AB_NYC ⭐⭐⭐⭐⭐

```{r}
fedr <- read_csv("_data/FedFundsRate.csv")
```

### Briefly describe the data

```{r}
summary(fedr)
```

### Tidying the Data

```{r}
fedr$Date <- as.Date(with(fedr,paste(Day,Month,Year,sep="-")),"%d-%m-%Y")
fedr
```

## Time Dependent Visualization

```{r warning=FALSE}
fedr %>% 
  ggplot(aes(x = Date, y = `Unemployment Rate`)) + 
  geom_line() + 
  labs( x = "Date", y = "Unemployment Rate", title = "Date vs Unemployment rate")
```

```{r}
data_fill <- fedr %>% 
  fill(`Unemployment Rate`, .direction = 'updown')
ggplot(data_fill, aes(x = Date, y = `Unemployment Rate`)) + 
  geom_line() + 
  labs(x = "Date", y = "Effective Federal Funds Rate", title = "Date vs Effective Fund Rate" )
  
```

## Visualizing Part-Whole Relationships

```{r}
data_fill <- fedr %>% 
            fill(`Unemployment Rate`, .direction = 'updown')

  data_fill %>%
  filter(Year > 1999) %>%
  ggplot(aes(x = Date, y = `Unemployment Rate`)) +
  geom_line() +
  labs( x = "Date", y = "Unemployment Rate", title = "Unemployment Rate from 2000 to 2017")
```
