---
title: "Challenge 1"
author: "Gauri Naik"
date: "03/29/2023"
output: html_document
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Challenge 1

### Dataset: Birds.csv

When I was doing this challenge, I was unaware that I had to assign an object name to the file path. However, once I figured that out, as it is not clear in the tutorials given, I managed to manipulate the dataset birds.csv.

```{r, include=FALSE}
library(tidyverse)
library(readr)

  birds<-read_csv("_data/birds.csv")
```

Here is the first couple lines of the table, using the command `head()` as there is a large amount of data involved in the set and would not show up properly.

```{r, echo=FALSE}
birdstop<-head(birds)

library(knitr)
knitr::kable(birdstop, caption = NULL)
```

After the table, I learned how to select, filter, and arrange data. Here I will select only area, item, year, value, and unit as the columns to show. I will then filter by year, for data after 2000, and then arrange it based on value. I also want to show countries with only chickens. I used `%>%` to show one table with all the changes I made. The table is rather large as I did not add a scroll.

```{r, echo=FALSE}
select(birds, 'Area', 'Item', 'Year', 'Value', 'Unit') %>%
  filter_at(vars(3:3), any_vars(. > 2000)) %>%
  filter(grepl("Chickens", Item)) %>%
    arrange('Year') %>%
  knitr::kable(caption = NULL)
```
