---
title: "Challenge 5"
author: "Abhinav Reddy Yadatha"
description: "Introduction to Visualization"
date: "05/14/2023"
format:
  html:
    toc: true
    code-copy: true
    code-tools: true
categories:
  - challenge_5
  - Abhinav Reddy Yadatha
  - AB_NYC_2019.csv
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)
library(ggplot2)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Challenge Overview

Today's challenge is to:

1)  read in a data set, and describe the data set using both words and any supporting information (e.g., tables, etc)
2)  tidy data (as needed, including sanity checks)
3)  mutate variables as needed (including sanity checks)
4)  create at least two univariate visualizations
   - try to make them "publication" ready
   - Explain why you choose the specific graph type
5)  Create at least one bivariate visualization
   - try to make them "publication" ready
   - Explain why you choose the specific graph type

[R Graph Gallery](https://r-graph-gallery.com/) is a good starting point for thinking about what information is conveyed in standard graph types, and includes example R code.

(be sure to only include the category tags for the data you use!)

## Read in data

Read in one (or more) of the following datasets, using the correct R package and command.

-   cereal.csv ⭐
-   Total_cost_for_top_15_pathogens_2018.xlsx ⭐
-   Australian Marriage ⭐⭐ 
-   AB_NYC_2019.csv ⭐⭐⭐
-   StateCounty2012.xls ⭐⭐⭐
-   Public School Characteristics ⭐⭐⭐⭐ 
-   USA Households ⭐⭐⭐⭐⭐

```{r}
ab_nyc_data <- read.csv("_data/AB_NYC_2019.csv")
head(ab_nyc_data)
summary(ab_nyc_data)
unique(ab_nyc_data$room_type)



```

### Briefly describe the data

The dataset comprises 16 columns and primarily focuses on neighborhood groups and room types to determine housing prices. The price range in the dataset spans from 0 to 10000.

## Tidy Data (as needed)

Is your data already tidy, or is there work to be done? Be sure to anticipate your end result to provide a sanity check, and document your work here.

The available room options are categorized into three types: Private room, Shared room, and Entire home/apartment.

```{r}

sum(is.na(ab_nyc_data$reviews_per_month))

ab_nyc_data <- ab_nyc_data %>%  replace_na(list(reviews_per_month = 0))


```

Are there any variables that require mutation to be usable in your analysis stream? For example, do you need to calculate new values in order to graph them? Can string values be represented numerically? Do you need to turn any variables into factors and reorder for ease of graphics and visualization?

Upon observation, I noticed that there were 10052 rows with missing values in the "reviews_per_month" column. In order to provide potential buyers with a more accurate understanding of the reviews for the property, I decided to replace these missing values with 0.

Document your work here.

```{r}



```

## Univariate Visualizations
Let us visualize on how the prices are scattered accross the dataset. 

```{r}
ggplot(ab_nyc_data,aes(price))+
  geom_histogram()

airb_filt_data <- ab_nyc_data %>%
  filter(price>0 & price<2500)

ggplot(airb_filt_data,aes(price))+
  geom_histogram()

```


## Bivariate Visualization(s)

Now, let us visualize two variables : price and room_type across the whole data.

```{r}
ggplot(ab_nyc_data,aes(room_type,price))+geom_boxplot()+labs(title = "Distribution of Airbnb prices across various ranges of room types.")


```

Any additional comments?
