---
title: "Challenge 8"
author: "Rahul Somu"
description: "Joining Data"
date: "05/16/2023"
format:
  html:
    toc: true
    code-copy: true
    code-tools: true
categories:
  - challenge_8
  - railroads
  - snl
  - faostat
  - debt
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)
library(ggplot2)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Challenge Overview

Today's challenge is to:

1)  read in multiple data sets, and describe the data set using both words and any supporting information (e.g., tables, etc)
2)  tidy data (as needed, including sanity checks)
3)  mutate variables as needed (including sanity checks)
4)  join two or more data sets and analyze some aspect of the joined data

(be sure to only include the category tags for the data you use!)

## Read in data

Read in one (or more) of the following datasets, using the correct R package and command.

  - military marriages ⭐⭐
  - faostat ⭐⭐
  - railroads  ⭐⭐⭐
  - fed_rate ⭐⭐⭐
  - debt ⭐⭐⭐
  - us_hh ⭐⭐⭐⭐
  - snl ⭐⭐⭐⭐⭐

```{r}




```

### Briefly describe the data

## Tidy Data (as needed)

Is your data already tidy, or is there work to be done? Be sure to anticipate your end result to provide a sanity check, and document your work here.



After checking the sanity of the data by using glimpse function to check the data types & column names and NA count, it has been found that casts data has more NA values. So it would be ideal to move ahead with actors and seasons data.

```{r}
library(tidyverse)

# Load the data sets
snl_actors <- read_csv("_data/snl_actors.csv")
snl_casts <- read_csv("_data/snl_casts.csv")
snl_seasons <- read_csv("_data/snl_seasons.csv")

# Check the structure of the data sets
names(snl_actors)
glimpse(snl_actors)

names(snl_casts)
glimpse(snl_casts)

names(snl_seasons)
glimpse(snl_seasons)

sum(is.na(snl_actors))
sum(is.na(snl_casts))
sum(is.na(snl_seasons))


```

Are there any variables that require mutation to be usable in your analysis stream? For example, do you need to calculate new values in order to graph them? Can string values be represented numerically? Do you need to turn any variables into factors and reorder for ease of graphics and visualization?

## Join Data

Be sure to include a sanity check, and double-check that case count is correct!

Document your work here.

Below we join both the datasets, then convert dates to appropriate format and create a new colummn with season duration.
Featured, updated_anchor are converted to factors. Then we create a visualisation with number of episodes per season.

```{r}

library(dplyr)

snl <- left_join(snl_seasons, snl_casts, by = "sid")

str(snl)

library(ggplot2)

snl$first_epid.x <- as.Date(as.character(snl$first_epid.x), format = "%Y%m%d")
snl$last_epid.x <- as.Date(as.character(snl$last_epid.x), format = "%Y%m%d")

snl$featured <- as.factor(snl$featured)
snl$update_anchor <- as.factor(snl$update_anchor)

snl$aid <- factor(snl$aid, levels = unique(snl$aid[order(table(snl$aid), decreasing = TRUE)]))

snl$season_duration <- snl$last_epid.x - snl$first_epid.x

snl$season_fraction <- as.numeric(snl$season_fraction)

ggplot(snl, aes(x = sid, y = n_episodes.x)) +
  geom_bar(stat = "identity") +
  labs(x = "Season", y = "Number of Episodes")

```


```{r}

```
