---
title: "Challenge 2 Instructions"
author: "Rahul Somu"
description: "Data wrangling: using group() and summarise()"
date: "03/06/2023"
format:
  html:
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
  - challenge_2
  - railroads
  - faostat
  - hotel_bookings
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)
library(dplyr)
library(ggplot2)
library(readxl)
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Challenge Overview

Today's challenge is to

1)  read in a data set, and describe the data using both words and any supporting information (e.g., tables, etc)
2)  provide summary statistics for different interesting groups within the data, and interpret those statistics

## Read in the Data

Read Birds.csv and get summary for each item

```{r}

df_birds <- read_csv("_data/birds.csv")
df_birds_select <- df_birds %>%
  select('Area', 'Item', 'Year', 'Unit', 'Value', 'Flag', 'Flag Description')

df_birds_select %>%
  filter(Value != "") %>%
  group_by(Year, Item) %>%
  summarize(mean = mean(Value),
            median = median(Value))

data <- na.omit(df_birds_select)

# Create a plot for each item
items <- unique(data$Item)

for (item in items) {
  item_data <- subset(data, Item == item)
  p <- ggplot(item_data, aes(x = Year, y = Value, group = 1)) +
    geom_line() +
    labs(title = item)
  print(p)
}

```


## Describe the data

Using a combination of words and results of R commands, can you provide a high level description of the data? Describe as efficiently as possible where/how the data was (likely) gathered, indicate the cases and variables (both the interpretation and any details you deem useful to the reader to fully understand your chosen data).

Summary of the entire dataframe

```{r}
#| label: summary

summary(df_birds_select)

df_birds_select %>%
  group_by(Item, Year) %>%
  summarise(mean = mean(Value, na.rm = TRUE),
            median = median(Value, na.rm = TRUE))
```

## Provide Grouped Summary Statistics

Conduct some exploratory data analysis, using dplyr commands such as `group_by()`, `select()`, `filter()`, and `summarise()`. Find the central tendency (mean, median, mode) and dispersion (standard deviation, mix/max/quantile) for different subgroups within the data set.

Summary table for world statistics.

```{r}
# Filtering the world data
data_filtered <- filter(df_birds_select, Area == "World")

# Group the data by item
grouped_data <- data_filtered %>% 
  group_by(Item)

# Compute mean, median, and mode for each item
summary_stats <- grouped_data %>% 
  select(Item, Value) %>% # select only the relevant columns
  summarise(mean = mean(Value, na.rm = TRUE), # compute mean and exclude NAs
            median = median(Value, na.rm = TRUE), # compute median and exclude NAs
            mode = as.numeric(names(table(Value))[which.max(table(Value))])) # compute mode using table() function

# Compute standard deviation and quantiles for each item
dispersion_stats <- grouped_data %>% 
  select(Item, Value) %>% # select only the relevant columns
  summarise(sd = sd(Value, na.rm = TRUE), # compute standard deviation and exclude NAs
            min = min(Value, na.rm = TRUE), # compute minimum and exclude NAs
            q1 = quantile(Value, probs = 0.25, na.rm = TRUE), # compute 1st quartile and exclude NAs
            median = median(Value, na.rm = TRUE), # compute median and exclude NAs
            q3 = quantile(Value, probs = 0.75, na.rm = TRUE), # compute 3rd quartile and exclude NAs
            max = max(Value, na.rm = TRUE)) # compute maximum and exclude NAs

# Combine the summary statistics and dispersion statistics into a single table
summary_table <- inner_join(summary_stats, dispersion_stats, by = "Item")

# View the summary table
summary_table

```

### Explain and Interpret

I have picked the highest and least consumed item of the world data.

Based on the data, the mean number of chickens produced is approximately 11.6 million with a standard deviation of approximately 6.1 million. The median number of chickens produced is approximately 10.4 million. The mode is 3.9 million. The minimum number of chickens produced is zero, while the maximum number is approximately 23.7 million. the average (mean) number of chickens increased from around 2 million in 1961 to over 23 million in 2018 in the world. However, there is also a large variability in the number of chickens, as indicated by the high standard deviation of over 6 million.

For pigeons, the mean number produced is approximately 29,400 with a standard deviation of approximately 11,510. The median is approximately 32,222 and the mode is 12,068. The minimum number produced is 12,068, while the maximum is approximately 57,909.Based on the data, the mean and median values for pigeons have remained relatively stable over the years. However, the mode value for pigeons has fluctuated significantly, which indicates that there have been some years where the production of pigeons was significantly higher than others. Overall, there is not a clear upward or downward trend in pigeon production based on the provided data.
