---
title: "Challenge 6 - Hotel Bookings"
author: "Megan Galarneau"
description: "Visualizing Time and Relationships"
date: "04/07/2023"
format:
  html:
    df-print: paged
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
    css: "style.css"
categories:
  - challenge_6
  - Megan Galarneau
  - hotel_bookings
---

```{r}
#| label: setup
#| warning: false
#| message: false
library(tidyverse)
library(ggplot2)
library(dplyr)
library(lubridate)
library(patchwork)
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Challenge Overview

Today's challenge is to:

1)  Read in a data set, and describe the data set using both words and any supporting information (e.g., tables, etc)
2)  Tidy data (as needed, including sanity checks)
3)  Mutate variables as needed (including sanity checks)
4)  Create at least one graph including time (evolution)

-   try to make them "publication" ready (optional)
-   Explain why you choose the specific graph type

5)  Create at least one graph depicting part-whole or flow relationships

-   try to make them "publication" ready (optional)
-   Explain why you choose the specific graph type

## Read in data

```{r}
#read in the data set, raw
library(readr)
raw_hotel <- read_csv("_data/hotel_bookings.csv")
raw_hotel
```

### Briefly describe the data

This data set contains information of over 119k city & resort hotel bookings from July 2015 to August 2017 in 178 different countries. Detailed information about arrival date/times, length of stay, cancellations, room type, reservation status, # of people (adults, childern, or babies), market segment & more are also included in this data set. At first glance, there are more reported city hotel bookings (66.4%) with almost half of the bookings from 2015 (47.5%).


```{r}
#summary of data set statistics
print(summarytools::dfSummary(raw_hotel,
                        varnumbers = FALSE,
                        plain.ascii  = FALSE, 
                        style        = "grid", 
                        graph.magnif = 0.70, 
                        valid.col    = FALSE),
      method = 'render',
      table.classes = 'table-condensed')
```

## Tidy Data

Immediately, I noticed that arrival dates are separated into three columns (year, month, day). Let's mutate these columns in order to graph it later on.


```{r}
#recode month characters to numeric
month_hotel<-raw_hotel%>%
  mutate(tidy_arrival_date_month = case_when(
  arrival_date_month == "January" ~ 1,
  arrival_date_month == "February" ~ 2,
  arrival_date_month == "March" ~ 3,
  arrival_date_month == "April" ~ 4,
  arrival_date_month == "May" ~ 5,
  arrival_date_month == "June" ~ 6,
  arrival_date_month == "July" ~ 7,
  arrival_date_month == "August" ~ 8,
  arrival_date_month == "September" ~ 9,
  arrival_date_month == "October" ~ 10,
  arrival_date_month == "November" ~ 11,
  arrival_date_month == "December" ~ 12)
)

#create new arrival date column
arrival_date_hotel<-month_hotel%>%
  mutate(arrival_date = make_date(arrival_date_year, tidy_arrival_date_month, arrival_date_day_of_month))

#create new total bookings per arrival date column
tidy_time_hotel<-arrival_date_hotel%>%
  mutate(month=floor_date(arrival_date,unit="month")) %>%
  group_by(arrival_date, hotel) %>%
  summarise(n=n()) %>%
  ungroup()

#pivot_longer adults, child, babies

people_hotel <-pivot_longer(raw_hotel, col = c("adults", "children", "babies"),
                 names_to="Person",
                 values_to = "Person Number",
                 values_drop_na = TRUE)

#arrival date sanity check
tidy_time_hotel

#adult, child, baby sanity check
people_hotel
```

## Time Dependent Visualization

Now that our data is tidy, I chose to graph the number of bookings per hotel type over time according to arrival date. I choose a line graph because it gives a granular perspective of each booking and date.


```{r}

time_hotel <- tidy_time_hotel %>%
  ggplot(aes(arrival_date,n,col=hotel))+
  geom_line()+
  scale_x_date(NULL, date_labels = "%b %y",breaks="2 months")+
  scale_y_continuous(limits=c(0,500))+
  labs(x="Arrival Date",y="# of bookings", title = "Hotel Booking Arrival Dates over Time")+
  theme(axis.text.x=element_text(angle=90))

time_hotel


```

## Visualizing Part-Whole Relationships

I wanted to graph the breakdown of adult, child & babies for city vs. resort bookings below. I don't think the quantities are accurately represented though.

```{r}
bar_hotel <- people_hotel %>%
ggplot(mapping = aes(x = hotel, fill = Person), position = "fill") + 
  geom_bar() +
   labs(x="Hotel Type",y="# of Bookings", title = "Occupants of Hotel Rooms by Type")

bar_hotel
```


