---
title: "Challenge 1"
author: "Pooja Shah"
description: "Reading in data and creating a post"
date: "04/26/2023"
format:
  html:
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
  - challenge_1
  - railroads
  - wildbirds
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)
library(readxl)
library(summarytools)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Challenge Overview

Today's challenge is to

1)  read in a dataset, and

2)  describe the dataset using both words and any supporting information (e.g., tables, etc)

## Read in the Data

Read in one (or more) of the following data sets, using the correct R package and command.

-   railroad_2012_clean_county.csv ⭐
-   birds.csv ⭐⭐
-   FAOstat\*.csv ⭐⭐
-   wild_bird_data.xlsx ⭐⭐⭐
-   StateCounty2012.xls ⭐⭐⭐⭐

Find the `_data` folder, located inside the `posts` folder. Then you can read in the data, using either one of the `readr` standard tidy read commands, or a specialized package such as `readxl`.

Reading the railroad dataset
```{r}
#Reading the railroad_2012_clean_county dataset
railroad <- read_csv("_data/railroad_2012_clean_county.csv")
```
Printing first few rows of the railroad dataset
```{r}
#Printing only few rows
head(railroad)
```
Reading the wild bird dataset
```{r}
#Reading the wild_bird_data dataset
wildbird <- read_excel("_data/wild_bird_data.xlsx", skip=1)
```
Printing first few rows of the wildbird dataset
```{r}
#Printing only few rows
head(wildbird)
```
Add any comments or documentation as needed. More challenging data sets may require additional code chunks and documentation.

## Describe the data

Using a combination of words and results of R commands, can you provide a high level description of the data? Describe as efficiently as possible where/how the data was (likely) gathered, indicate the cases and variables (both the interpretation and any details you deem useful to the reader to fully understand your chosen data).

```{r}
#| label: summary
dfSummary(railroad)
dfSummary(wildbird)
```
