---
title: "Challenge 1 Paritosh"
author: "Paritosh Gandhi"
description: "Challenge_1_Final"
date: "03/28/2023"
format:
  html:
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
  - challenge_1
  - railroads
  - faostat
  - wildbirds
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Challenge Overview

Today's challenge is to

1)  read in a dataset, and

2)  describe the dataset using both words and any supporting information (e.g., tables, etc)

## Read in the Data

Read in one (or more) of the following data sets, using the correct R package and command.

-   railroad_2012_clean_county.csv ⭐
-   birds.csv ⭐⭐
-   FAOstat\*.csv ⭐⭐
-   wild_bird_data.xlsx ⭐⭐⭐
-   StateCounty2012.xls ⭐⭐⭐⭐

## Find the `_data` folder, located inside the `posts` folder. Then you can read in the data, using either one of the `readr` standard tidy read commands, or a specialized package such as `readxl`.

------------------------------------------------------------------------

```{r}
library(readxl)
library(tidyverse)
library(summarytools)


df = read_excel("_data/wild_bird_data.xlsx", skip=1)

## instead os using skip we can also use the below line of code

#df <- df[2:147,]







```

## Describe the data

The wild bird data conatins 2 columns and 147 entries. The first columns consists of wet body weights of the birds in grams and and second columns consists of population size.The dataset is different as the reference column which provides the information about wet body weight in grams is stored in the form of character which is converted into numeric form using the "as.numeric" function The minimum wet body weight in grams is 5.45 gms and the mean being 363.74 gms while the max is 9639.84

## ***Using as.numeric(), Min, Max,Mean***

```{r}
#| label: summary 
min(as.numeric(df$`Wet body weight [g]`), na.rm = T)
mean(as.numeric(df$`Wet body weight [g]`), na.rm = T)
max(as.numeric(df$`Wet body weight [g]`), na.rm = T)



```

## ***Using Select()***

```{r}
df %>% 
  select(`Wet body weight [g]`) %>% 
  n_distinct()
```

## ***Using Filter()***

```{r}
df %>% filter(`Wet body weight [g]` > 3000)
```

-   summary(df)

```{r}
dfSummary(df)
```

------------------------------------------------------------------------
