---
title: "Challenge 6"
author: "Gauri Naik"
date: "04/26/2023"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo=TRUE)
```

```{r data, echo = FALSE, message = FALSE}
#tidy data and mutate var as needed
library(tidyverse)
library(summarytools)
library(knitr)
library(treemap)
library(readxl)
debt <- read_xlsx("_data/debt_in_trillions.xlsx")
knitr::kable(head(debt))
```

Here is a preview of the data we will be using.

```{r tidymanip, echo = FALSE}
debt1 <- debt %>%
  separate(`Year and Quarter`, c("Year", "Quarter"), sep = ":") %>%
  arrange(`Year`, `Quarter`) %>%
  group_by(`Year`)

debt_means <- debt1 %>%
  summarise(`Mortgage Mean` = mean(`Mortgage`),
            `Auto Loan Mean` = mean(`Auto Loan`),
            `Credit Card Mean` = mean(`Credit Card`),
            `Student Loan Mean` = mean(`Student Loan`),
            `Mean Miscellaneous` = mean(`Other`),
            `Total Mean` = mean(`Total`))

debt_no21 <- debt1 %>%
  group_by(`Year`, `Quarter`) %>%
  subset(`Year` != 21)
  
debt4 <- rowsum(debt_no21[3:9],rep(1:18,each=4))
Year <- c("03", "04", "05", "06", "07", "08", "09", "10", "11", "12", "13", "14", "15", "16", "17", "18", "19", "20")
debt_prop <- debt4/debt4$Total
debt_prop <- cbind(debt_prop, Year) %>%
  relocate(.before = Mortgage, Year) %>%
  group_by(`Year`)

 debt_yearascol <- data.frame(t(debt4))
 colnames(debt_yearascol) <- c("03", "04", "05", "06", "07", "08", "09", "10", "11", "12", "13", "14", "15", "16", "17", "18", "19", "20")
 
 debt_tree <- debt1
 debt_tree %>%
   group_by(`Year`) %>%
   transform(Year = as.double(Year)) %>%
 pivot_longer(cols = c(`Mortgage`, `HE.Revolving`, `Auto.Loan`, `Credit.Card`, `Student.Loan`, `Other`, `Total`),
              names_to = "Type.of.Debt",
              values_to = "Value")

debt_tree2 <- debt_means
invisible(debt_tree)
```
Above, I've printed one of the tables that I transformed, but eventually I ended up not using it to create any graphs.

Below, I've created four graphs, one of which is evolution, one is part-whole, and two are scatterplots. The evolution one shows the total debt mean over over the years 2003 to 2021, while the part-whole gives a comparative look at the mortgage debt mean in each year, and how each year's debt compares to the other.
```{r graphs, echo = FALSE, message = FALSE, results ='hide'}
#create one graph including time (evolution)
#create one graph depicting part-whole or flow relationships

total_noavg <- ggplot(debt_no21, aes(x = Year, y = Total, color = `Quarter`)) +
  geom_point() +
  theme_minimal() +
  labs(title = "Total Debt per Quarter per Year",
       x = "Year",
       y = "Total Debt per Quarter (in Trillions)") +
  ylim(0, 15)
total_noavg

total_mean <- ggplot(debt_means, aes(x= Year, y = `Total Mean`)) +
  geom_point() + 
  labs(title = "Average of Quarters") +
  ylim(0, 15) +
  geom_curve(aes(x = 1, y = 7.558950, xend = 19, yend = 14.9), position = "identity", curvature = 0, arrow = NULL, ncp = 25)
total_mean

ggplot(debt_means, aes(x=`Year`)) +
  geom_point(aes(y=`Credit Card Mean`, color = "Credit Card Mean")) +
  geom_point(aes(y=`Auto Loan Mean`, color = "Auto Loan Mean")) +
  labs(title = "Mean of Auto Loan Compared to Credit Card over 2003 to 2021",
       caption = "Means were averaged given data for each quarter of each year.",
       x = "Year",
       y = "Mean Debt per Year (in Trillions)",
       color = "Mean Debt") +
  scale_color_manual(values = c("#2781d1", "#d13e27")) +
  theme_minimal()
 
debt_treemap <- treemap(debt_tree2,
            index="Year",
            vSize="Mortgage Mean",
            type="index"
            )
debt_treemap


```
