---
title: "Coding Demonstration Example"
author: "Sean Conway"
description: "Creating a function"
date: "04/22/2023"
format:
  html:
    df-print: paged
    toc: false
    code-fold: false
    code-copy: true
    code-tools: true
    css: styles.css
    mermaid-format: png
categories:
  - Coding_Example
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup
#| warning: false
#| message: false
#| include: false

library(tidyverse)
library(lubridate)
library(ggplot2)
library(here)
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

##Overview

This coding demonstration example presents three cases of R function coding: create a function, and use it to perform a data analysis / cleaning / visualization task:

Examples of such functions are:
1) A function that reads in and cleans a dataset.  
2) A function that computes summary statistics (e.g., computes the z score for a variable).  
3) A function that plots a histogram.  

That's it!

## A function that reads in and cleans a dataset  

This very simple function reads in the [SNL Cast Dataset](https://github.com/hhllcks/snldb) and converts the first and last episode date columns to date format^[Note that most values here are `NA`].  

We also include an optional argument that allows us to only look at those cast members who were an anchor on [Weekend Update](https://en.wikipedia.org/wiki/Weekend_Update). This can simplify the process of filtering later on.  

```{r}
read_snl_casts <- function(f,update_only=F){
  d <- read_csv(f)
  d <- d %>%
    mutate(across(c(first_epid,last_epid),~ymd(as.character(.x))))
  if(update_only){
    d <- d %>%
      filter(update_anchor)
  }
  return(d)
}
snl_casts <- here("posts","_data","snl_casts.csv") %>%
  read_snl_casts()
snl_casts
snl_update_casts <- here("posts","_data","snl_casts.csv") %>%
  read_snl_casts(update_only = T)
snl_update_casts
```

## A function that computes summary statistics  

To use such a function, we'll first import the [hotel bookings](https://www.kaggle.com/datasets/jessemostipak/hotel-booking-demand) dataset, which includes plenty of quantitative variables.  

```{r}
hotels <- here("posts","_data","hotel_bookings.csv") %>%
  read_csv()
hotels
```

Our function will compute the mean, median, and sd for a variable. Specifically, the function will take a vector (or a column from a data frame) and output a tibble of these summary statistics.  

```{r}
sum_stat <- function(x){
  stat <- tibble(
    mean=mean(x,na.rm=T),
    median=median(x,na.rm=T),
    sd=sd(x,na.rm=T)
  )
  return(stat)
}
sum_stat(hotels$lead_time)
```


<!-- ```{r} -->
<!-- map_dfr(list(hotels$lead_time,hotels$babies,hotels$days_in_waiting_list),sum_stat) %>% -->
<!--   mutate(id=c("lead time","babies","days in waiting list")) %>% -->
<!--   relocate(id,.before=everything()) -->
<!-- ``` -->

## A function that plots a histogram  

This is also a very simple function - essentially a [wrapper function](https://en.wikipedia.org/wiki/Wrapper_function) around `geom_histogram()` that plots a histogram of a variable. We'll use the `hotel bookings` dataset again - specifically the `lead_time` variable.  
 
We can use a double-brace around the name of the column we want a histogram of. For more details around this approach (known as data masking), visit [this link](https://dplyr.tidyverse.org/articles/programming.html#indirection).  

```{r}
make_my_hist <- function(dat, colname, fill="purple", xlab="x", ylab="n"){
  dat %>%
    ggplot(aes({{colname}}))+
    geom_histogram(fill=fill)+
    labs(x=xlab,
         y=ylab)
}
make_my_hist(hotels,lead_time,xlab="lead time")
```

We can save our plot as an object and add to it using the `+` operator.  

```{r}
my_graph <- make_my_hist(hotels,lead_time,xlab="lead time")
my_graph_w_title <- my_graph +
  labs(title="Lead time histogram")
my_graph_w_title
```

## Conclusion  

There's clearly a lot we can do with functions. This solution comprises a small chunk of how to do so. A crucial aspect of programming with functions is knowing the extent to which your function will work. For example, the `read_snl_casts()` function will fail on any dataset that isn't `snl_casts.csv`, while the `make_my_hist()` function will work on many different datasets.  

Functions are great for automating processes and avoiding copy-paste. We will see more of how to do so in the next Function Coding Example - `purrr`.  

