---
title: "Final Project Assignment#1: Susmita Madineni"
author: "Susmita Madineni"
description: "Global Internet users dataset from 1980-2020"
date: "04/12/2023"
format:
  html:
    df-print: paged
    toc: true
    code-copy: true
    code-tools: true
    css: styles.css
categories:
  - final_Project_assignment_1
  - final_project_data_description
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Part 1. Introduction {#describe-the-data-sets}

In this part, you should introduce the dataset(s) and your research questions.

1.  Dataset Introduction:

The internet users have grown exponentially since the time of its inception in the late 1960s. It is one of the most transformative inventions of the modern times. In the early days, primarily scientists, researchers and academics used the internet to share information across on the collaborated projects. But, later on due to the technological advancements, many people started to use internet for various activities. It evolved into a platform that connects people from all over the world  for various reasons. 

There are a number of factors that has contributed to the growth of internet, such as faster and more reliable connections, the development and increase in usage of smartphones and other mobile devices. Due to the internet's increasing accessibility and user-friendliness, the usage has risen exponentially over the years.

Internet usage has become an integral part of a person's life today. From communication and entertainment to education and commerce, billions of people utilize the internet on a daily basis. As the internet continues to develop and new technologies are emerging, it is foreseen that its use will continue to grow and transform our way of living, working and interacting with the world. 

The dataset consists of country-wise internet users from 1980 to 2020. Each row in the dataset represents the mobile phone subscription per 100 people along with percentage of internet users using the internet, number of people using the internet and number of fixed broadband subscriptions per 100 people for a particular country and an year.


2.  What questions do you like to answer with this dataset(s)?

With the help of the data analysis, I would like to answer the following questions:

1. How the general trends of internet usage(mobile phone, internet and broadband subscriptions) has been over the years from 1980 to 1920 for the countries? Are we observing a general pattern of user growth based on the telecommunication mode?

2. What does the relationship between the mobile phone subscriptions and internet users looks like over the years? Are they strongly correlated to each other? Does increase in mobile phone subscriptions has an effect on the usage of the internet? 

3. Similarly what does the relationship between the fixed broadband subscriptions and internet users looks like over the years? Are they strongly correlated to each other? Does increase in fixed broadband subscriptions has an effect on the usage of the internet? 

4. Which countries has the most number of internet users in the last 10 years? How much percentage population of internet users are being occupied by the top 10 countries?

 Many organizations can benefit with the help of these data visualizations. For example,
 - Market research companies can understand consumer behavior, trends and preferences in different parts of the world.They can make informed decisions while marketing the products, in product development and target the suitable demographics.
 - Advertising agencies will be able to identify the most popular and effective advertising channels and strategize accordingly to reach their target audience.
 - E-commerce companies can also benefit from this. They can understand the customer behavior, trends and preferences in different parts of the world. It will help them create a better experience for their customers and offer targeted promotions and discounts. 
 - Government agencies can use these to monitor the usage patterns and identify any potential security threats, gain insights to what the public wants and also understand the sentiment of the public regarding various issues.
 - Social media platforms can use it to understand trends and preferences and improve user satisfaction and engagement.
 - Educational institutions, can understand how students are using the internet for research and learning, and to develop more effective online education tools and platforms.

## Part 2. Describe the data set(s) {#describe-the-data-sets-1}

This part contains both a coding and a storytelling component.

In the coding component, you should:

1.  read the dataset;

```{r}
library(readr)
globalInternetUsers <- read_csv("C:/Users/SUSMITA/OneDrive/Documents/Spring23/Courses/DACSS_601/DACSS_601/posts/SusmitaMadineni_FinalProjectData/Final.csv")
View(globalInternetUsers)

```

This dataset consists of 229 countries, where each row represents the mobile phone subscription per 100 people(Cellular Subscription) along with percentage of internet users using the internet(Internet Users(%)), number of people using the internet(No. of Internet Users) and number of fixed broadband subscriptions per 100 people(Broadband Subscription) for a particular country(Entity) and an year. The dataset has the above mentioned information from the year 1980 to 2020 for all the 229 countries.

2.  present the descriptive information of the dataset(s) using the functions in Challenges 1, 2, and 3;

    -   for examples: dim(), length(unique()), head();

```{r}
# Preview the first few rows of the dataset
head(globalInternetUsers)

# Understanding the dimensions of the dataset 
dim(globalInternetUsers)

# Identifying the column names of the dataset 
colnames(globalInternetUsers)

# Identifying unique values for column - entity
length(unique(globalInternetUsers$Entity))

# Identifying the data types of the columns
table(sapply(globalInternetUsers, function(x) typeof(x)))

sapply(globalInternetUsers, class)

```



There are 8867 rows and 8 columns("...1" , "Entity", "Code", "Year", "Cellular Subscription", "Internet Users(%)",    "No. of Internet Users", "Broadband Subscription") in total in the dataset.There are 2 character datatype columns - "Entity", "Code" and there are 6 numeric columns - "...1" , "Entity", "Cellular Subscription", "Internet Users(%)",    "No. of Internet Users", "Broadband Subscription".  


3.  conduct summary statistics of the dataset(s); especially show the basic statistics (min, max, mean, median, etc.) for the variables you are interested in.

```{r}
# Summary statistics of the datasets 

summary(globalInternetUsers$`Cellular Subscription`)
summary(globalInternetUsers$`Internet Users(%)`)
summary(globalInternetUsers$`No. of Internet Users`)
summary(globalInternetUsers$`Broadband Subscription`)

```

Based on the summary statistics for the cellular subscription per 100 people, we can say that average is 39.990 with a maximum value as 436.103. The average value for the percentage of internet users is 17.0436% across all the years for the countries with a maximum value as 100%. The average value for the number of internet users is 1.005e+04 across all the years for the countries with a maximum value as 4.700e+09. The average value for the fixed Broadband Subscription is 4.441 across all the years for the countries with a maximum value as 78.524.


## 3. The Tentative Plan for Visualization {#the-tentative-plan-for-visualization}

With a bivariate visualization the relationship between the variables can be clearly depicted. For example, the increase in usage of internet over the years for various telecommunication modes(internet, mobile phones and broadband subscriptions) can be visualized using Time Series. We can plot the evolution using ggplot2 and build an interactive Candlestick plot using dygraphs package. The dataset is put together in a way that each country has percentage of Cellular subscriptions, internet users(%), percentage of Broadband subscriptions along with total number of internet users in the country. I would pivot the dataset using `pivot_longer` and create a column-"Telecommunication mode", where the type of modes(3) and the respective percentage will be tracked. 

Likewise, the relationship between mobile phone subscriptions and internet users and the relationship between the fixed broadband subscriptions and internet users over the years can be depicted using correlogram. The GGALLY package has ggpairs() function that can build a classic correlogram with scatterplot to visualize the relationship between the aforementioned numeric variables. With the use of interactive Choropleth map with leaflet package we can build an interactive map in R to understand the number of internet users across the world in the past 10 years. 

The data is tidy enough to conduct these visualizations. There are no null data and outliers in the dataset, which makes it easy to use it for data analysis and visualization. To visualize the number of internet users in the last 10 years, I would mutate the dataset by creating a new variable and store the average value of the number of internet users for ten years for each country. For visualizing the correlation between other numeric data, I would map the country and number of mobile phone subscriptions and internet users and visualize the correlation between them. I would do a similar visualization for fixed broadband subscriptions and internet users.
