---
title: "Final Project: Kevin Martell Luya"
author: "Kevin Martell Luya"
description: "Project submission"
date: "05/31/2023"
format:
  html:
    df-print: paged
    toc: true
    code-copy: true
    code-tools: true
    css: styles.css
categories:
  - final_Project
editor_options: 
  chunk_output_type: console
editor: 
  markdown: 
    wrap: 72
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)
library(ggplot2)
library(here)
library(mosaic)
library(tidyverse)
library(lubridate)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)

```

## Abstract {#abstract}

Swimming has become a fundamental aspect of human culture and has been
practiced for thousands of years. They provide a platform for physical
activity, competition, and entertainment. Swimming is one of the many
sports that have gained popularity worldwide due to its many health
benefits and inclusion in major international events like the Olympic
Games. Competitive swimming features a variety of events, including
freestyle, breaststroke, backstroke, and butterfly. The world records
for each event are continually being broken by exceptional world top
athletes who push the limits of human performance medley, with races
ranging from 50 meters to 1500 meters. This paper will focus on the Top
200 world times data set, which contains historical records of the best
200 swimming times from 1980 to 2020, and answer the following research
questions.

1\. Are there any patterns or trends in the progression of swimming
world records over time? For example, are records being broken more
frequently during certain events or specific periods? Or Are world
records more likely to be broken in shorter or longer races or specific
stroke categories?

2\. Is there a relationship between a swimmer's age, gender, or
nationality and their likelihood of breaking a world record? For
example, do younger or older swimmers tend to break records more often
than their peers?

3\. Have advancements in technology or changes in swimming equipment
impacted the frequency or magnitude of world record-breaking? For
example, have swimsuits or other equipment innovations led to more world
records being broken in recent years?

## Data set {#describe-the-data-sets}

Data.world is a platform for data collaboration, analysis, and sharing
designed to help teams and organizations discover, use, and publish
high-quality data in different contexts, such as competitive swimming.
Data has been centralized by this organization resulting in the Top 200
world times. In total, there are 5,200 instances and 14 columns. In this
data set, each row represents a different swim record. Each column
contains a specific piece of information such as the event names, swim
times, swim date, event description, and athlete name, as well as
information about the team they represent, their full name, gender,
birth date, rank order, duration of their swim and last but not the
least the city and country code where the event took place in the world.
The data is sourced from various international swimming competitions,
including the Olympic Games, the FINA World Championships, and national
championships in different countries.

```{r}
# read in the swim data frame
df_swim<- read_csv("./_data/swimming_database.csv")
# look at the head of the data
head(df_swim)
```

This data set has records from 716 swimmers and 474 events worldwide
from 1980 to 2020.

```{r}
# dimensions of the data set
dim(df_swim)
# number of unique Event name
length(unique(df_swim$`Event Name`))
# number of unique Team name
length(unique(df_swim$`Team Name`))
# number of unique Athletes
length(unique(df_swim$`Athlete Full Name`))
# table of the 26 unique Swimming Events
table(df_swim$`Event description`)
```

Before conducting a summary statistics of the data set, we need to
compute and store in a dedicated column 1) the swimmers' age when they
competed at a given event in time and 2) the swimming event year when it
took place. Then, we will mutate the table and factor the columns. This
process will facilitate and shorten the code implementation for the
following research questions.

Let us extract the year from the Swim Date column.

```{r}
# extract the event year
date <- parse_date_time(df_swim$`Swim date`, orders = "mdy")
# concatenate the years as a new column
df_swim$event_year = year(date)
```

Because there are swimmers that were born before 1968 and Libridate has
a difficult time parsing years before 1968 due to the "Year 2000
problem," we created a function that add the century to the year: from
this "dd-mm-yy" to "dd-mm-yyyy" format.

Now, let's extract the swimmers' age in the year of the competition
using add_century function.

```{r}
# extract the athlete birth date year
date <- parse_date_time2(as.character(df_swim$`Athlete birth date`), "mdy", cutoff_2000 = 0)
# concatenate the event_athle_age as a new column
df_swim$event_athle_age = df_swim$event_year -  year(date)
```

Now, let's factor the columns of interest and show the statistic of the
variables swim_time and event_athle_age.

```{r}

df_swim <- df_swim %>%
  mutate(
    `Event Name` = factor(`Event Name`),
    `Team Code` = factor(`Team Code`),
    `Team Name` = factor(`Team Name`),
    `Athlete Full Name` = factor(`Athlete Full Name`),
    `Gender` = factor(`Gender`),
    `Athlete birth date` = factor(`Athlete birth date`),
    `Rank_Order` = factor(`Rank_Order`),
    `City` = factor(`City`),
    `Country Code` = factor(`Country Code`)
  )

summary(df_swim)
# stats of Swim time 
fav_stats(df_swim$`Swim time`)
# stats of swimmer's age at time of competition
fav_stats(df_swim$event_athle_age)
```

Let us take a quick view of each country's records from 1980 to 2022.

```{r}
df_swim %>%
  group_by(`Team Code`) %>%
  summarise(record_counts = length(`Team Code`)) %>%
  mutate(`Team Code` = fct_reorder(`Team Code`, record_counts)) %>%
  ggplot() + geom_col(aes(record_counts, `Team Code`)) +
  labs(title = "Country record (1980 - 2022) ",
       x = "Number of records", y ="Country")
 
```

We can see that the dominant countries are The United States (USA),
Australia(AUS), Japan(JPN), China(CHN), and Great Britain(GBR). Does
this mean we must focus only on these five countries to answer the
research questions? The answer is no. All countries are representative
of our analysis, for hidden patterns might be found at the event/style
level, which leads us to answer the first question. Are there any
patterns or trends in the progression of swimming world records over
time? We can inspect the records that have been broken more frequently
in certain events and during what periods.

Let us create a bar graph that shows the number of records broken per
year and see if the swimming historical data reveals a trend between
1980 and 2020.

```{r}
# minimum and maximum years
min_year <- min(df_swim$event_year)
max_year <- max(df_swim$event_year)

df_swim %>%
  ggplot(aes(x = event_year)) + 
  geom_bar(stat = 'count') +
  scale_x_continuous(breaks = seq(min_year, max_year, by = 2 )) +
  labs(title = "Year records (1980 - 2022)",
       x = "Years", y ="Number of records") +
  theme(axis.text.x = element_text(angle = 45))
```

The data shows that over this range of years, swimmers of different
ages, nationalities, and categories have continuously been setting new
world records, suggesting that the number of swimming records has
increased as time progressed.

There are five crucial peaks (2008, 2009, 2016, 2019, and 2021) in the
bar graph that we will analyze since the number of records increased
continuously yearly.

```{r}
# filter data by years of interest
peaks_yrs <- df_swim %>%
  filter(
      df_swim$event_year == 2008 |
      df_swim$event_year == 2009 |
      df_swim$event_year == 2016 |
      df_swim$event_year == 2019 |
      df_swim$event_year == 2021 )

peaks_yrs <- peaks_yrs %>%
  mutate(event_year  = factor(event_year))

```

Analyzing such years in the context of the length of race, we will see
whether there is a certain distance in which swimmers have set more new
records compared to others' distances. For this graph, we do not
consider the nationality, gender, or age of the swimmer who set a record
since this is an analysis at the swimming-length level.

```{r}
# extract the length of the races
event_descriptions <- peaks_yrs$`Event description`
reg <- regexpr("\\d+", event_descriptions)
peaks_yrs$length_race <-regmatches(event_descriptions, reg)

# create a histogram of the peak years
ggplot(peaks_yrs, aes(length_race, fill = event_year)) +
  geom_histogram(stat = "count",position="dodge2") +
  theme(legend.position = "top") +
  labs(x = "Length of races in meters (m)", 
       y = "Number of broken records", 
       title = "Records by swimming length") +
  guides(fill = guide_legend(title = "Year of observation")) +
  coord_flip()
```

Then, let us analyze such years in the context of swimming events and
see whether there are certain swimming events in which swimmers have set
more new records than other events. Even though the names of the
swimming event involve gender, they are officially unique swimming
events. For this graph, we do not consider the nationality, gender, or
age of the swimmer who set a record since this is an analysis at the
swimming-event level.

```{r}
# Create a bar graph of the filtered data
peaks_yrs %>% 
  ggplot(aes(`Event description` )) +
    geom_bar(aes(fill= event_year), 
             position = position_stack(reverse = TRUE)) +
    theme(legend.position = "top") +
    labs(x = "Type of swimming event", 
         y = "Number of broken records", 
         title = "Broken records by swimming event") +
    guides(fill = guide_legend(title = "Year of observation")) +
    coord_flip() 
```

Thanks to this visualization, in general, swimmers are more likely to
set new records more frequently in the following swimming events: Women
100 Backstroke LCM Female, Men 100 Breaststroke LCM Male, and Men 100
Freestyle LCM Male, for the proportions of new records increased year
after year during this periods. Likewise, based on the graph, the
swimming strokes more likely to be broken are breaststroke, butterfly,
and freestyle: the proportions are increasing yearly.

*Moving forward to the second question, t*o investigate the relationship
between age, gender, or nationality and their likelihood of breaking a
world record, we will use histograms for 1) swimmers; age at year
competition, 2) swimmers' gender, and 3) swimmers' nationality.

Let us plot the age histograms by age from the whole data set.

```{r}
df_swim %>%
  ggplot( aes(event_athle_age, center = event_athle_age), bins = 100 ) +
  geom_histogram(binwidth = 1, 
                 fill = "skyblue", 
                 color = "gray",
                 breaks = seq(5, 40, by = 1)) +
  scale_x_continuous(breaks = seq(5, 40, by = 2),
                     labels = seq(5, 40, by = 2)) +
  labs(title = "Age distribution at a year of competition (1980 - 2020)",
       x = "Age", 
       y ="Number of swimmers")

```

We observe that swimmers aged between 21 and 25 years old at the year of
the competition have set more records from 1980 to 2020. Physical
development would be the most reasonable explanation. Younger swimmers
may have a greater potential for physical development, as their bodies
are still growing and developing. This can give them an advantage in
terms of strength, power, and flexibility, which can help them to swim
faster and break records.

Let us plot a histograms of the age distributions by country.

```{r}
df_swim %>%
  ggplot(aes(event_athle_age)) +
  geom_histogram(binwidth = 1, 
                 fill = "skyblue", 
                 color = "gray",
                 show.legend = TRUE) +
  xlim(10,40) +
  facet_wrap(~df_swim$`Team Code`) +
  labs(title = "Age distribution by country",
       x = "Age", 
       y ="Number of swimmers")

```

We can see that The United States (USA), Australia(AUS), Japan(JPN),
China(CHN), Great Britain(GBR), Italy(ITA), and Russia(RUS) have the
swimmers whose ages were between 21 and 25 at a year of competition,
suggesting that swimmers from these countries and between those ages are
more likely to set new world records in future swimming champions.
Genetics and body type could one explanation. Certain countries may have
a higher prevalence of individuals with genetic traits that are
beneficial for swimming, such as a tall and lean body type or a high
percentage of fast-twitch muscle fibers.

Let us plot a third histogram disaggregated by gender.

```{r}
df_swim %>%
  ggplot( aes(event_athle_age)) +
  geom_histogram(binwidth = 1, 
                 fill = "skyblue", 
                 color = "gray" ,
                 breaks = seq(5, 40, by = 1)) +
  scale_x_continuous(breaks = seq(5, 40, by = 2),
                     labels = seq(5, 40, by = 2)) +
  facet_wrap(~df_swim$Gender) +
  labs(title = "Age distribution by gender",
       x = "Age", 
       y ="Number of swimmers")
```

These histograms suggests that males between 21 and 25 years old have a
higher likelihood of setting new records compared to females swimmer.
Why is that? One factor could be physiology - men generally have more
muscle mass and larger body size than women, which can give them a
physical advantage in swimming. This can be particularly beneficial in
events that require power and strength, such as the sprint events.
However, the gender gap in world record-breaking is narrowing. In recent
years, female swimmers have been breaking records with increasing
frequency, and the gap between male and female records is slowly
closing.

*Have advancements in technology or changes in swimming equipment had an
impact on the frequency or magnitude of world record breaking? For
example, have swimsuits or other equipment innovations led to more world
records being broken in recent years?*

To answer the last question regarding the advancements in
technology or changes in swimming equipment that had an impact on the frequency or magnitude of world record-breaking, we will 1) use all the data set, 2) create eight year categories (10-15, 16-20, 21-25, 26-30, 31-35, 36-40, 41-45, and 46+), and 3) create a new column with the swimming style (freestyle, butterfly, backstroke, breaststroke, and medley). This will help us abstract the number of swimmers based on their ages for each swimming style and see which of them they have broken more records for a given year. We want to visualize jumps in the number of records per year. 
```{r}
# divide categories according to age
best_performace <- df_swim %>%
  mutate(age_category = case_when(
    event_athle_age > 10  & event_athle_age <= 15  ~ "10-15",
    event_athle_age > 15  & event_athle_age <= 20  ~ "16-20",
    event_athle_age > 20  & event_athle_age <= 25  ~ "21-25",
    event_athle_age > 25  & event_athle_age <= 30  ~ "26-30",
    event_athle_age > 30  & event_athle_age <= 35  ~ "31-35",
    event_athle_age > 35  & event_athle_age <= 40  ~ "36-40",
    event_athle_age > 40  & event_athle_age <= 45  ~ "41-45",
    TRUE ~ "+46" ))

# extract swimming style from Event description
for (i in 1:dim(peaks_yrs)[1]) {
  best_performace$swim_style_category[i] <- strsplit(best_performace$`Event description`, " ")[[i]][3]
}
```

Let us plot the graph.
```{r}
# Create a bar graph of the filtered data
best_performace %>% 
  ggplot(aes(x = event_year,
             y = age_category,
             color =  swim_style_category,
             alpha = 0.3)) +
  geom_count(aes(size = after_stat(prop), group = 1)) +
  facet_wrap(~swim_style_category) +
  scale_size_area(max_size = 8) +
  labs(title = "Records by year and swimimg category",
       x = "Year", 
       y ="Age category")
```

Advancements in technology and changes in swimming equipment have had a
significant impact on the frequency and magnitude of world record
breaking in recent years. High-tech swimsuits were introduce in the early 2000s had a major impact on the number of world records being broken. we can observe that the proportions of world records started to increase from 2000's. These swimsuits were designed to reduce drag and improve buoyancy, which allowed swimmers to swim faster and more efficiently. The suits were so effective that they were eventually banned by the International Swimming Federation (FINA) in 2010.

## Conclusion and Discussion {#conclusion}

This data analysis suggests that there are patterns or trends in the progression of swimming, such swimmers between 21 and 25 have a higher likelihood of breaking world records in Women
100 Backstroke LCM Female, Men 100 Breaststroke LCM Male, and Men 100 Freestyle LCM Male. We also explored the relationship between a swimmer's age, gender, or nationality and their likelihood of breaking a world record. Finally, high-tech swimsuits have impacted the frequency or magnitude of world record-breaking starting in 2010. The top 200 world records allowed us to explore historical data, suggesting the findings we presented in this paper.

## Bibliography

    - For reference, you can check out [the source of the dataset](https://data.world/romanian-data/swimming-dataset-top-200-world-times)
    - R as a programming language
