---
title: "Final Project Assignment#2"
author: "Rahul Somu"
description: "Exploratory Analysis and Visualization"
date: "05/15/2023"
format:
  html:
    df-print: paged
    toc: true
    code-copy: true
    code-tools: true
    css: styles.css
categories:
  - final_Project_assignment_2
  - final_project_Exploratory_Analysis_and_Visualization
editor_options: 
  chunk_output_type: inline
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)
library(dplyr)
library(ggplot2)
library(readxl)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Overview of the Final Project {#overview-of-the-final-project}

Introduction:

Housing affordability is one of the major problems in the United States. With many households struggling to afford the rising cost of housing, it is critical to analyze the data trends over the country. According to statistics from the U.S. Census Bureau on median household income by city and region from 1996 to the present, Housing affordability has changed significantly over the time.

Although there has been general increase in median household income over the previous few decades, it has not been uniform across all the regions. Some regions have seen greater growth than others and similar swings have been observed in the median home prices, with some regions seeing more significant price increases than others.

The widening disparity between median household incomes and median property prices is one of the most obvious patterns in the statistics which is becoming more challenging for many Americans to afford homeownership as median property prices in many locations have been rising more quickly than median household incomes. The affordability issue has also been made worse by the fact that, despite the median household income rising, it has not kept up with the cost of living.

Overall, according to data from the U.S. Census Bureau, the affordability of housing has become a major worry over time, with many Americans finding it difficult to keep up with the rising cost of housing. The data emphasize the necessity of making policy changes to address the housing affordability crisis and guarantee that all Americans have access to safe, reasonably priced housing.

As part of this project I'm trying to focus on below: Visualize the trends in home values, the average household income, and inflation rates through time, broken down by area and city size. Create graphs showing the evolution of housing affordability over time in the country's largest cities and how it differs by area and city size.

Data Sources:

As part of this project I found below resources that will help me understand the issue. Zillow Home Value Index (ZHVI) - a dataset from Zillow with the median home value for each region and city in the United States from 1996 to the present. https://www.kaggle.com/datasets/robikscube/zillow-home-value-index?resource=download

U.S. Census Bureau data on median household income by city and region from 1996 to the present. https://www.census.gov/data/tables/time-series/demo/income-poverty/historical-income-households.html

U.S. Bureau of Labor Statistics data on inflation rates by region and year from 1996 to the present. https://www.bls.gov/cpi/regional-resources.htm

1.  Util to Create a look up for states to region

```{r}

state_to_region <- function(state) {
  region_lookup <- list(
    Northeast = c("Connecticut", "Maine", "Massachusetts", "New Hampshire", "New Jersey", "New York", "Pennsylvania", "Rhode Island", "Vermont"),
    South = c("Delaware", "District of Columbia", "Florida", "Georgia", "Maryland", "North Carolina", "South Carolina", "Virginia", "West Virginia", "Alabama", "Kentucky", "Mississippi", "Tennessee", "Arkansas", "Louisiana", "Oklahoma", "Texas"),
    Midwest = c("Illinois", "Indiana", "Michigan", "Ohio", "Wisconsin", "Iowa", "Kansas", "Minnesota", "Missouri", "Nebraska", "North Dakota", "South Dakota"),
    West = c("Arizona", "Colorado", "Idaho", "Montana", "Nevada", "New Mexico", "Utah", "Wyoming", "Alaska", "California", "Hawaii", "Oregon", "Washington")
  )
  region <- NA
  for (r in names(region_lookup)) {
    if (state %in% region_lookup[[r]]) {
      region <- r
      break
    }
  }
  return(region)
}

```

2.  Read The Datasets

```{r}
getwd()
income_data <- read_excel("Rahul_Final_projectData/income_state.xlsx")

df_inflation <- read_excel("Rahul_Final_projectData/USA_inflation_data.xlsx")

df_rentalindex <- read.csv("Rahul_Final_projectData/ZHVI.csv")

```

3.  Create functions for getting data for each state

```{r}
#Function to get income data by state and year

get_median_se <- function(income_data, year, state){
  rownum <- which(income_data$State == state)
  colnum <- which(colnames(income_data) == year)
  median_income <- as.numeric(income_data[rownum, colnum])
  se <- as.numeric(income_data[rownum, colnum+1])
  #diff <- median_income - se
  return(c(as.character(median_income), as.character(se)))
}

get_median_se(income_data, "2021", "Alabama")


#Function to get inflation data

avg_by_region_and_year <- function(df, state, year) {
  region <- state_to_region(state)
  subset_df <- subset(df, Region == region & Year == year)
  row_avgs <- apply(subset_df[, 4:ncol(subset_df)], 1, mean, na.rm = TRUE)
  overall_avg <- mean(row_avgs, na.rm = TRUE)
  return(overall_avg)
}

# example usage
avg_by_region_and_year(df_inflation, "Texas", 2000)


df_rentalindex$Date <- as.Date(df_rentalindex$Date, format = "%Y-%m-%d")

# function to calculate average rental index for a state and year
get_state_rentalindex <- function(df,state, year) {
  df_filtered <- subset(df, format(as.Date(Date), "%Y") == year)
  state_new <- chartr(" ", ".", state)
  state_avg <- mean(df_filtered[[state_new]])
  return(state_avg)
}
# example usage
get_state_rentalindex(df_rentalindex,"District of Columbia", "2000")

```

2.  present the descriptive information of the dataset(s) using the functions in Challenges 1, 2, and 3;

```{r}
dim(income_data)
dim(df_inflation)
dim(df_rentalindex)
head(income_data)
```

3.  conduct summary statistics of the dataset(s); especially show the basic statistics (min, max, mean, median, etc.) for the variables you are interested in.

```{r}

summary(df_rentalindex)

```

Combining all the dataframes to a single dataframe
```{r}
# Create an empty data frame to store the results
result_df <- data.frame(State = character(),
                        Year = character(),
                        Median_Income = numeric(),
                        Standard_Error = numeric(),
                        Inflation_Average = numeric(),
                        Rental_Index = numeric(),
                        stringsAsFactors = FALSE)

# Loop through each state and year
for (state in unique(income_data$State)) {
  for (year in 2000:2020) {
    median_se <- get_median_se(income_data, as.character(year),as.character(state))
    inflation_avg <- avg_by_region_and_year(df_inflation, as.character(state), as.character(year))
    rental_index <- get_state_rentalindex(df_rentalindex, as.character(state), as.character(year))
    #print(median_se,inflation_avg,rental_index)
    # Add the calculated values to the result data frame
    result_df <- rbind(result_df, data.frame(State = state,
                                             Year = as.character(year),
                                             Median_Income = as.numeric(median_se[1]),
                                             Standard_Error = as.numeric(median_se[2]),
                                             Inflation_Average = inflation_avg,
                                             Rental_Index = rental_index,
                                             stringsAsFactors = FALSE))
  }
}

# Print the resulting data frame
result_df <- result_df %>%
  filter(!is.na(State) & State != "United States")

library(ggplot2)

# Calculate the affordability score
result_df$Affordability_Score <- result_df$Rental_Index / result_df$Median_Income * (1 + result_df$Inflation_Average)

# Categorize states based on affordability score
# Categorize states based on affordability score range
result_df$Affordability_Category <- cut(result_df$Affordability_Score,
                                        breaks = c(-Inf, 1000, 1500, Inf),
                                        labels = c("Unaffordable", "Moderately Affordable", "Affordable"),
                                        include.lowest = TRUE)


result_df

```

1. Heatmap of Affordability Scores: The heatmap, also known as a "heatmap" in the code, shows the evolution of a state's affordability score over time. Each heatmap cell represents a state and a certain year, while the fill color represents the affordability category. Green is used to indicate "Affordable," yellow to indicate "Moderately Affordable," and red to indicate "Unaffordable."

We can see how affordability categories have changed over time, across states and years, thanks to the heatmap. It makes it easier to spot states that consistently fall into a particular affordability group as well as those where affordability changes over time. Policymakers and academics can easily identify states that need attention in terms of housing affordability by displaying the data in this way and can then prioritize measures accordingly.

2. Median Real Incomes Over Time: The median real incomes over time for each state are shown in a line plot (referred to as "income_plot" in the code). Each line displays the median real income trend for a particular state, with the years on the x-axis and the median income on the y-axis.

With the help of this image, we can see how the median real income has changed among states and spot any trends or inequalities. It offers insights into the economic health of various states and aids in understanding how median wages have changed over time. This visualization can be used by researchers and policymakers to examine probable causes of income inequality, identify states with large changes in median earnings, and evaluate patterns in income growth.
```{r}
# Heatmap of affordability scores
heatmap <- ggplot(result_df, aes(x = Year, y = State, fill = Affordability_Category)) +
  geom_tile() +
  scale_fill_manual(values = c("Affordable" = "green", "Moderately Affordable" = "yellow", "Unaffordable" = "red")) +
  labs(x = "Year", y = "State", fill = "Affordability Category") +
  theme_classic()
# Display the visualizations
heatmap
```
2. Median Real Incomes Over Time: The median real incomes over time for each state are shown in a line plot (referred to as "income_plot" in the code). Each line displays the median real income trend for a particular state, with the years on the x-axis and the median income on the y-axis.

With the help of this image, we can see how the median real income has changed among states and spot any trends or inequalities. It offers insights into the economic health of various states and aids in understanding how median wages have changed over time. This visualization can be used by researchers and policymakers to examine probable causes of income inequality, identify states with large changes in median earnings, and evaluate patterns in income growth.
```{r}
# Plot median real incomes over time
income_plot <- ggplot(result_df, aes(x = Year, y = Median_Income, group = State)) +
  geom_line() +
  labs(x = "Year", y = "Median Income", title = "Median Real Incomes Over Time") +
  theme_bw()


income_plot


```

Median Income by Year: Plot 1 depicts the median income trend for all states throughout time. It enables us to track changes in median income over time and get knowledge of state-by-state income growth trends.

```{r}
library(ggplot2)

plot1 <- ggplot(result_df, aes(x = Year, y = Median_Income, color = State)) +
  geom_point() +
  labs(x = "Year", y = "Median Income", title = "Median Income by Year") +
  theme_bw()

# Adjust the size of the points if needed
plot1 + geom_point(size = 3)

```

Rental Index by State: Plot 2 displays a box plot that shows how the rental index values are distributed among the states. It aids in comprehending price fluctuation and locating possible outliers or states with higher or lower rental rates.

```{r}
plot2 <- ggplot(result_df, aes(x = State, y = Rental_Index)) +
  geom_boxplot() +
  labs(x = "State", y = "Rental Index", title = "Rental Index by State") +
  theme_bw()

plot2
```

Distribution of Affordability Categories: Plot 3 shows how States are distributed across Affordability Categories. It gives a general summary of how many states fall into each of the three affordability categories—"Affordable," "Moderately Affordable," and "Unaffordable." This image makes it possible to quickly evaluate the affordability environment as a whole.

```{r}
plot3 <- ggplot(result_df, aes(x = Affordability_Category, fill = Affordability_Category)) +
  geom_bar() +
  labs(x = "Affordability Category", y = "Count", title = "Affordability Category Distribution") +
  theme_bw()

plot3
```

Average inflation versus median income:
Plot 4 shows a scatter plot of the median income versus the average inflation rate, with each point denoting a different state. This map makes it easier to spot any possible connections or correlations between state-by-state inflation rates and income levels.
```{r}
plot4 <- ggplot(result_df, aes(x = Median_Income, y = Inflation_Average, color = State)) +
  geom_point() +
  labs(x = "Median Income", y = "Inflation Average", title = "Median Income vs. Inflation Average") +
  theme_bw()

plot4
```



Plot 6 displays a bar chart comparing the number of states in each affordability category across various states. It allows us to track the regional distribution of the affordability categories and pinpoint which states have more or lower representation in each one.

```{r}
plot6 <- ggplot(result_df, aes(x = State, fill = Affordability_Category)) +
  geom_bar() +
  labs(x = "State", y = "Count", title = "Affordability Category by State") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

plot6
```

Rental Index by Affordability Category: Plot 7 makes use of a violin plot to show how the rental index values are distributed among different affordability groups. With an awareness of the rental market within each affordability category, this graphic offers insights on the relationship between affordability and rental costs.

```{r}
plot7 <- ggplot(result_df, aes(x = Affordability_Category, y = Rental_Index)) +
  geom_violin() +
  labs(x = "Affordability Category", y = "Rental Index", title = "Rental Index by Affordability Category") +
  theme_bw()

plot7
```

State-by-State Median Income: Plot 8 displays a bar graph showing the median income for each state. It makes it simple to compare income levels between states, making it easier to determine which states have higher or lower median incomes.

```{r}
plot8 <- ggplot(result_df, aes(x = State, y = Median_Income, fill = State)) +
  geom_bar(stat = "identity") +
  labs(x = "State", y = "Median Income", title = "Median Income by State") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

plot8
```

Inflation Average by Year Plotted as a Bar Graph: The bar graph shows the inflation average by year, with each bar representing a different year. Each bar's height shows the average inflation rate, while the fill color denotes various states.

```{r}
plot9 <- ggplot(result_df, aes(x = Year, y = Inflation_Average, fill = State)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(x = "Year", y = "Inflation Average", title = "Inflation Average by Year") +
  theme_bw()

plot9


```

The analysis of housing affordability trends utilizing information on median household income, inflation rates, and rental indices has, in conclusion, shown the complex housing affordability situation in the United States. The results emphasize the growing gap between median salaries and housing prices, which makes it more challenging for many Americans to finance homeownership.

We have learned a lot about how housing affordability has changed over time, how it differs between states, and how income, inflation, and rental costs are related by utilizing visualizations like heatmaps, line graphs, box plots, scatter plots, and bar charts. Policymakers, researchers, and other stakeholders can use these visualizations to better understand the dynamics of housing affordability and to help them decide how to handle the housing problem.

Significant differences in housing affordability between states have been found by the investigation, with most of states constantly falling into the "Unaffordable" category. This emphasizes the requirement for focused governmental interventions and actions to tackle affordability issues and guarantee that all Americans have access to safe and reasonably priced housing.

The visualizations have also demonstrated how economic factors like inflation affect housing affordability. In order to devise methods to lessen the impact of growing expenses on household budgets, it can be helpful to understand the relationship between the median income and inflation rates.

The report as a whole emphasizes how urgent it is to address the housing affordability situation in the US. It is possible to work toward a future where affordable housing is available to everyone by utilizing data-driven insights and enacting tailored policies, allowing individuals and families to prosper and contribute to vibrant and sustainable communities.

