---
title: "Final Project Assignment#1: PoChun Yang"
author: "PoChun Yang"
description: "Project & Data Description"
date: "04/10/2023"
format:
  html:
    df-print: paged
    toc: true
    code-copy: true
    code-tools: true
    css: styles.css
categories:
  - final_Project_assignment_1
  - final_project_data_description
editor_options: 
  chunk_output_type: console
---

## Overview of the Final Project {#overview-of-the-final-project}

The goal is to tell a coherent and focused story with your data, which answers a question (or questions) that a researcher, or current or future employer, might want to have answered. The goal might be to understand a source of covariance, make a recommendation, or understand change over time. We don't expect you to reach a definitive conclusion in this analysis. Still, you are expected to tell a data-driven story using evidence to support the claims you are making on the basis of the exploratory analysis conducted over the past term.

In this final project, statistical analyses **are not required**, but any students who wish to include these may do so. However, **your primary analysis should center around visualization rather than inferential statistics.** Many scientists only compute statistics after a careful process of exploratory data analysis and data visualization. Statistics are a way to gauge your certainty in your results - **NOT A WAY TO DISCOVER MEANINGFUL DATA PATTERNS**. **Do not run a multiple regression with numerous predictors and report which predictors are significant!!**

## Tasks of Assignment#1 {#Tasks-of-assignment1}

This assignment is the first component of your final project. Together with the later assignments, it make up a short paper/report. In this assignment, you should introduce a dataset(s) and how you plan to present this dataset(s). This assignment should include the following components:

1.  **A clear description of the dataset(s) that you are using.**

2.  **What "story" do you want to present to the audience? In other words, what "question(s)" do you like to answer with this dataset(s)?**

3.  **The Plan for Further Analysis and Visualization.**

We will have a special class meeting on April 12 to review and discuss students' proposed datasets for the final project. **If you want your project being discussed in the class, please submit this assignment before April 12.**

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)
library(readxl)
library(ggplot2)
library(lubridate)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Part 1. Introduction {#describe-the-data-sets}

In this part, you should introduce the dataset(s) and your research questions.

1.  Dataset(s) Introduction:

I have been a devoted follower of basketball for as long as I can remember, and I have always been eager to merge my love for the sport with my interest in data analysis. To achieve this, I used the NBA Stats API to collect a dataset that I believe will be just as captivating for you to work with as it has been for me.

To begin with, I am interested in exploring the careers of my favorite players and analyzing their data. I am not particularly interested in following new star players, so I will analyze the data for their careers. My aim is to gain a deeper understanding of their performances through this report.

Secondly, I am curious to investigate whether height has an impact on a player's rebound skills. I want to determine if there is evidence to support this theory or not.

**Additionally, I want to use the data to analyze the body mass index (BMI) of players. I am interested in exploring whether players who have a high weight are more prone to injuries than players who have a lower BMI. However, this data does not provide injure data. Thus, I would analysis of player with high BMI and investigate how often they are injured.**

Lastly, I want to explore the impact of age on player performance. I would like to investigate whether age has a significant effect on the performance of players or not.

2.  What questions do you like to answer with this dataset(s)?

-   Analysis of several my favorite players

-   Analysis of **Height and Weight** of NBA Players

-   Height and **rebound** relative analyze

-   Analysis players **age with their net rating**

    做summary statistics for以上這些要研究的columns

## Part 2. Describe the data set(s) {#describe-the-data-sets-1}

The dataset that I have compiled using the NBA Stats API includes information on all players who have been a part of an NBA team's roster for the past twenty years. It encompasses various demographic variables such as age, height, weight, and place of birth, along with biographical details like the team they played for, draft year, and round. Furthermore, the dataset contains fundamental box score statistics, such as the number of games played, average points, rebounds, assists, and other relevant information.

1.  read the dataset;

```{r}
df <- read_csv("Po-chunYang_FinalProjectData/all_seasons.csv")

```

The 23 column show players' name, team abbreviation, age, height, weight, college, country, draft round, draft number, games played, points, rebound, assist, net rating, Offensive Rebounds, Defensive Rebounds, Usage Percentage, True Shooting Percentage, Assists Percentage, and season. In addition, there are 12305 rows of data. Those rows shows all the players' data every season.

```{r}
dim(df)
length(unique((df)))
head(df)
colnames(df)
df%>%
  select(player_name)%>%
  n_distinct(.)
#描述有多少columns以及每一行代表什麽（球員）；一共有多少cases（球員)
```

There are a lot of NBA players who play for several years so they have many rows of data and I try to merge all of them. After we tidy the data, I got the 2463 rows of data. Moreover, the summary shows 2463 players' average height, weight, point, rebound, and assist. In addition, I tidy players' data of their all careers with point, rebound, assist, and net ratings. Additionally, I integrate the players' average points, average assists, and average rebounds.

```{r}


summarise(df,height = mean(player_height), weight = mean(player_weight),
          point = mean(pts), rebound = mean(reb), assist = mean(ast))

dp<-mutate(df,point= pts*gp, assist = ast*gp, rebound = reb*gp)
performance<-dp%>%
  group_by(player_name)%>%
  summarise(game_play=sum(gp),total_point = sum(point),
            total_rebound=sum(rebound), total_assist = sum(assist),
            net_rating = mean(net_rating))%>%
  mutate(average_point = total_point/game_play,
         average_rebound = total_rebound/game_play,
         average_assist = total_assist/game_play)

body_data <- df %>%
  group_by(player_name)%>%
  summarise(height = mean(player_height), weight=mean(player_weight))


```

In the storytelling component, you should describe the basic information of the dataset(s) and the variables in a way that corresponds to your descriptive and summary statistics in the above coding component. **DO NOT** simply report the number of rows. Instead, describe the dataset(s) fully by **specifying what each row and column mean.** In other words, your description should be comprehensive and detailed enough for readers to picture or envision the dataset(s) in their brains.

## 3. The Tentative Plan for Visualization {#the-tentative-plan-for-visualization}

-   Analysis of several my favorite players I am interested in exploring the careers of my favorite players and analyzing their data. I am not particularly interested in following new star players, so I will analyze the data for their careers. My aim is to gain a deeper understanding of their performances through this report.

```{r}
Kobe<-df[df$player_name=='Kobe Bryant',]
Durant<-df[df$player_name=='Kevin Durant',]
Curry<-df[df$player_name=='Stephen Curry',]
Thompson<-df[df$player_name=='Klay Thompson',]
Tatum<-df[df$player_name=='Jayson Tatum',]
favorite_player<-rbind(Kobe, Durant, Curry, Thompson, Tatum)
favorite_player<- subset(favorite_player, select = -...1)
player_point<- select(favorite_player, player_name, age, pts)


  
```

-   Analysis of **Height and Weight** of NBA Players Additionally, I want to use the data to analyze the body mass index (BMI) of players. I am interested in exploring whether players who have a high weight are more prone to injuries than players who have a lower BMI. However, this data does not provide injure data. Thus, I would analysis of player with high BMI and investigate how often they are injured.

-   Height and **rebound** relative analyze I am curious to investigate whether height has an impact on a player's rebound skills. I want to determine if there is evidence to support this theory or not.

-   Analysis players **age with their net rating** I want to explore the impact of age on player performance. I would like to investigate whether age has a significant effect on the performance of players or not.

```{r}

```

## Special Note on the role of statistics {#special-note-on-the-role-of-statistics}

Statistical analyses **are not required**, but any students who wish to include these may do so. However, your primary analysis should center around visualization rather than inferential statistics. Many scientists only compute statistics after a careful process of exploratory data analysis and data visualization. Statistics are a way to gauge your certainty in your results - **NOT A WAY TO DISCOVER MEANINGFUL DATA PATTERNS**. Do not run a multiple regression with numerous predictors and report which predictors are significant!!

**Remember: The goal is to tell a story about the data**. For example, you might identify sports where winning athletes are younger or older than average and then try to see if you can find some sort of pattern that accounts for this difference. Or perhaps you might compare a country's performance to GDP and see if this changes over time. The goal is not to get a significant statistical result but to identify an interesting pattern in the data and then extract some sort of meaningful recommendation or information from it.

## Evaluation {#evaluation}

You will be evaluated on both the quality of your source code and your written report, with a greater emphasis on the clarity and details of the description of your dataset(s) and your research questions.
