---
title: "Final Assignment 1"
author: "Janani Natarajan"
date: "05/21/2023"
format:
  html:
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)
library(ggplot2)


knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

# **Part 1**

## Introduction 

This dataset provides information on various cars, including their attributes such as name, year, selling price, kilometers driven, fuel type, seller type, transmission, and previous owners. It covers a wide range of car models and characteristics, offering valuable insights into factors influencing car prices and market dynamics. With data on petrol and diesel cars, different transmission options, and both dealer and individual sellers, this dataset allows for analyzing pricing strategies and understanding the impact of age, mileage, fuel type, and seller type on selling prices. The dataset includes popular car models across various segments, enabling comprehensive analysis of pricing trends and identification of outliers. By studying this dataset, users can uncover patterns, trends, and preferences in the used car market, aiding informed decision-making for buyers, sellers, and industry professionals.

**Source of the dataset(s):**

kaggle.com

## Data Description

-   Name: This column contains the name or model of each car. It provides a unique identifier for each vehicle in the dataset, allowing easy reference and identification.

-   Year: The Year column represents the manufacturing year of each car. It indicates the age or vintage of the vehicle, which is an important factor to consider when evaluating its value and condition.

-   Selling Price: This column contains the listed selling price of each car. It represents the monetary value at which the owner or seller is offering the vehicle for sale. The selling price can vary depending on factors such as brand reputation, model popularity, condition, mileage, and market demand.

-   Kilometers Driven: The Kilometers Driven column records the total distance traveled by each car in kilometers. It provides an insight into the usage and wear-and-tear of the vehicle. Generally, higher mileage suggests that the car has been driven more extensively, potentially impacting its value and condition.

-   Fuel Type: This column specifies the type of fuel used by each car, such as petrol, diesel, or electric. It reflects the energy source that powers the vehicle's engine. Fuel type is an important consideration for buyers, as it affects running costs, environmental impact, and availability of refueling options.

-   Seller Type: The Seller Type column indicates whether the seller is a dealer or an individual. It helps differentiate between professional car dealerships and private individuals selling their cars. Seller type can influence factors such as negotiation flexibility, warranty options, and overall trustworthiness of the transaction.

-   Transmission: This column represents the type of transmission system present in each car. It can be either manual or automatic. Manual transmission requires manual shifting of gears, while automatic transmission shifts gears automatically. Transmission type can impact driving experience, convenience, and preferences of potential buyers.

-   Owner: The Owner column denotes the number of previous owners a car has had. It provides insight into the vehicle's history and ownership stability. Generally, a car with fewer previous owners may be perceived as more desirable and potentially better maintained, which can affect its value.

## Questions the dataset will answer

1.  How many cars in the dataset are fuel-efficient or hybrid?

2.  Are there any correlations between engine size, horsepower, and fuel efficiency?

3.  What is the most common transmission type among the cars?

4.  How does the average price vary across different car makes and models?

5.  Can we predict the price or mileage of a car based on its attributes?

# Part 2 Describe the data set

```{r}
data<-read.csv(file = "/Users/jananinatarajan/Downloads/cardata.csv",
                header=TRUE,
                sep = ","
                )

```

```{r}
dim(data)
```

```{r}
summary(data)
```

```{r}
head(data)
```

```{r}
tail(data)
```

```{r}
mean(data$Selling_Price)
```

```{r}
mean(data$Present_Price)
```

```{r}
median(data$Selling_Price)
```

```{r}
median(data$Present_Price)
```

```{r}
min(data$Kms_Driven)
```

```{r}
max(data$Selling_Price)
```

```{r}
max(data$Present_Price)
```

```{r}
length(unique(data$Car_Name))
```

**Why I choose vehicles as my data set**

The choice of the "cars" dataset as a data set is likely based on several factors. Firstly, cars are a ubiquitous mode of transportation and play a significant role in our daily lives. The dataset provides an opportunity to explore various aspects of the automotive industry, including car makes, models, attributes, and market trends. Additionally, analyzing car data can be relevant for multiple stakeholders, such as car manufacturers, dealerships, buyers, and policymakers. It allows for studying factors like car popularity, performance metrics, fuel efficiency, pricing, and more. The dataset's comprehensive information and its relevance to real-world scenarios make it an attractive choice for conducting analyses and gaining insights into the automotive domain.

##  Part - 3 The Tentative Plan for Visualization

I will be plotting different histograms, box plots, scatter plots and linear graphs

-   Histograms: Histograms are used to display the distribution of a single variable. They are particularly useful for understanding the shape, center, and spread of a dataset. In the context of the car dataset, a histogram can be used to visualize the distribution of variables such as the year of manufacture or the number of doors.

-   Box Plots: Box plots, also known as box-and-whisker plots, provide a summary of the distribution of a continuous variable. Box plots allow for a quick comparison of multiple groups or variables and provide information about central tendency, variability, and potential outliers. In the car dataset, box plots can be used to compare variables like price or mileage across different car makes or models.

-   Scatter Plots: Scatter plots are used to visualize the relationship between two continuous variables. They plot individual data points as dots on a graph, with one variable on the x-axis and the other on the y-axis. Scatter plots help in understanding the nature and strength of the relationship between variables, such as engine size and horsepower or engine size and fuel efficiency. They can reveal patterns like positive or negative correlation, clusters, or outliers.

-   Linear Graphs: Linear graphs, also known as line graphs, are used to depict the relationship between two variables over time or a continuous range. They are particularly useful for showing trends, patterns, and changes in variables. Linear graphs can be used to analyze variables like price or mileage over a period of time or based on a specific continuous variable, such as engine size. They help in identifying patterns of development, growth, or decline.
