---
title: "Final Project"
author: "Xinyang Mao"
description: "Project & Data Description"
date: "05/12/2023"
format:
  html:
    df-print: paged
    toc: true
    code-copy: true
    code-tools: true
    css: styles.css
categories:
  - final_Project
editor_options: 
  chunk_output_type: console
---


```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Introduction

Every year the lives of approximately 1.3 million people are cut short as a result of a road traffic crash. Between 20 and 50 million more people suffer non-fatal injuries, with many incurring a disability as a result of their injury.

Road traffic injuries cause considerable economic losses to individuals, their families, and to nations as a whole. These losses arise from the cost of treatment as well as lost productivity for those killed or disabled by their injuries, and for family members who need to take time off work or school to care for the injured. Road traffic crashes cost most countries 3% of their gross domestic product.

In 2020, a total of 35,766 fatal car accidents occurred on roadways across the United States. Another 1,593,390 crashes resulted in injuries and 3,621,681 caused property damage. That means a total of 5,250,837 collisions happened over the course of a single year.Road crashes are the single greatest annual cause of death of healthy U.S. citizens travelling abroad.

source:
https://www.who.int/news-room/fact-sheets/detail/road-traffic-injuries
https://www.forbes.com/advisor/legal/car-accident-statistics/

The goal of this project is to analyze these data statistically, to find the states with the most accidents, and to explore when accidents are likely to occur, the weather conditions at the time of the accident, and to visualize this: to summarize the information obtained from the analysis, to tell the overall picture of accidents in the United States, and to identify the factors that influence the occurrence of accidents and the severity of accidents.

## Describe the data set(s)

This part contains both a coding and a storytelling component.

```{r}
library(dplyr)
library(DMwR2)
library(lubridate)
library(zoo)
# install.packages('Rmisc')
library(Rmisc)
library(mice)
# install.packages('VIM')
library(VIM)
library(zoo)
library(reshape2)

```
```{r}
library(hrbrthemes)
```
```{r}
data <- read.csv("D:/Files/workspace/R/US_Accidents_Dec21_updated.csv")
head(data)
```

```{r}
colnames(data)
```

```{r}
dim(data)
```

```{r}
length(unique(data))

```


```{r}
print(summarytools::dfSummary(data,
                        varnumbers = FALSE,
                        plain.ascii  = FALSE, 
                        style        = "grid", 
                        graph.magnif = 0.70, 
                        valid.col    = FALSE),
      method = 'render',
      table.classes = 'table-condensed')
```

```{r}
colSums(is.na(data))
```

```{r}
table(is.na(data))
```

```{r}
# show the na value amount for each attribute
mis_col <- colSums(is.na(data))
mis_col <- sort(mis_col,decreasing  = TRUE)
mis_col
```

```{r}
# select the columns which have NA value
mis_col <- mis_col[mis_col>0]
mis_col
```

```{r}
# Showing missing values of column variables
library(ggplot2)
```
```{r}
class(mis_col)
```
```{r}
mis_col_df <- data.frame(
  attributes = names(mis_col),
  missing_amount = mis_col
)
class(mis_col_df)
```
```{r}
mis_col_df
```

```{r}
# Create the bar plot
ggplot(mis_col_df, aes(x = attributes, y = missing_amount)) +
  geom_bar(stat = "identity") +
  labs(title = "Missing Values by Column", x = "attributes", y = "Number of Missing Values") +
  theme_minimal() +
  ylim(c(0, 2000000))
```
```{r}
mis_row <- rowSums(is.na(data))/ncol(data) >=0.1
data_1 <- data[!mis_row, ]
head(mis_row)
```

```{r}
colSums(is.na(data_1))
```

```{r}
dim(data_1)
```

```{r}
data_del <- subset(data_1, select = -c(ID,End_Time,End_Lat,End_Lng,
                                   Distance.mi.,Description,Number,Zipcode,Timezone,Airport_Code,
                                   Weather_Timestamp,Wind_Chill.F.,
                                   Civil_Twilight,Nautical_Twilight,
                                   Astronomical_Twilight))

```

```{r}
# 32 columns
dim(data_del)
```
```{r}
colSums(is.na(data_del))
```
We can see there still are several attributes have na values.
Deal with the NA values
```{r}
#Mean interpolation
func_mean <- function(x){
  x[is.na(x)] <- mean(x,na.rm = TRUE)
}
```
```{r}
data_del$Temperature.F. <- sapply(data_del$Temperature.F.,func_mean)
data_del$Humidity... <- sapply(data_del$Humidity... ,func_mean)
data_del$Pressure.in. <- sapply(data_del$Pressure.in.,func_mean)
data_del$Visibility.mi. <- sapply(data_del$Visibility.mi.,func_mean)
data_del$Wind_Speed.mph. <- sapply(data_del$Wind_Speed.mph.,func_mean)
data_del$Precipitation.in. <- sapply(data_del$Precipitation.in.,func_mean)
```

```{r}
# Write the modified data back to the original data box
data_del[] <- lapply(data_del, function(x) {
  if(is.numeric(x)) replace(x, is.na(x), mean(x, na.rm = TRUE)) else x
})
```

```{r}
colSums(is.na(data_del))
```

```{r}
colnames(data_del)
```
## Total number of accidents by state
```{r}

```

```{r}
data_state <- data_del %>%
  group_by(State) %>%
  dplyr::summarize(count = n()) %>%
  ungroup()
data_state
```
```{r}
# draw accident map
# install.packages('maps')
# install.packages('ggmap')
library(maps)
library(ggmap)
m <- borders("state",colour = "gray50",fill="white")
df2 <- data_del %>% select(Start_Lat,Start_Lng)  
point <- df2[sample(nrow(df2),3000),]
p1 <- ggplot(point,aes(x=Start_Lng,y=Start_Lat)) + m 
p1 + geom_point(color='red',alpha=.6) + 
  labs(title='U.S. Traffic Accident Map')+ 
  theme(plot.title = element_text(hjust = 0.5))

```


The raw data includes data for the five-year period 2016-2021, with the six-year accident totals shown in the table below:
```{r}
data_del$year <- year(data_del$Start_Time)
summary(data_del$year)
```
```{r}
colnames(data_del)
```
```{r}
data_year <- data_del %>%
  group_by(year) %>%
  dplyr::summarize(count = n()) %>%
  ungroup()
data_year
```
Because this dataset is so large,so I will only choose 1 year of data to do the analysis.
```{r}
data_2020 <- data_del[data_del$year == 2020, ]
head(data_2020)
```
```{r}
colnames(data_2020)
```
## Influencing Factors
### Weather condition
The Weather_Condition variable in the data is presented for analysis.

```{r}
data_2020$Severity <- as.factor(data_2020$Severity)
data_2020$Weather_Condition <- as.factor(data_2020$Weather_Condition)
str(data_2020)
```

```{r}
table(data_2020$Weather_Condition)
```
We can see the first one is blank,so we need to select valid weather type to analyze.
```{r}
data_weather <- data_2020[data_2020$Weather_Condition !="", ]
table(data_weather$Weather_Condition)
```
```{r}
weather <- data_weather %>%
  group_by(Weather_Condition) %>%
  dplyr::summarize(count = n()) %>%
  ungroup()
weather
```
Count the number of weather conditions by category and select the 10 most common types of weather
```{r}
weather <- arrange(weather,desc(count))
print(n=20,weather)
```

```{r}
weather <- filter(weather,count > 2000)
weather_10 <- weather$Weather_Condition
weather_10
```
```{r}
df_weather_1 <- filter(data_weather,Weather_Condition %in% weather_10)
df_weather <- subset(df_weather_1,select=c(Weather_Condition,Severity))

```
```{r}
ggplot(df_weather, aes(x=Weather_Condition, fill = Severity)) + 
    geom_bar(position = 'fill') + 
    labs(x = 'Weather', y = 'Severity') + 
    coord_flip() +
    theme_minimal()
```
```{r}
weather_count <- df_weather %>%
  group_by(Severity) %>%
  dplyr::summarize(count = n()) %>%
  ungroup()
df_weather <- left_join(df_weather, weather_count, by = "Severity")
colnames(df_weather)
```
```{r}
ggplot(df_weather, aes(x=Weather_Condition, y= count)) + 
  geom_bar(stat="identity") + 
  labs(x = 'Weather', y = 'Severity') + 
  coord_flip() +
  theme_minimal() +
  facet_wrap(~Severity, ncol = 2)
```

### Visibility
```{r}
data_vis <- subset(data_2020,select=c(Visibility.mi.,Severity))
sum(is.na(data_vis$Visibility.mi.))
```
```{r}
ggplot(data_vis,aes(x=Visibility.mi.)) +
  geom_density(fill="#69b3a2", color="#e9ecef", alpha=0.8) +
    ggtitle("The density of accidents at different visibility levels") +
    theme_ipsum()
```
```{r}
```
```{r}

ggplot(data_vis,aes(x=Visibility.mi.)) +
  geom_histogram(fill="#69b3a2", color="#e9ecef", alpha=0.8) +
    ggtitle("The density of accidents of different levels at different visibility levels") +
    theme_ipsum() +
    facet_wrap(~Severity, ncol = 2)
```

### Time Factor
```{r}
# adjust date format
data_time <- data_del
data_time$Start_Time <- as.POSIXct(data_time$Start_Time)
data_time$month <- month(data_time$Start_Time)
data_time$hour <- hour(data_time$Start_Time)

```
```{r}
head(data_time$month)
head(data_time$hour)
```

```{r}
# Month
df_month <- data_time %>%
  group_by(month) %>%
  dplyr::summarize(mon_count = n()) %>%
  ungroup()
```
```{r}
# hour
df_hour <- data_time %>%
  group_by(hour) %>%
  dplyr::summarize(hour_count = n()) %>%
  ungroup()
```
```{r}
data_time <- left_join(data_time,df_month,by ="month")
data_time <- left_join(data_time,df_hour,by = "hour")
colnames(data_time)
```
```{r}
```
```{r}
# Number of car accidents by month
# create a vector of month names
month_names <- month.abb

ggplot(data_time, aes(x=month, y=mon_count)) +
  geom_line( color="grey") +
  geom_point(shape=21, color="black", fill="#69b3a2", size=6) +
  scale_x_continuous(breaks = 1:12, labels = month_names) +
  labs(x = "Month", y = "Number of Accidents") +
  ggtitle("Number of Car Accidents by Month") +
  theme_ipsum()
```

```{r}
# Number of car accidents by hour
# create a vector of hour
hour_names <- 0:23

ggplot(data_time, aes(x=hour, y=hour_count)) +
  geom_line( color="grey") +
  geom_point(shape=21, color="black", fill="#69b3a2", size=6) +
  scale_x_continuous(breaks = hour_names, labels = hour_names) +
  labs(x = "Hour", y = "Number of Accidents") +
  ggtitle("Number of Car Accidents by Hour") +
  theme_ipsum()
```

### Road condition factor

```{r}
df_road <- subset(data_del,select = c(Amenity,Bump,Crossing,Give_Way,Junction,No_Exit,Railway,Roundabout,Station,Stop,Traffic_Calming,Traffic_Signal,Turning_Loop))
```
```{r}
colnames(df_road)
```
```{r}
df_road <- rbind(colSums(df_road=='True'),colSums(df_road=='False')) %>%
  cbind(Boolean=c('True','False')) %>% 
  data.frame()
```
```{r}
head(df_road)
```
```{r}
data_env <- melt(df_road,
                id.vars = 'Boolean',
                variable.name = 'Conditions',
                value.name = 'Accidents'
)
str(data_env)  
```
```{r}
data_env$Accidents <- as.numeric(data_env$Accidents)
```
```{r}
env_plot <- ddply(data_env,'Conditions',transform,
            percent_num = Accidents/sum(Accidents)*100)
```
```{r}
ggplot(env_plot,aes(x=Conditions,y=percent_num,fill=Boolean)) +
  geom_bar(stat = 'identity',color='blue')+ 
  scale_fill_brewer(palette = 11)+
  geom_text(aes(label=paste(round(percent_num),"%"),vjust=-0.3))+
  labs(title='Environmental factors stacking bar chart')+ 
  theme_minimal()

```

## Evaluation {#evaluation}

You will be evaluated on both the quality of your source code and your written report, with a greater emphasis on the clarity and details of the description of your dataset(s) and your research questions.
