---
title: "Final Project: Connor Landreth"
author: "Connor Landreth"
description: "Determinants of Sleep Efficiency"
date: "05-18-2023"
format:
  html:
    df-print: paged
    toc: true
    code-copy: true
    code-tools: true
    css: styles.css
categories:
  - Final_Project_Submission
  - Sleep_Efficiency
  - Connor Landreth
editor_options: 
  chunk_output_type: console
editor: 
  markdown: 
    wrap: 72
---

```{r setup}
#| label: setup
#| warning: false
#| message: false

#load in relevant packages
library(psych)
library(tidyverse)
library(ggplot2)
library(ggthemes)
library(summarytools)
library(kableExtra)
library(dplyr)
library(reshape2)
library(SmartEDA)


knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

# Introduction

Today, there are a slew of sleep inhibitors that are hindering young
individuals from obtaining adequate and or efficient sleep. Today, there
is overwhelmingly easy access to alcohol, caffeine, and nicotine, and
they have all evolved in their own ways to become very socially
accepted. We will look at which of these negative variables, as well as
a more positive variable (exercise), affect sleeping trends in young
individuals, aged 21-35. The data set, which I will discuss in greater
detail below, analyzes sleep efficiency and time spent sleeping. It also
notes whether or not/how much alcohol and caffeine was consumed the
prior day. Additionally, it looks at frequency of exercise, and whether
or not the individual smokes. All this combined will enable us to draw
insights on what some drivers of poor or strong sleep efficiency are.
Doing so may curate actionable results for individuals looking to
improve sleep efficiency and duration by taking into account the
findings here. As someone who has slept poorly essentially forever, I am
very interested to see what trends emerge.

# Data Set Introduction & Background

The data set contains information about a group of test subjects and
their sleep patterns. Each test subject is identified by a unique
"Subject ID" and their age and gender are also recorded. The "Bedtime"
and "Wake-up time" features indicate when each subject goes to bed and
wakes up each day, and the "Sleep duration" feature records the total
amount of time each subject slept in hours. The "Sleep efficiency"
feature is a measure of the proportion of time spent in bed that is
actually spent asleep. The "REM sleep percentage", "Deep sleep
percentage", and "Light sleep percentage" features indicate the amount
of time each subject spent in each stage of sleep. The "Awakenings"
feature records the number of times each subject wakes up during the
night. Additionally, the data set includes information about each
subject's caffeine and alcohol consumption in the 24 hours prior to
bedtime, their smoking status, and their exercise frequency. This will
all be outlined in a descriptive table down below, with the addition of
measurement scale.

The data set provided was collected as part of a study conducted in the
UK by a research team from The University of Oxfordshire[^1]. The study
aimed to investigate the impact of lifestyle factors such as caffeine,
alcohol, and exercise on sleep patterns and sleep quality. The research
team recruited participants from the local community and collected data
over a period of several months. The data was collected using a
combination of self-reported surveys, actigraphy, and polysomnography
which is a sleep monitoring technique. The data was then analyzed to
understand the relationship between lifestyle factors and sleep patterns
and to identify any potential areas for intervention to improve sleep.

[^1]: "Sleep Efficiency Dataset." University of Oxfordshire , 14 Jan.
    2023

I am interested in finding sleep inhibitors for young adults, so I'll
filter for ages from 21-35. I would like to look at those under 35 more
generally, but I cannot go below 21 because nicotineand alcohol are not
legally available for individuals under the age of 21. The results,
therefor, could be erroneous without filtering out below 21. Also note,
35 was an abstract selection and where I chose the cut-off for "young
adults", although this is not an objectively true measure of "young", of
course.

# Data Description

I am loading in the raw data, which is rather tidy to begin with. I will
be altering the data to fit my specific analysis, but spending a bit
more time analyzing and less cleaning. Any NAs will be dropped and
disregarded, because there is a very slim percentage of NAs present in
the dataset, so the results will not be impacted from disregarding them.

```{r read_data}
getwd()
#Set wd to folder with data
setwd("C:/Github Projects/601_Spring_2023/posts/ConnorLandreth_FinalProjectData")
#read in raw sleep data
sleep <- read.csv("Sleep_Efficiency.csv")
#View(sleep)
#head(sleep)
#View Dimensions of dataframe 
dim(sleep)
```

Dimensions on the raw sleep data include 452 observations over 15
variables

```{r Summarize}
#evaluate unique values and summarize to evaluate number of respondents, age range, and gender options
length(unique(sleep$ID))
length(unique(sleep$Age))
length(unique(sleep$Gender))

#I am going to remove bedtime and wakeup time, as the information is irrelevant for this analysis and is encapsulated in the sleep duration column. Then I will recode variables to make them numeric to summarize.

# delete columns 4 & 5
sleep <- sleep[, -c(4:5)] 

#Make gender and smoking status binary to get breakdown in summary stats
sleep$Gender[sleep$Gender=="Male"] <- "0"
sleep$Gender[sleep$Gender=="Female"] <- "1"
sleep$Smoking.status[sleep$Smoking.status=="No"] <- "0"
sleep$Smoking.status[sleep$Smoking.status=="Yes"] <- "1"
sleep$Gender <- as.numeric(sleep$Gender)
sleep$Smoking.status <- as.numeric(sleep$Smoking.status)

#sleep <- n_gender = as.numeric(sleep$Gender)

#use summarytools to view summary of each variable (More aesthetically pleasing/organized than base R)
```

```{r}
describe(sleep)

```

There are 452 unique IDs, so each row is an individual. Ages 9-69 are
included across the 452 subjects, and 2 genders. There are four more men
than women, but very close to an even split.

Some important notes to summarize variables, the mean age here is just
over 40, so we will filter to look solely at young adults later. Mean
sleep duration is about 7.5 hours.

Below, I will present the relevant variables in a more consumable format
using kable that will break down these variables by meaning and
measurement . I'll do this by creating a data frame.

```{r Index}

# Create variables
Variables <- c('Age', 'Gender', 'Sleep Duration', 'Sleep Efficiency', 'REM Sleep percentage', 'Deep Sleep Percentage', 'Light Sleep Percentage', 'Awakenings', 'Caffeine Consumption', 'Alcohol Consumption', 'Smoking Status', 'Exercise Frequency')
# Add meaning for each variable
Meaning <- c('Age of the test subject', 'Biological gender of subject', 'the total amount of time the test subject slept (in hours)', 'Percentage of time in bed spent sleeping', 'Length of sleep spent in REM portion of sleep cycle', 'Length of time spent in the deep sleep portion of the sleep cycle', 'Length of time spent in the light sleep portion of the sleep cycle', 'the number of times the test subject wakes up during the night', 'the amount of caffeine consumed in the 24 hours prior to bedtime (in mg)', 'the amount of alcohol consumed in the 24 hours prior to bedtime (in Standard Drinks)', ' whether or not the test subject smokes', ' the number of times the test subject exercises each week')
# Add measurement scale for each variable 
Measurement.Scale <- c('Continuous', 'Binary(male and female)', 'Continuous','Continuous', 'Continuous','Continuous', 'Continuous', 'Discrete', 'Continuous','Continuous', 'Binary(yes or no)', 'Discrete')

# Join the variables to create a data frame
Index <- data.frame(Variables, Meaning, Measurement.Scale)
View(Index)

```

Finally, we will use our new data frame to create another table with all
the new information on each variable.

```{r Variable Summary}
Index %>% 
kable(Variables, format = "html") %>%
  kable_styling() %>%
  column_spec(2, bold = TRUE) %>% 
 # column_spec(1, background = "lightblue") %>% 
  row_spec(1:12, color = "black") %>%
  row_spec(0, bold = TRUE, font_size = 16, color = "black", 
           background = "plum") #%>%
  #row_spec(1:10, background = "lightpink1")
```

# Questions & Analysis Plan

I want to answer two main questions, one on a specific covariate, and
one more broad:

1.  Does moderate alcohol consumption (2-3 standard drinks) improve
    sleep efficiency in young adults?
2.  What are the main inhibitors of sleep efficiency in young adults?

Next, I'll pick from the desired variables above and omit those that
won't lend to the analysis. I am removing sleep and wake time because
those give me the sleep duration, and that is already a variable. Also,
ID won't be necessary, nor will the rem, light, or any other sleep
factor beyond duration and efficiency.

```{r Filter}
#use select and filter to omit undesirable columns and hone in on ages 21 to 35. Again, to reiterate, I want to look at young adults, but I cannot go below 21 because smoking and alcohol is not legally available for individuals under the age of 21. The results, therefor, could be erroneous without filtering out below 21. 
n_sleep <- sleep %>% 
  select(Age, Gender, Sleep.duration, Sleep.efficiency, Awakenings, Caffeine.consumption, Alcohol.consumption, Smoking.status, Exercise.frequency) %>% 
  filter(Age <= "35"& Age>="21")
#check
head(n_sleep)
```

That mutation left me with 155 observations and nine nine variables,
which is a bit low, but enough to figure out what causes young adults to
miss out on sleep. Let us look at some histograms to help visualize
averages and trends in the data.

```{r Caffeine Hist}
#Building a histogram of caffeine consumption to evaluate dispersion of consumption
hist_caffeine<-ggplot(n_sleep, aes(x= Caffeine.consumption)) +
  geom_histogram() +
  labs(title = "Histogram of Caffeine Consumption", x = 'Caffeine Consumption (mg.)', y="Count") +
theme_economist()

hist_caffeine

```

In this filtered out sample of only people 21-35, caffeine consumption
is concentrated on the lower end of the spectrum, with nearly all
individuals consuming under 100mg for the day, with many consuming zero.

```{r Alcohol Hist}
#Building a histogram of alcohol consumption to evaluate dispersion of consumption

hist_alcohol<-ggplot(n_sleep, aes(x= Alcohol.consumption)) +
  geom_histogram() +
  labs(title = "Histogram of Alcohol Consumption", x = 'Alcohol Consumption (Standard Drinks)', y="Count") +
theme_economist()

hist_alcohol
```

Alcohol consumption is on the lesser side, with the mass majority
consuming little to no alcohol. Many are between zero and one standard
drink for the day.

Down below as we look at smoking, I will also mutate those "yes" and
"no" responses to 0 and 1 to subset more easily moving forward as I did
above for the summary stats.

```{r Smoking Hist}
#Not building a histogram for smoking because it is clear for in the summary statistics, as it's binary.
```

The majority of young people do not smoke, while roughly sixty people 35
and under do engage in smoking.

```{r Exercize Hist}
#Building a histogram of alcohol consumption to evaluate dispersion of excersize frequency
hist_exercise<-ggplot(n_sleep, aes(x= Exercise.frequency)) +
  geom_histogram() +
  labs(title = "Histogram of Exercise Frequency", x = 'Exercise Frequency', y="Count") +
theme_economist()

hist_exercise
```

For exercise, most work out 0-1 times per week, with a decreasing trend
from 1 to 5 times per week.

```{r Duration Hist}
#Building a histogram of alcohol consumption to evaluate dispersion of sleep duration

hist_sleep_duration<-ggplot(n_sleep, aes(x= Sleep.duration)) +
  geom_histogram() +
  labs(title = "Histogram of Sleep Duration", x = 'Sleep Duration (Hours)', y="Count") +
theme_economist()

hist_sleep_duration

```

Sleep duration sees a slightly more normal distribution with it's mean,
again, falling right around 7.5 hours a night.

```{r Efficiency Hist}
#Building a histogram of alcohol consumption to evaluate dispersion of sleep efficiency

hist_sleep_efficiency<-ggplot(n_sleep, aes(x= Sleep.efficiency)) +
  geom_histogram() +
  labs(title = "Histogram of Sleep Efficiency", x = 'Sleep Efficiency', y="Count") +
theme_economist()

hist_sleep_efficiency
```

The distribution of sleep efficiency is a bit left skewed, seeing sleep
efficiency rated all across the board, but a bit more concentrated on
the efficient end. There is a spike at 0.5, signifying a large sum of
individuals with inefficient sleep.

Overall, based on the trends I can already see emerging and the
questions I hope to answer, my plan for analysis, more specifically, is
to use **geom_point/jitter** to look at sleep efficiency by gender. I
will facet wrap by gender in the visualizations and look at the effect
alcohol, caffeine, smoking, and exercise has on sleep efficiency for
each gender of the young adult subset. In addition, I will compare sleep
efficiency of non-smokers vs smokers using a set of **geom_boxplot**,
again, facet wrapped by gender.

# Analysis & Visualizations



I will look at the comparison between high sleep efficiency/duration and
alcohol/caffeine/exercise/smoking. I'll start by filtering sleep
efficiency to be greater than the mean and look at consumption levels
for those high efficiency sleepers.

::: {#Alcohol .panel-tabset}
# Question 1: Alcohol

This section will answer question 1, 'Does moderate alcohol consumption
(2-3 standard drinks) improve sleep efficiency?'

```{r alc_by_age}
n_sleep$Gender[n_sleep$Gender=="0"] <- "Male"
n_sleep$Gender[n_sleep$Gender=="1"] <- "Female"

n_sleep %>% 
  ggplot(aes(Alcohol.consumption, Age, color=Gender, fill = Gender))+
  #jitter vs point will provide a nice random variation in points
  geom_jitter(size=2, alpha = 0.5)+
  #add range
  geom_smooth()+
  #facet wrap by gender will allow us to look at each gender separately tabset
 
  facet_wrap(~Gender)+
  labs(x='Alcohol Consumption', y='Age') +
  theme_linedraw()+
  labs(title="Alcohol Consumption by Gender & Age (21-35)")
```

For females, it appears the younger the women are, the more they tend to
drink. The mean age for those who don't drink at all is 28, and the
graph peaks at the intersection of roughly age 31 and one standard
drink. As standard drinks increase, mean age decreases. This is not
surprising.

For men, the trend is less intense, showing a mean age of closer to 27
who do not drink at all, but the mean age here stay more steady as
standard drinks increase, with the mean age for those who consume closer
to five drinks around 25 years old, so a much smaller range than the
female data. Next, I'll explore the consumption trends for efficient
sleepers between the ages of 21-35.

```{r S.E._Alcohol}
summary(n_sleep)
# The mean sleep efficiency for these individuals is .7889, so I will make that the baseline for "good" sleep efficiency
  

n_sleep %>% 
  filter(Sleep.efficiency >= ".7889") %>% 
  ggplot(aes(Alcohol.consumption,Sleep.efficiency, color=Gender))+
  geom_jitter(size=2, alpha = 0.5)+
  geom_smooth()+
  facet_wrap(~Gender)+
  labs(x='Alcohol Consumption', y='Sleep Efficiency') +
  theme_linedraw()+
  labs(title="Alcohol Consumption & Sleep Efficiency by Gender")

```

These results show, generally speaking, there is a trend that shows more
alcoholic drinks translating to less sleep efficiency for both males and
females. Males do show a resurgence in sleep efficiency from 2 to 3
drinks, then even goes on to gradually increase as intoxication goes up.
The way I interpret this is that women in the sample all experience
reduced sleep efficiency the more they drink, but for men, two drinks
actually sees the nadir of sleep efficiency.

Perhaps because the mind often becomes more active off of 1-2 drinks,
men are experiencing increased mental stimulus and activity, making it
harder to sleep[^2]. Then as they get up to 3-4 drinks (likely a bit
drunk and relaxed), efficiency goes back up, but still not as high as it
would be with no drinks. So in conclusion, 3 drinks doesn't see the
worst sleep efficiency, but the highest efficiency comes from no alcohol
at all.


[^2]: Leadem, Rose. "11 Ways Drinking Alcohol Can Make You Smarter,
    Healthier and More Creative." Entrepreneur, 25 Oct. 2017,
    www.entrepreneur.com/living/11-ways-drinking-alcohol-can-make-you-smarter-healthier/303584

Now we will look at the other variables below.

# Question 2: Other Variables

This section will answer my second question, 'What are the main
inhibitors of sleep efficiency?'

```{r S.E._Caffeine}
n_sleep %>% 
    filter(Sleep.efficiency >= ".7889") %>% 
  ggplot(aes(Caffeine.consumption,Sleep.efficiency, color=Gender))+
  geom_jitter(size=2, alpha = 0.5)+
  #width = x
  geom_smooth()+
  facet_wrap(~Gender)+
  labs(x='Caffeine Consumption', y='Sleep Efficiency') +
  theme_linedraw()+
  labs(title="Caffeine Consumption & Sleep Efficiency by Gender")
```

These are interesting results. For females, the most caffeine any of
these high-efficiency-sleeping women consume a day is 75mg, and most
consume less. For the few young women who do consume 75mg, their average
sleep efficiency is the highest. I would be inclined to surmise that
caffeine effects them little, or they drink it only in the early
morning.

For men, most consume none, but oddly, as consumption increases, so too
does sleep efficiency. Again though, much likely has to do with time of
day caffeine is consumed, so I am inclined to once again believe these
individuals aren't effected much by caffeine, so are able to consume
more. Overall though, all of these high-efficiency-sleepers consume a
reasonable amount, and sleep very well. Thirdly, we'll look at exercise.
How often are these efficient sleepers exercising?

```{r S.E._Exercise}
#since we mutated gender to be binary earlier, we will revert back to the original names of 'male', 'female' for the visualization. 
#n_sleep$Gender[sleep$Gender=="0"] <- "Male"
#n_sleep$Gender[sleep$Gender=="1"] <- "Female"

n_sleep %>% 
  ggplot(aes(Exercise.frequency,Sleep.efficiency, color=Gender))+
  geom_jitter(size=2, alpha = 0.5)+
  geom_smooth()+
  facet_wrap(~Gender)+
  labs(x='Exercise Frequency', y='Sleep Efficiency') +
  theme_linedraw()+
  labs(title="Exercise Frequency & Sleep Efficiency by Gender")
```

Of this subset of male sleepers, the least efficient are those who do no
exercise at all. For women, the least effiicient appear to exercise
onece per week and most efficient female sleepers appear to exercise
between 3-4 times per week. For men, The maximum amount of days these
efficient sleepers exercised was 4 days. 4 days also led to maximum
sleep efficiency at over 0.9. So, it appears for both men an women,
although they follow different paths to get there, four days of exercise
yields maximum sleep efficiency.

Lastly, I will take a look at smoking category. First, I will use
n_sleep data set to look at all efficiency, then filter for smoking and
see if the trends change. I would imagine those who smoke wake up more
often and sleep less efficiently because their body is craving
nicotine/. We will look at two sets of boxplots. First is derived from
the original raw data and includes all 452 subjects. It splits up by
gender again and looks at smokers vs. non-smokers sleep efficiency. The firm horizontal line shows the mean overall, while the fluid, curved line shows the mean sleep efficiency for non-smokers on the left side and the mean efficiency for smokers on the right side. They all show decreasing trends. 

```{r S.E._Smoking_OG}
#For the binary smoking, boxplots will help us see the mean sleep efficiency between smokers and non-smokers mroe easily. 
sleep$Gender[sleep$Gender=="0"] <- "Male"
sleep$Gender[sleep$Gender=="1"] <- "Female"
sleep$Gender <- as.character(sleep$Gender)

sleep %>% 
  ggplot(aes(Smoking.status,Sleep.efficiency, color=Gender, fill = Gender))+
  geom_boxplot(size=2, alpha = 0.5)+
  geom_smooth()+
  facet_wrap(~Gender)+
  labs(x='Smoking Status', y='Sleep Efficiency') +
  theme_linedraw()+
  labs(title="Smoking Status & Sleep Efficiency by Gender (Whole Data Set)")
```

The following set of boxplots references n_sleep, and therefor
specifically 21-35 year old individuals.

```{r S.E._Smoking_new}

n_sleep$Gender[n_sleep$Gender=="0"] <- "Male"
n_sleep$Gender[n_sleep$Gender=="1"] <- "Female"
n_sleep$Gender <- as.character(n_sleep$Gender)

n_sleep %>% 
  ggplot(aes(Smoking.status,Sleep.efficiency, color=Gender, fill = Gender))+
  geom_boxplot(size=2, alpha = 0.5)+
  geom_smooth()+
  facet_wrap(~Gender)+
  labs(x='Smoking Status', y='Sleep Efficiency') +
  theme_linedraw()+
  labs(title="Smoking Status & Sleep Efficiency by Gender (21-35)")
```
:::


# Conclusion

In general, young adults (aged 21-35) saw enhanced sleep efficiency
(proportion of time spent in bed actually spent asleep) when they had
higher exercise rates and lower alcohol consumption (with the male 2-3
drink caveat). The genesis of this analysis is that exercise for anyone
(male or female) can promote higher sleep efficiency, that amount of
exercise will likely vary from person to person. For young men, smoking
is a major sleep inhibitor. For young women, alcohol appears to be the
greatest sleep inhibitor. The remaining paragraphs blow will draw
further conclusions from the analysis for each specific variable.

The fact 2-3 drinks leads to slightly better sleep in men is not a
promotion or sign to go drink alcohol every night. This evidence
actually suggests that while alcohol may help you get to sleep quicker,
it often leads to reduced sleep quality.[^3] For women and men both,
zero drinks saw optimal sleep quality. For women, every additional drink
led to lower and lower quality sleep, with a very slight increase in
quality around 5 drinks. It's probable that 5 drinks would lead a woman
to a rather high level of intoxication and result in a much deeper sleep
and the inability to even stay awake once their head hits the pillow.
Considering our definition of sleep efficiency (time in bed spent
sleeping), the slight increase makes sense.

[^3]: \[1\] Roehrs, T. and Roth, T., (2001) Sleep, sleepiness, and
    alcohol use. Alcohol research and Health, 25(2), pp.101-109.

Next, for caffeine, this data is not the most useful because the mass
majority of men consume zero caffeine and they have the worst sleep
quality. Then we see an increase in sleep efficiency the more caffeine
men consume. Again, I think if the sample size were larger, we'd see a
different trend emerge, and that the high caffeine consumers are more
immune to caffeine or they consume solely in the early morning, so their
sleep is hardly affected. Similarly, the female graph shows odd results.
No women consumed more than 75mg of caffeine. 0mg and 75mg show similar
sleep quality with a substantial dip at 50mg. Similar to the men, the
heavier consumers are in the minority, so the results may be skewed.

For smokers, we looked at the sample for all smokers/non-smokers and
smokers/non-smokers aged 21-35. For the entire data set, non-smoking
females saw approximately 13% higher sleep efficiency than female
smokers, while the mean benefit for non-smoking men was negligible at
only \~1-2%. For those aged 21-35, women saw a less substantial change,
with non-smoking females experiencing roughly a 5% higher sleep
efficiency rate. Meanwhile, men saw a glaring difference, with
non-smoking men aged 21-35 operating at 83-84% mean sleep efficiency and
smoking men of the same age group seeing a mean of under 60%, which is
far worse off than non-smokers. This means that smoking is a massive
inhibitor of sleep efficiency in young men. Smoker vs non-smoker sleep
efficiency for women tightened with the age filter, but vastly widened
for men.

Lastly, for exercise, there is a sporadic/widespread sleep efficiency
recorded for all days listed. For example, men who worked out just one
day showed superior sleep efficiency to men who worked out two days,
then the trend decreased further at three days before increasing greatly
again at four days. Despite this tumultuous journey across the graph
plane, the line still tells us that from 1-4 days of working out, sleep
efficiency was better for those individuals than it was for those who
didn't work out at all. So, we can't say how many days you should work
out, but we can say you should try a different mix of days and be
cognizant of your sleep quality week to week. One can likely find the
optimal number of weekly workouts with this strategy, and the answer
does not appear to be zero.

# Limitations

The evidence against caffeine was inconclusive, but many caffeine
consumers were high-efficiency sleepers. Overall, a larger sample could
have led to a spread with greater validity, but I believe many of these
trends (alcohol, smoking, excessive) would likely be perpetuated with a
greater sample.

Additionally, we know people consumed alcohol and caffeine, but there is
no data that tells us when they consumed it. That is a huge gap because
if one drank 5 beers at between 11pm-12am, they'd probably pass out and
have high sleep efficiency vs someone who had 4 beers from the hours of
1pm-10pm. Similarly for caffeine, someone drinking 75mg at 8am will not
sleep at the same rate of efficiency as someone who consumed 75mg at
10pm then proceeded to try try to sleep a couple hours later. The
inclusion of this data could certainly lend to a more precise analysis.

#### References

```{r}
citation()
citation("tidyverse")
RStudio.Version()
```

