---
title: "Challenge 8"
author: "Harsha Kanaka Eswar Gudipudi"
description: "Joining Data"
date: "05/16/2022"
format:
  html:
    toc: true
    code-copy: true
    code-tools: true
categories:
  - challenge_8
  - Harsha Kanakeswar Gudipudi
  - faostat
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)
library(ggplot2)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Challenge Overview

Today's challenge is to:

1)  read in multiple data sets, and describe the data set using both words and any supporting information (e.g., tables, etc)
2)  tidy data (as needed, including sanity checks)
3)  mutate variables as needed (including sanity checks)
4)  join two or more data sets and analyze some aspect of the joined data

(be sure to only include the category tags for the data you use!)

## Read in data

Read in one (or more) of the following datasets, using the correct R package and command.

  - military marriages ⭐⭐
  - faostat ⭐⭐
  - railroads  ⭐⭐⭐
  - fed_rate ⭐⭐⭐
  - debt ⭐⭐⭐
  - us_hh ⭐⭐⭐⭐
  - snl ⭐⭐⭐⭐⭐

```{r}
eggs_chicken_df <- read_csv("_data/FAOSTAT_egg_chicken.csv")
livestock_df <- read_csv("_data/FAOSTAT_livestock.csv")
dairy_df <- read_csv("_data/FAOSTAT_cattle_dairy.csv")
```
Printing the first five rows of every dataset
```{r}
head(eggs_chicken_df) 
head(livestock_df) 
head(dairy_df) 
```

Understanding the dimensions of data set 1
```{r}
dim(eggs_chicken_df)
```
Understanding the dimensions of data set 2
```{r}
dim(livestock_df)
```
Understanding the dimensions of data set 3
```{r}
dim(dairy_df)
```

Understanding the columns of data set 1
```{r}
colnames(eggs_chicken_df)
```
Understanding the columns of data set 2
```{r}
colnames(livestock_df)
```
Understanding the columns of data set 3
```{r}
colnames(dairy_df)
```




### Briefly describe the data

I am using three data sets having information on eggs and chicken, livestock and dairy producing animals. The data sets contain information on livestock, live animals, eggs and chicken country wise. I would like to combine these three datasets to do an exploratory analysis country wise, item code wise etc.

## Tidy Data (as needed)

Is your data already tidy, or is there work to be done? Be sure to anticipate your end result to provide a sanity check, and document your work here.

Checking to see how many entries in the 1st dataset has null values.
```{r}
sum(is.na(eggs_chicken_df))
```
Checking to see how many entries in the 1st dataset has null values.
```{r}
sum(is.na(livestock_df))
```

Checking to see how many entries in the 1st dataset has null values.
```{r}
sum(is.na(dairy_df))
```


The datasets are not very tidy but so I would like to clean them.

Cleaning the 3 data sets
```{r}
eggs_chicken_df<- na.omit(eggs_chicken_df)
livestock_df<- na.omit(livestock_df)
dairy_df<- na.omit(dairy_df)
```

Understanding the dimensions of the data sets post cleaning.
```{r}
dim(eggs_chicken_df)
dim(livestock_df)
dim(dairy_df)
```

##Join Data
```{r}
dairy_livestock_df <-full_join(dairy_df,livestock_df)

dairy_livestock_df
```
Understanding the columns of Dairy-Livestock combined data frame.
```{r}
colnames(dairy_livestock_df)
dim(dairy_livestock_df)

```


```{r}
dairy_livestock_eggschicken_df <- full_join(dairy_livestock_df,eggs_chicken_df)
dairy_livestock_eggschicken_df

```
Understanding columns of overall combined data
```{r}
colnames(dairy_livestock_eggschicken_df)

```

Understanding dimensions of overall combined data
```{r}
dim(dairy_livestock_eggschicken_df)
summary(dairy_livestock_eggschicken_df)
```
 
With the final combined data set which was created by joining 3 individual data sets, we can answer a lot of questions. As an example, I attempted to see how the total value of Milk Production in Afghanistan is changing over years.I observed an exponential increase in terms of value as illustrated in the graph below.
```{r}

milk_production <- dairy_livestock_eggschicken_df %>%
  filter(Element == "Production", Item == "Milk, whole fresh cow", Area == "Afghanistan" )

# Plotting the Milk Production over the years
ggplot(milk_production, aes(x = Year, y = Value)) +
  geom_line() +
  labs(x = "Year", y = "Milk Production") +
  ggtitle("Milk Production in Afghanistan") +
  theme_minimal()
```


