---
title: "Challenge 2"
author: "Harsha Kanaka Eswar Gudipudi"
description: "Data wrangling: using group() and summarise()"
date: "05/15/2023"
format:
  html:
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
  - challenge_2
  - Harsha Kanaka Eswar Gudipudi
  - birds.csv
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Challenge Overview

Today's challenge is to

1)  read in a data set, and describe the data using both words and any supporting information (e.g., tables, etc)
2)  provide summary statistics for different interesting groups within the data, and interpret those statistics

## Read in the Data

Read in one (or more) of the following data sets, available in the `posts/_data` folder, using the correct R package and command.

-   railroad\*.csv or StateCounty2012.xls ⭐
-   FAOstat\*.csv or birds.csv ⭐⭐⭐
-   hotel_bookings.csv ⭐⭐⭐⭐

```{r}
df <- read_csv('_data/birds.csv', show_col_types = FALSE)
head(df)
```

Add any comments or documentation as needed. More challenging data may require additional code chunks and documentation.

## Describe the data

Using a combination of words and results of R commands, can you provide a high level description of the data? Describe as efficiently as possible where/how the data was (likely) gathered, indicate the cases and variables (both the interpretation and any details you deem useful to the reader to fully understand your chosen data).

Description : This data description presents information on live bird stocks like chicken, geese,etc in different areas such as Afganisthan, Albania, Algeria..etc, for the years from 1961 to 2018.

```{r}
#| label: summary
summary(df)
dim(df)
```

All varitey of live stocks in the data:

```{r}
unique_items <- unique(df$Item)
num_unique_items <- length(unique_items)
cat(paste("There are", num_unique_items, "unique Varitey of animals present in the dataset"))
cat(paste(unique_items, collapse = ", "))
```
The data is collected from various areas like:

```{r}
unique_items <- unique(df$Area)
num_unique_items <- length(unique_items)
cat(paste("Total no of areas: ", num_unique_items))
cat(paste(unique_items, collapse = ", "))
```


## Provide Grouped Summary Statistics

Conduct some exploratory data analysis, using dplyr commands such as `group_by()`, `select()`, `filter()`, and `summarise()`. Find the central tendency (mean, median, mode) and dispersion (standard deviation, mix/max/quantile) for different subgroups within the data set.

```{r}
library(dplyr)

birds_grouped <- df %>%
  group_by(Year)
  
print("Data grouped by year:")
print(birds_grouped)

# Filter the data to include only chickens
chickens_only <- birds_grouped %>%
  filter(Item == "Chickens")
  
print("Data filtered to include only chickens:")
print(chickens_only)

# Calculate the mean, median, and mode of chicken stocks for each year
chicken_summary <- chickens_only %>%
  summarise(mean_stocks = mean(Value, na.rm = TRUE),
            median_stocks = median(Value, na.rm = TRUE),
            mode_stocks = as.numeric(names(sort(-table(Value))))[1])

print("Summary statistics of chicken stocks:")
print(chicken_summary)

# Calculate the minimum, maximum, and quantiles of chicken stocks for each year
chicken_range <- chickens_only %>%
  summarise(min_stocks = min(Value, na.rm = TRUE),
            max_stocks = max(Value, na.rm = TRUE),
            q1_stocks = quantile(Value, 0.25, na.rm = TRUE),
            q3_stocks = quantile(Value, 0.75, na.rm = TRUE))

print("Range of chicken stocks:")
print(chicken_range)
```

### Explain and Interpret

Be sure to explain why you choose a specific group. Comment on the interpretation of any interesting differences between groups that you uncover. This section can be integrated with the exploratory data analysis, just be sure it is included.

Here, I choose to group the data by year to see if there are any interesting trends in chicken stocks over time .The chicken_summary and chicken_range tables provide useful insights into the central tendency and dispersion of chicken stocks over years. we can see the highest and lowest mean, median, and mode chicken stocks, as well as the minimum, maximum, and quartile ranges. We can also compare the trends across different years and see if there are any interesting patterns.