---
title: "Challenge 1"
author: "Ishan Bhardwaj"
desription: "Reading in data and creating a post"
date: "03/14/2023"
format:
  html:
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
  - challenge_1
  - Ishan Bhardwaj
  - railroads
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)
library(readr)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Challenge Overview

Today's challenge is to

1)  read in a dataset, and

2)  describe the dataset using both words and any supporting information (e.g., tables, etc)

## Read in the Data

Read in one (or more) of the following data sets, using the correct R package and command.

-   railroad_2012_clean_county.csv ⭐
-   birds.csv ⭐⭐
-   FAOstat\*.csv ⭐⭐
-   wild_bird_data.xlsx ⭐⭐⭐
-   StateCounty2012.xls ⭐⭐⭐⭐

Find the `_data` folder, located inside the `posts` folder. Then you can read in the data, using either one of the `readr` standard tidy read commands, or a specialized package such as `readxl`.

```{r}
rail <- read_csv("_data/railroad_2012_clean_county.csv")
rail
```

Add any comments or documentation as needed. More challenging data sets may require additional code chunks and documentation.

## Describe the data

Using a combination of words and results of R commands, can you provide a high level description of the data? Describe as efficiently as possible where/how the data was (likely) gathered, indicate the cases and variables (both the interpretation and any details you deem useful to the reader to fully understand your chosen data).

```{r}
dim(rail)
```

This dataset has 2930 rows and 3 columns.

```{r}
colnames(rail)
head(rail)
```

The first column (variable) lists states, the second column lists counties in those states, and the third column lists the total number of employees in those counties. Evidently (from the name of the dataset), these employees must be those working in the railroad industry in 2012.

```{r}
head(select(rail, 1))
```

The states in the first column are denoted by their two-letter abbreviations.

```{r}
state_table <- table(select(rail, 1))
prop.table(state_table)
```

Listed out here is the proportion/percentage of how many times each state in this dataset is mentioned. With this information, we can get an idea of where the railroad industry is more prominent.