---
title: "Challenge 4"
author: "Janani Natarajan"
desription: "More data wrangling: pivoting"
date: "05/08/23"
format:
  html:
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
  - challenge_4
  - abc_poll
  - eggs
  - fed_rates
  - hotel_bookings
  - debt
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Read in data

Read in one (or more) of the following datasets, using the correct R package and command.

-   abc_poll.csv ⭐
-   poultry_tidy.xlsx or organiceggpoultry.xls⭐⭐
-   FedFundsRate.csv⭐⭐⭐
-   hotel_bookings.csv⭐⭐⭐⭐
-   debt_in_trillions.xlsx ⭐⭐⭐⭐⭐

```{r}
debt_t <- readxl::read_excel("_data/debt_in_trillions.xlsx", sheet="Sheet1")
table(select(debt_t, `Year and Quarter`))
      
```

### Briefly describe the data

```{r}
view(debt_t)
ncol(debt_t)
```

```{r}
nrow(debt_t)
```

```{r}
nrow(debt_t) * (ncol(debt_t)-1)
```

```{r}
debt_t2 <- debt_t %>%
  pivot_longer(col = -c(`Year and Quarter`), names_to="debt_type", values_to = "debt_value")
```

## Identify variables that need to be mutated

The Year and Quarter variable has to be mutated to make analysis more efficient. Since Year and Quarter denotes a date or time, this is not right and needs to be converted to Date object.

```{r}
library(lubridate)
debt_t3 <- debt_t2 %>%
  mutate(year = str_c("20", str_sub(`Year and Quarter`, 1, 2)),quarter = str_sub(`Year and Quarter`, 5, 5), year_and_quarter = quarter(as_date(str_c(year, quarter), format="%Y%q"), with_year=TRUE)) %>% 
  select(-c(`Year and Quarter`, `year`, `quarter`)) %>%
  relocate(debt_type, debt_value, year_and_quarter) %>%
  mutate(debt_value = str_remove(as.character(debt_value), ".0+$"))

```

```{r}
head(debt_t3)
```
