---
title: "Challenge 8"
author: "Jaswanth Reddy Kommuru"
description: "Joining Data"
date: "05/11/2023"
format:
  html:
    toc: true
    code-copy: true
    code-tools: true
categories:
  - challenge_8
  - Jaswanth Reddy Kommuru
  - snl
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)
library(ggplot2)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Challenge Overview

Today's challenge is to:

1)  read in multiple data sets, and describe the data set using both words and any supporting information (e.g., tables, etc)
2)  tidy data (as needed, including sanity checks)
3)  mutate variables as needed (including sanity checks)
4)  join two or more data sets and analyze some aspect of the joined data

(be sure to only include the category tags for the data you use!)

## Read in data

Read in one (or more) of the following datasets, using the correct R package and command.

  - military marriages ⭐⭐
  - faostat ⭐⭐
  - railroads  ⭐⭐⭐
  - fed_rate ⭐⭐⭐
  - debt ⭐⭐⭐
  - us_hh ⭐⭐⭐⭐
  - snl ⭐⭐⭐⭐⭐

```{r}
adata <- read.csv("~/Documents/601/601_Spring_2023/posts/_data/snl_actors.csv")
cdata <- read.csv("~/Documents/601/601_Spring_2023/posts/_data/snl_casts.csv")
sdata <- read.csv("~/Documents/601/601_Spring_2023/posts/_data/snl_seasons.csv")
```

### Briefly describe the data

```{r}
head(sdata)
```

```{r}
head(cdata)
```

```{r}
head(adata)
```



```{r}
str(sdata)
```

```{r}
str(cdata)
```

```{r}
str(adata)
```

```{r}
dim(adata)
```

```{r}
dim(sdata)
```

```{r}
dim(cdata)
```





```{r}
colnames(snldata)
```

```{r}
dim(snldata)
```

```{r}
summary(adata)
```

```{r}
summary(cdata)
```

```{r}
summary(sdata)
```


Separate and well-organized data are contained in the data frames. The list of people who have appeared on the show, including actors, guests, musical guests, and staff members, is vast in the Actors Data Frame. Each item is specific to each actor and includes information about their gender and type of role. The Casts Data Frame, on the other hand, concentrates on actors who were a part of the cast during a specific season. Each item corresponds to a particular actor and contains details like the number of episodes they appeared in that season, the dates of their first and last appearances, whether they were a Weekend Update anchor, and their featured status. The Seasons Data Frame, contains details about each distinct season. Each entry is for a specific season and contains information about that season, including the year it debuted, the premiere and finale dates, and the total number of episodes.







## Tidy Data (as needed)

Is your data already tidy, or is there work to be done? Be sure to anticipate your end result to provide a sanity check, and document your work here.

```{r}

sdata <- sdata %>%
  mutate(across(c(first_epid, last_epid), ymd))

cdata <- cdata %>%
  mutate(across(c(first_epid, last_epid), ymd))

```

Are there any variables that require mutation to be usable in your analysis stream? For example, do you need to calculate new values in order to graph them? Can string values be represented numerically? Do you need to turn any variables into factors and reorder for ease of graphics and visualization?

Document your work here.

```{r}

```

## Join Data

Be sure to include a sanity check, and double-check that case count is correct!

```{r}
adata <- adata %>%
  filter(type == "cast") %>%
  select(-url)

member_count <- n_distinct(cdata$aid)
member_count

```
```{r}
joined_data <- full_join(adata, cdata, by = "aid") %>%
        select(c(aid, gender, sid, featured, update_anchor))

head(joined_data)
```





