---
title: "Challenge 5-AB_NYC_2019"
author: "Jinxia Niu"
description: "Introduction to Visualization"
date: "3/28/2023"
format:
  html:
    toc: true
    code-copy: true
    code-tools: true
categories:
  - challenge_5
  - Jinxia Niu
  - AB_NYC_2019
  
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)
library(ggplot2)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Challenge Overview

Today's challenge is to:

1)  read in a data set, and describe the data set using both words and any supporting information (e.g., tables, etc)
2)  tidy data (as needed, including sanity checks)
3)  mutate variables as needed (including sanity checks)
4)  create at least two univariate visualizations
   - try to make them "publication" ready
   - Explain why you choose the specific graph type
5)  Create at least one bivariate visualization
   - try to make them "publication" ready
   - Explain why you choose the specific graph type

[R Graph Gallery](https://r-graph-gallery.com/) is a good starting point for thinking about what information is conveyed in standard graph types, and includes example R code.

(be sure to only include the category tags for the data you use!)

## Read in data

Read in one (or more) of the following datasets, using the correct R package and command.

-   cereal.csv ⭐
-   Total_cost_for_top_15_pathogens_2018.xlsx ⭐
-   Australian Marriage ⭐⭐ 
-   AB_NYC_2019.csv ⭐⭐⭐
-   StateCounty2012.xls ⭐⭐⭐
-   Public School Characteristics ⭐⭐⭐⭐ 
-   USA Households ⭐⭐⭐⭐⭐

```{r}
library(tidyverse)
 AB_NYC_2019 <- read.csv("_data/AB_NYC_2019.csv")
 dim(AB_NYC_2019)
 colnames(AB_NYC_2019)
head(AB_NYC_2019,10)


```

```{r}
AB_NYC_2019 %>% 
    select( "room_type",  "neighbourhood", "price", everything())%>%
    group_by(room_type) %>%
    summarise(avg_price = mean(price, na.rm = TRUE))

```


### Briefly describe the data
It's a data set of the Airbnb data in New York City in 2019, it includes 16 variables.48895 observations. The key variables includes the neighborhoods the Airbnb rooms located in, the location, the prices and the reviews they got. There are total three kind of rooms:Emtir home/apt, Private room and Shared room, and the avarage price for each different kind of room is $211, $89 and $70.

The key questions I am interested are: what's the most expensive neighborhoods? What's the relationship between the price and the location of each different kind of rooms? 

## Tidy Data (as needed)

```{r}
AB_NYC_2019 %>% 
  select(-c("id","host_id","host_name","last_review","calculated_host_listings_count" )) %>%
  print(AB_NYC_2019)

  # what's the codes to combine two variables into one in R? like combine "latitude" and "longitude"? 
```
## Univariate Visualizations

```{r}
  ggplot(AB_NYC_2019,aes(price))+
  geom_histogram()


```
```{r}
AB_NYC_2019 %>%
  ggplot(aes(neighbourhood_group,price))+
  geom_boxplot()
```
```{r}
AB_NYC_2019 %>%
  ggplot(aes(number_of_reviews,neighbourhood_group,))+
  geom_point()+
  facet_wrap(vars(neighbourhood_group))
```

```{r}
  AB_NYC_2019 %>%
  ggplot(aes(price, number_of_reviews))+
  geom_line()
 

```


Is your data already tidy, or is there work to be done? Be sure to anticipate your end result to provide a sanity check, and document your work here.

Are there any variables that require mutation to be usable in your analysis stream? For example, do you need to calculate new values in order to graph them? Can string values be represented numerically? Do you need to turn any variables into factors and reorder for ease of graphics and visualization?

Document your work here.

```{r}




```

## Univariate Visualizations


## Bivariate Visualization(s)
Any additional comments?
