---
title: "Challenge 2 by Jinxia Niu"
author: "Jinxia Niu"
desription: "Data wrangling: using group() and summarise()"
date: "03/14/2023"
format:
  html:
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
  - challenge_2
  - Jinxia Niu
  - birds.csv

---

## Challenge Overview

Today's challenge is to

1)  read in a data set, and describe the data using both words and any supporting information (e.g., tables, etc)
2)  provide summary statistics for different interesting groups within the data, and interpret those statistics

## Read in the Data

Read in one (or more) of the following data sets, available in the `posts/_data` folder, using the correct R package and command.

-   railroad\*.csv or StateCounty2012.xls ⭐
-   FAOstat\*.csv or birds.csv ⭐⭐⭐
-   hotel_bookings.csv ⭐⭐⭐⭐

```{r}
install.packages("tidyverse")
library(tidyverse)
birds <- read_csv("_data/birds.csv") %>%
  select(-c(contains("Code"),Element,Domain,Unit))%>% 
  filter(Flag!="A")
View(birds)
```

## Describe the data

Birds.csv that includes estimates of the stock of five different types of poultry (Chickens, Ducks, Geese and guinea fowls, Turkeys, and Pigeons/Others) for 248 areas for 58 years between 1961-2018. Estimated stocks are given in 1000 head.


## Provide Grouped Summary Statistics
Conduct some exploratory data analysis, using dplyr commands such as `group_by()`, `select()`, `filter()`, and `summarise()`. Find the central tendency (mean, median, mode) and dispersion (standard deviation, mix/max/quantile) for different subgroups within the data set.

```{r}
birds %>%
  group_by(Item)%>%
  summarise(avg_stocks = mean(Value, na.rm = TRUE))
```
### Explain and Interpret
Try to understand the sizes of stocks of each of the five types of poultry in the dataset and try to understand which arears/countries has how many of poultry during this time. On average, we can see that countries have far more chickens as livestock (=58.4million head) than other livestock birds.

```{r}
birds %>%
  group_by(Area)%>%
  summarize(avg_stocks = mean(Value, na.rm = TRUE))

```




