---
title: "Challenge 1 by Jinxia"
author: "Jinxia Niu"
desription: "Reading in data and creating a post"
date: "02/22/2023"
format:
  html:
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
  - challenge_1
  - Jinxia Niu
  - datasets:birds.csv and wild_birds_data.xlsx
---

```{r}
#Reading in data

library(tidyverse)
df <- read.csv("_data/birds.csv")
View(df)
nrow(df)
install.packages("readxl")
library("readxl")
df1 <- read_excel("_data/wild_bird_data.xlsx")

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)

```
# Describe the data

Birds.csv is a data sets of a total of 14 columns,30977 rows.It includes 4-6 kinds of birds gatherer from 9 areas, with the years of each kind of births discovered.

Wildbirds.xlsl is a dataset includes 2 columns and 147 rows.The birds are sorted by Wet body weight and population size,






## Read in the Data

Read in one (or more) of the following data sets, using the correct R package and command.

-   railroad_2012_clean_county.csv ⭐
-   birds.csv ⭐⭐
-   FAOstat\*.csv ⭐⭐
-   wild_bird_data.xlsx ⭐⭐⭐
-   StateCounty2012.xls ⭐⭐⭐⭐

Find the `_data` folder, located inside the `posts` folder. Then you can read in the data, using either one of the `readr` standard tidy read commands, or a specialized package such as `readxl`.

```{r}

```

Add any comments or documentation as needed. More challenging data sets may require additional code chunks and documentation.

## Describe the data

Using a combination of words and results of R commands, can you provide a high level description of the data? Describe as efficiently as possible where/how the data was (likely) gathered, indicate the cases and variables (both the interpretation and any details you deem useful to the reader to fully understand your chosen data).

```{r}
#| label: summary

```
