---
title: "Challenge 1 - Railroad Employees"
author: "Joseph Vincent"
desription: "Reading in data, describing, and creating first post"
date: "02/15/23"
format:
  html:
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
  - challenge_1
  - railroads
  - Joseph Vincent
---

```{r}
#| label: setup
#| warning: false

library(tidyverse)
library(summarytools)
library(ggplot2)

knitr::opts_chunk$set(echo = TRUE)
```

## Reading in the data

-   railroad_2012_clean_county.csv ⭐

```{r}
# loading in dataset and assigning to variable 'railroad'
# using head to preview the dataset

railroad <- read_csv("_data/railroad_2012_clean_county.csv")
head(railroad)

```

## Describing the dataset

This data set consists of 3 columns: State, County and Total Employees. It appears that the data set is showing the number of railroad employees by county in the United States. 

The least number of railroad employees in a given county is 1, and the greatest number of railroad employees in a given county is over 8,000. 

There is a mean number of employees per county of about 87, but with a large standard deviation (283).

I learned that there are 31 counties with the name "Washington" and 26 counties with the name "Jefferson".

## Data summary

```{r}
#| label: summary

# finding the dimensions of 'railroad'
dim(railroad)

# finding the column names of 'railroad'
colnames(railroad)

#using summary tools
dfSummary(railroad)

# practice selecting employees column and calc min/max manually
employees <- select(railroad, total_employees)
head(employees)
min(employees)
max(employees)
```