---
title: "Challenge 4 Instructions"
author: "Kevin Martell Luya"
description: "More data wrangling: pivoting"
date: "04/23/2023"
format:
  html:
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
  - challenge_4
  - abc_poll
  - eggs
  - fed_rates
  - hotel_bookings
  - debt
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)
library(readxl)
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Challenge Overview

Today's challenge is to:

1)  read in a data set, and describe the data set using both words and any supporting information (e.g., tables, etc)
2)  tidy data (as needed, including sanity checks)
3)  identify variables that need to be mutated
4)  mutate variables and sanity check all mutations

## Read in data

Read in one (or more) of the following datasets, using the correct R package and command.

-   abc_poll.csv ⭐
-   poultry_tidy.xlsx or organiceggpoultry.xls⭐⭐
-   FedFundsRate.csv⭐⭐⭐
-   hotel_bookings.csv⭐⭐⭐⭐
-   debt_in_trillions.xlsx ⭐⭐⭐⭐⭐

```{r}
debt <- read_xlsx("./_data/debt_in_trillions.xlsx")
debt
```

### Briefly describe the data

The data shows year and quaters and different types of loans. Regarding the first column, we can modify the year and quarter format.
## Tidy Data (as needed)

Is your data already tidy, or is there work to be done? Be sure to anticipate your end result to provide a sanity check, and document your work here.

As data it is not tidy, there is work to be done.

```{r}
summary(debt)
```
As we can see, Year and Quarter column summary does not tell us much in terms of statistical data.
Let's fix this.

## Identify variables that need to be mutated

Are there any variables that require mutation to be usable in your analysis stream? For example, are all time variables correctly coded as dates? Are all string variables reduced and cleaned to sensible categories? Do you need to turn any variables into factors and reorder for ease of graphics and visualization?

Document your work here.

```{r}
debt_quarte_format <- debt %>%
  mutate(date = parse_date_time(`Year and Quarter`,orders = "yq"))%>%
  select(!contains("Year and Quarter"))

debt_quarte_format
```

Any additional comments?
Now let see the date summary
```{r}
summary(debt_quarte_format$date)
```

