---
title:  "Challenge 1 Instructions"
author:  "Khadijat Adeleye"
description: "Reading in data and creating a post"
date: "03/01/2023"
format:
  html:
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
  - challenge_1
  - KhadijatAdeleye
  - Birds
  - ggplot2
---

##Reading in data
```{r}
df <- read.csv("_data\\birds.csv")

view(df)
```

#Description of Datasets
The second dataset has 14 columns and 30977 observations. From colnames, we get to know that the dataset gives us the values of the dietary energy intake for different countries across different years from 1961-2017. Data types of the columns, value could actually be converted into double type. There were around 11000 missing values found and removed from the data. Many countries were included in this dataset and there are 6 types of birds but only one domain of animals present.
```{r}
dim(df) 
```
```{r}
list(colnames(df))
```
```{r}
str(df)
```

```{r}
df <- transform(df,val= as.numeric(Value))
sum(is.na(df))
```
```{r}
df<-na.omit(df)
dim(df)
```
```{r}
area<-select(df,Area)
table(area)
```
#list of items in the dataset
```{r}
item<-select(df,Item)
table(item)
```

```{r}
domain<-select(df,Domain)
table(domain)
```
```{r}
plot(val~Year,df)
```


```{r}
df_Afghanistan <-df%>% filter(`Area`=='Afghanistan')
plot(val~Year,df_Afghanistan)
```