---
title: "Challenge 4 Instructions"
author: "Khadijat Adeleye"
desription: "More data wrangling: pivoting"
date: "03/29/2023"
format:
  html:
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
  - challenge_4
  - abc_poll
  - eggs
  - fed_rates
  - hotel_bookings
  - debt
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Read in data

Read in one (or more) of the following datasets, using the correct R package and command.

-   abc_poll.csv ⭐
-   poultry_tidy.xlsx or organiceggpoultry.xls⭐⭐
-   FedFundsRate.csv⭐⭐⭐
-   hotel_bookings.csv⭐⭐⭐⭐
-   debt_in_trillions.xlsx ⭐⭐⭐⭐⭐


```{r}
  df <- read.csv("_data/FedFundsRate.csv")
view(df)
```
####DATASET DESCRIPTION
The federal funds dataset shows  the trend of united states economy related to federal funds. The economy was measured by the unemployment rate, inflation rate and GDP. The data set consist of 10 column and 904rows. There are 3196 null values in entire data frame. There are seven columns with more than one null Values.
```{r}
# num of rows of dataset
nrow(df)
```

```{r}
#num of cols of dataset
ncol(df)
```
```{r}
#name of all the columns
colnames(df)
```
```{r}
   tail(df)
```

```{r}
###count for number of null values
sum(is.na(df))
```
#####TIDY DATASET
when tidying up this data, we first want to check if there are any missing entries in the data set.

```{r}
#count number of missing entries 
num_missing_cols<-colSums(is.na(df))
print(num_missing_cols)
```

# The chosen format  was pivoted longer and the  unspecified data values(N/A values) in the dataset. The columns  and rows were remove from dataframe
```{r}
#remove missing values
#pivot longer of the federal fund rate columns, removed NA values
clean_FedFundsRate <-pivot_longer(df, col = c("Federal.Funds.Target.Rate" , "Federal.Funds.Lower.Target" , "Effective.Federal.Funds.Rate"),
                 names_to="Federal Fund Type",
                 values_to = "Federal Fund Rate",
                 values_drop_na = TRUE)
```
```{r}
#pivot longer of the federal fund rate columns, removed NA values. 

clean_FedFundsRate
clean_FedFundsRate <-p<-pivot_longer(df, col = c("Real.GDP..Percent.Change.", "Unemployment.Rate", "Inflation.Rate"),
                 names_to="GDP Condition",
                 values_to = "GDP Rate",
                 values_drop_na = TRUE)

clean_FedFundsRate
```

####Mutating Data
Also merge date with month, day and year into one easily readable “Date” column. 
```{r}
###margin month and date  variable names to make easier to reference in code
clean_FedFundsRate <- mutate(clean_FedFundsRate, Date = make_date(Year, Month, Day))
```


```{r}
clean_FedFundsRate<-clean_FedFundsRate[complete.cases(clean_FedFundsRate$"Inflation.Rate"),]
select(clean_FedFundsRate,Date,"Inflation.Rate","Unemployment.Rate","Federal Fund Rate","Real.GDP..Percent.Change.","Federal Fund Type")
```



