---
title: "Final Project : Project & Data Description"
author: "KhadijatAdeleye"
description: "Project & Data Description"
date: "05/21/2023"
format:
  html:
    df-print: paged
    toc : true
    code-copy: true
    code-tools: true
    css: styles.css
categories:
  - final_Project_assignment_1
  - final_project_data_description
editor_options: 
  chunk_output_type: inline
---

```{r}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
# Install packages (Run Only ONCE)
# install.packages("haven")
# install.packages("devtools")
# install.packages("scales")
# install.packages('rlang')
###install.packages(c("ggplot2", "ggpubr", "tidyverse", "broom", "AICcmodavg"))
pacman :: p_load(pacman,stats, dplyr, knitr, ggplot2, plotly, psych, gridExtra,
                 waffle, emojifont,tidyr, tidytext, wordcloud, GGally, viridis,
                tidyverse, rnaturalearth, rnaturalearthdata)
```


```{r}
# Load packages
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE)
library(tidyverse)
library(haven)
library(ggplot2)
library(lubridate)
library(dplyr)
```


# INTRODUCTION
This cross-sectional study utilized a survey to collect data. The analysis presented below focuses exclusively on individuals taking antiretroviral medication, resulting in a sample size of 485 women living with HIV.
This paper will explores the  relationship between social capital and adherence among women living with HIV in a community with High HIV prevalence rate.The paper specific objective is the factors influencing adherence to ante-retro viral medication. The paper will begin with a brief introduction of the dataset and a resulting tidy data part. In the third section, the paper will explore the clean data through visualization in order to answer a set of questions.
# Research Questions : what is the level of ante-retroviral medication adherence among women living with HIV? Is there an association between social capital  and adherence to ante-retroviral medication ? 
It is important to note that the relationship between social capital and medication adherence can be complex and influenced by various factors, including cultural context, socioeconomic status, and individual characteristics. However, by leveraging social capital and strengthening supportive networks, healthcare providers and communities can potentially improve medication adherence and overall health outcomes.
# The initial step in the study involved examining descriptive statistics for all the variables included. Descriptive statistics were calculated for all study variables to summarize their central tendency (e.g., mean) and variability (e.g., standard deviation). Analysis of variance (ANOVA) was utilized for continuous variables. These tests were conducted to examine potential differences in participant characteristics and explore bivariate relationships between study variables. The significance level (alpha) was set at 0.05 to determine statistical significance. In the case of multiple comparisons involving continuous variables, a Bonferroni correction was applied to adjust for the increased risk of false positives (Type I error).
 This  analysis aimed to examine the relationship between social capital and antiretroviral medication adherence while accounting for the effects of other relevant factors.

 



```{r}
compliances_df <- read_dta('./Adherence at postpartum (1).dta')
compliances_df
```

```{r}
labelled::generate_dictionary(compliances_df) %>% as_tibble 
```



```{r}
####checking the string and number of rows
str(compliances_df)
sum(is.na(compliance_df))
is.numeric(compliances_df$stopandtalkwithpeopleinneighbour)

```
## DATASET DESCRIPTION.
This is a data retrieved from women living with HIV on their social wellbeing, social capital and adherence to anteretroviral medications.The data set has 165 column and 485 rows. To determine the relationship between adherence and controlling variable. So I reorganize the dataset. I first use the “select’’ function to create a dataset covering all”adherence” questions. Then, I create another dataset containing all controlling variable (age, marital status, level of education, smoking, year of disclosure).

```{r}
colnames(compliances_df,15)
###head(compliances_df, 15)

```
 Inspecting the dataset
```{r}
###checking duplicate
compliances_df %>% skimr::skim()
sum(duplicated(compliances_df))

# Inspect data
check <- function(data) {
  l <- list()
  columns <- names(data)
  for (col in columns) {
    instances <- sum(!is.na(data[[col]]))
    dtypes <- class(data[[col]])
    unique <- length(unique(data[[col]]))
    sum_null <- sum(is.na(data[[col]]))
    duplicates <- sum(duplicated(data))
    l[[length(l) + 1]] <- c(col, dtypes, instances, unique, sum_null, duplicates)
  }
  data_check <- as.data.frame(do.call(rbind, l))
  names(data_check) <- c("column", "dtype", "instances", "unique", "sum_null", "duplicates")
  return(data_check)
}

check(df)
```

```{r}
## unique value in Adherence recoded variable
unique(compliances_df$Ahdrence_recode)
```
# Checking the Demographic variale
```{r}
library(psych)
describe(compliances_df)

```

# TIDY THE DATA SET
It was crucial to maintain the Data quality: hence,the dataset missing values will added to the dataset as the sample size was small that can impact the accuracy and reliability of the results.Variable selection was based on past literatures on the topic of medication adherence.



```{r}
#compliances_df %>% count(HaveyoudisclosedyourHIVstatustoy)

compliances_df %>% names() %>% str_subset("recode")
compliances_df_clean %>% names()




compliances_df_clean <- compliances_df %>%
  mutate(age_recode = case_when(
    age_recode == 1 ~ "Young Adult",
    age_recode == 2 ~ "Middleage Adult", 
    age_recode == 3 ~ "Adult", 
    age_recode == 4 ~ "Older Adult"
 )) %>% 
   mutate(maritalstatus_recode = case_when(
   MaritalStatus == 1 ~ "Single",
    MaritalStatus == 2 ~ "Married", 
    MaritalStatus == 3 ~ "Not in Union", 
    MaritalStatus == 4 ~ "Not in Union"
  )) %>% 
  mutate(mat_education = case_when(
    Education <= 2 ~ "Below Grade 8",
    Education == 3 ~ "Grade 8-12", 
    Education > 3 ~ "Above Grade 12"
  )) %>% 
   mutate(employ =ifelse(Salarypayingjob == 0, "Employ", "Unemploy"
   )) %>%
  replace_na(list(HaveyoudisclosedyourHIVstatustoy = 2)) %>% 
  mutate(mat_disclosure = case_when(
  HaveyoudisclosedyourHIVstatustoy  == 1 ~ "Disclosure", 
  HaveyoudisclosedyourHIVstatustoy  == 2 ~ "No disclosure"
   )) %>% 
  mutate(Larrangement = case_when(
     LivingArrangements <= 3 ~ "Isolated", 
     LivingArrangements > 3 ~ "Not Isolated"
    )) %>%
  mutate(
    Haveyouconsumedanalcoholwithinth = ifelse(
      Haveyouconsumedanalcoholwithinth  == 0, "No", "Yes"))
  
# head(compliances_df_age)
  
 ### head(compliances_df_clean)
  

# compliances_df_clean %>% count(age_recode, sort = TRUE)
# compliances_df_clean %>% count(maritalstatus_recode, sort = TRUE)
# compliances_df_clean %>% count(mat_education, sort = TRUE)
# compliances_df_clean %>% count (adherence1, sort = TRUE)

```

# SOCIODEMOGRAPHICAL VARIABLE

AGE OF PARTICIPANTS
```{r}
 
 ###Age was collected as a continuous variable


# Calculate mean age
mean_age <- mean(compliances_df$Age)
print(mean_age)
# Median
cat("Median:", median(compliances_df$Age))
#### Mode
mode <- function(x){
  ta <- table(x)
  tam <- max(ta)
  if(all(ta==tam))
    mod <- NA
  else
    if(is.numeric(x))
      mod <- as.numeric(names(ta)[ta==tam])
  else
    mod <- names(ta)[ta==tam]
  return(mod)
}

cat("Mode:", mode(compliances_df$Age))
    # Measure of Variability

# Std. Dev
cat("Standard Deviation:", sd(compliances_df$Age))
# Variance
cat("Variance:", var(compliances_df$Age))

# psych::describe(compliances_df$Age)
```

 The overall sample size is 485 women living with HIV. Of these the mean age is 32yrs(s.d=5.742912), and majority were between  28 and 34years.


Visualization of Participant Age distribution

```{r}

# Tabulate Age
tab_age <- table(compliances_df$Age)

# Tidy the tabulated results
tidy_tab_age <- as.data.frame(tab_age)

# Visualize the tabulated results

plot(density(compliances_df$Age),
     col="#F5B0CB",
     main="Density Plot for Age",
     xlab="Age",
     ylab="Density")
polygon(density(compliances_df$Age),
        col="#F5B0CB")
```

 HIV status disclosure
Participants were asked to report their HIV status disclosure to different individuals or groups. The categories mentioned in the study include disclosure to a partner, disclosure to a family member, and disclosure to a religious leader. Based on these responses, a new variable was created to dichotomize HIV disclosure. Participants who reported disclosing their HIV status were assigned a value of 1, indicating HIV disclosure. In contrast, those who did not disclose their HIV status were assigned a value of 0, representing no HIV status disclosure. l. 

```{r}


######percentage
percentage_mat_disclosure <- tab / sum(tab_mat_disclosure) * 100
print(percentage_mat_disclosure )



# Create a data frame with the data points
disclosure <- data.frame(Category = c("Disclosure", "No disclosure"),
                 Count = c(400, 82))

# Plot the scatter plot
ggplot(disclosure, aes(x = Category, y = Count)) +
  geom_point() +
  xlab("Category") +
  ylab("Count") +
  ggtitle("Count of Disclosure vs. No disclosure") +
  theme_minimal()



```

 RESULT: Majority 400(83%) of the participant  had disclosed their HIV status to either partner, family or religious leader.


Participants living arrangement: Participants were asked to report their Living arrangement. The categories mentioned in the study include living,Living with both parents, Living alone with children, Living with one biological parent, Single not living with children,   They are living with relatives, Married and living with a husband and children, Married Living with children but not living with husband, Single living with partner and children. Based on these responses, a new variable was created to dichotomize Living arrangement to isolated living  and not living isolated. 
```{r}
# Create a bar chart
bar_chart <- ggplot(LivingArrangement, aes(x = category, y = count, fill = category)) +
  geom_bar(stat = "identity") +
  labs(title = "Living Arrangement Distribution", x = "Living Arrangement", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for better readability

# Display the bar chart
print(bar_chart)



```

 RESULT:  About 274 women Lives with either family or children, while 211 lives alone.



EMPLOYMENT STATUS
```{r}
# Create a data frame with the counts and categories
employment_df <- data.frame(
  Category = c("Employed", "Unemployed"),
  Count = c(157, 328)
)

# Create the bar chart
ggplot(employment_df, aes(x = Category, y = Count, fill = Category)) +
  geom_bar(stat = "identity") +
  labs(x = "Employment Status", y = "Count", title = "Count of Employment Status") +
  theme_minimal()

```
Most participant are unemployed which can have negative implication on the medication adherence.

Analysis for Partners Status 
```{r}


cat("The dataset grasps", length(unique(compliances_df $Partner_status))," distinct partners status")

```



 Year of HIV Diagnosis and Adherence
```{r}
compliances_df %>%
  ggplot(aes("Aherence", Year_diagnosed))+
  geom_point()

```
 Findings: Majority of Participant diagnosed three years ago adhere to anteretroviral medication.



 DEPENDENT VARIABLE
 Antiretroviral Adherence: In order to assess adherence to antiretroviral medication, a 7-item questionnaire was used. This questionnaire was pretested and demonstrated validity in measuring adherence. Each item in the questionnaire asked participants about their medication-related behaviors and causal indicators. For example, participants were asked if there were any days in the past two weeks when they did not take their antiretroviral (ARV) medication. All seven items utilized a yes/no response format.
The responses to the seven items were summed to create a continuous measure of self-reported adherence. The resulting estimate ranged from 0 to 7, with higher scores indicating greater adherence to the prescribed antiretroviral medication regimen.
The adherence scores were then categorized using cut points. Based on previously defined categories, participants with a cumulative score of less than 6 were classified as having low adherence. Participants with a cumulative score of seven were classified as having higher adherence, meaning they regularly reported taking their medication as prescribed.


```{r}
compliances_df3 <- compliances_df %>%
  select(Sincedeliverydoyousometimesforge, Past2weekswerethereanydaysyoudid,
  StoppedtakingARVmedicationbecaus, Whentravellingorleavinghomedoyou,
  DidyouuseARVyesterday, Whenyoufeelhealthydoyousometimes,
  Doyouoftenhavedifficultyremember
  ) %>% 
  mutate(across(everything(), as.numeric)) %>% 
  mutate(across(everything(), .fns = ~{(max(.x)) - .x})) %>% 
  # mutate(DidyouuseARVyesterday = ifelse(DidyouuseARVyesterday == 0, 1, 0)) %>%
  rowwise() %>% 
  mutate(adherence = sum(c_across(cols = everything()))) %>% 
  ungroup()

Adherencel <- compliances_df3 %>% 
  summarise(
    mean_sc = mean(adherence),
    mode_sc = mode(adherence),
    median_sc = median(adherence),
    sd = sd(adherence)
  )

  Adherencel

 psych::describe(compliances_df3$adherence)


compliances_df3 %>% 
  select(-adherence) %>% 
  pivot_longer(
    cols = everything(), 
    names_to = "variables", 
    values_to = "values") %>% 
  group_by(variables, values) %>% 
  count() %>% 
  ungroup %>% 
  mutate(variables = case_when(
    variables == "Sincedeliverydoyousometimesforge" ~ "Near Missed Dose",
    variables == "Past2weekswerethereanydaysyoudid" ~ "Missed Dose", 
    variables == "StoppedtakingARVmedicationbecaus" ~ "Stopped ARV", 
    variables == "Whentravellingorleavinghomedoyou" ~ "Travelling",
    variables == "DidyouuseARVyesterday" ~ "ARV Daily Use",
    variables == "Whenyoufeelhealthydoyousometimes" ~ " Feel Healthy",
    variables == "Doyouoftenhavedifficultyremember" ~ "Rememberance"
  )) %>% 
  mutate(
    variables = as.factor(variables), 
    values = reorder_within(values, n, variables)) %>%
  ggplot(aes(values, n, fill = variables)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~variables, scales = "free_y") +
  coord_flip() +
  scale_x_reordered() +
  scale_y_continuous(expand = c(0,0)) +
  theme_minimal() +
  labs(
    y = "Count of Response",
    x = "Scores",
    title = "ARV Adherence",
    subtitle = "Participants Adherence Level"
  )

```
 The missing data were included to the upper border of the number as the lower border has same value. Findings implies most respondents do not take ARV on daily use
 
 Adherence and Agegroup
```{r}
adherence_age_df <- compliances_df_clean %>% 
  select(age_recode) %>% 
  mutate(id = row_number()) %>% 
  full_join(
    compliances_df3 %>% select(adherence) %>% mutate(id = row_number()),
    by = c("id")
  ) %>% 
  select(-id)

adherence_age_plot_df <- adherence_age_df %>% 
  group_by(age_recode, adherence) %>% 
  count(sort = TRUE) %>% 
  group_by(age_recode) %>% 
  mutate(prop = n/sum(n)) %>% 
  ungroup
  
  adherence_age_plot_df %>% count(adherence)

 adherence_age_plot_df %>% filter(is.na(adherence))



adherence_age_plot_df %>% 
  mutate(
    age_recode = as.factor(age_recode), 
    adherence = as.factor(adherence)#,
    #adherence = reorder_within(adherence, prop, age_recode)
    ) %>%
  ggplot(aes(adherence, prop, fill = age_recode)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~age_recode, scales = "free_y") +
  # coord_flip() +
  #scale_x_reordered() +
  scale_y_continuous(expand = c(0,0), labels = scales::percent_format()) +
  theme_minimal() +
  labs(
    y = "Adherence Rates",
    x = "Scores",
    title = "Compare Adherence Rates between Age Groups"
    
  )

```


```{r}


model <- aov(adherence ~ age_recode, data = adherence_age_df)

summary(model)




```
INTEPRETATION: The result of the one-way ANOVA analysis shows a significant effect of age group (age_recode) on the variable being analyzed. The F-value is 4.203, and the associated p-value is 0.00596. This indicates that there are statistically significant differences in the means of the variable across the different age groups.

The degrees of freedom for the age_recode factor is 3, representing the number of age groups minus 1. The sum of squares for the age_recode factor is 16.0, and the mean sum of squares is 5.321. The sum of squares for the residuals (unexplained variation) is 608.9, and the mean sum of squares is 1.266.



** INDEPENDENT VARIABLES
 Recoding Data to measure 
SOCIAL CAPITAL: Social capital refers to the resources, networks, and relationships that exist within a social group or community. It represents the value derived from social interactions, trust, cooperation, and reciprocity among individuals or groups.In  this context it was identified as resources available to improve medication.

In the study examining the correlates of antiretroviral adherence, perceived social capital was assessed as a self-assessment of an individual's understanding and knowledge about their social capital. To measure this, participants were asked to indicate their level of agreement on a 5-point Likert-type scale with the statement, "When I am feeling happy, I am careful not to show it." The response options ranged from 0, indicating "Strongly Disagree," to 4, indicating "Strongly Agree." Participants' responses were analyzed and averaged to construct a continuous measure of social capital. The final measure ranged from 18 to 40, signifying the total score derived by summing up the individual item replies.

```{r}
compliances_df2 <- compliances_df %>%
  select(haveaclosefriend, stopandtalkwithpeopleinneighbour,
  socialnetworkgroup, satifiedwithfamilylife,
  satisfiedwithsociallife, relyonfriendandfamilyifIhaveaser,
  givespecialhelptoatleastonesickd, borrowthingsandexchangefavourswi
  ) %>% 
  mutate(across(everything(), as.numeric)) %>% 
  mutate(satifiedwithfamilylife = ifelse(
    satifiedwithfamilylife == 22, 4, satifiedwithfamilylife)) %>% 
  mutate(across(everything(), .fns = ~{(1+max(.x)) - .x})) %>% 
  rowwise() %>% 
  mutate(soc_cap = sum(c_across(cols = everything()))) %>% 
  ungroup()

social_capital <- compliances_df2 %>% 
  summarise(
    mean_sc = mean(soc_cap),
    mode_sc = mode(soc_cap),
    median_sc = median(soc_cap),
    sd = sd(soc_cap)
  )

compliances_df2 %>% 
  select(-soc_cap) %>% 
  pivot_longer(
    cols = everything(), 
    names_to = "variables", 
    values_to = "values") %>% 
  group_by(variables, values) %>% 
  count() %>% 
  ungroup %>% 
  mutate(variables = case_when(
    variables == "borrowthingsandexchangefavourswi" ~ "Loan",
    variables == "givespecialhelptoatleastonesickd" ~ "Special Help", 
    variables == "haveaclosefriend" ~ "Closefriend", 
    variables == "relyonfriendandfamilyifIhaveaser" ~ "Trust",
    variables == "satifiedwithfamilylife" ~ "Family Life",
    variables == "satisfiedwithsociallife" ~ "social Life",
    variables == "socialnetworkgroup" ~ "Networking",
    variables == "stopandtalkwithpeopleinneighbour" ~ "Neighbourhood relationship"
  )) %>% 
  mutate(
    variables = as.factor(variables), 
    values = reorder_within(values, n, variables)) %>%
  ggplot(aes(values, n, fill = variables)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~variables, scales = "free_y") +
  coord_flip() +
  scale_x_reordered() +
  scale_y_continuous(expand = c(0,0)) +
  theme_minimal() +
  labs(
    y = "Count of Response",
    x = "Scores",
    title = "Social Capital",
    subtitle = "Participants SocialCapital"
  )

 
```


Findings reveals that majority of participant had score a total of 32 out of 40 for social capital. This implies that majority had high social capital. Most participants reported lower rates of social capital, with a mean (30.63093) s.d=+/-3.1  (scored from 18 to 40). with a median score of 30.
 
 
# INFERENTIAL STATISTICS
```{r}
model <- aov(adherence ~ soc_cap, data = adherence_age_df)

summary(model)

```
The result of the one-way ANOVA analysis shows a significant effect of social capital (soc_cap) on the variable being analyzed. The F-value is 6.385, and the associated p-value is 0.0118. This indicates that there are statistically significant differences in the means of the variable across different levels of social capital.

The degrees of freedom for the soc_cap factor is 1, representing the number of levels of social capital minus 1. The sum of squares for the soc_cap factor is 8.2, and the mean sum of squares is 8.153. The sum of squares for the residuals (unexplained variation) is 616.7, and the mean sum of squares is 1.277.



The relationship between Age and Marital status of the participants
```{r}

###
compliances_age_marital_df <- compliances_df_clean %>% 
  group_by(age_recode, maritalstatus_recode) %>% 
  count(sort = TRUE) %>% 
  ungroup()

levels <- compliances_age_marital_df %>%
  arrange(maritalstatus_recode == 'Single', desc(n)) %>%
  pull(age_recode) %>% unique

# Visualize age and marital status
compliances_age_marital_df %>% 
  mutate(age_recode = factor(age_recode, levels)) %>%
  ggplot(aes(
    x = age_recode, 
    y = n, 
    fill = maritalstatus_recode)) +
  # geom_bar(stat = "identity", position = position_dodge()) +
  geom_col(position = position_stack(reverse = FALSE)) +
  # scale_fill_brewer(palette="Paired") +
  theme_minimal() +
  labs(
    x = "Age", y = "Population", fill = "Marrital \nStatus",
    title = "Relationship Between Age and Marital Status"
  )
```
 FINDINGS: Participants self-reported information was used to determine participants' relationship status, which was categorized into married, single, cohabiting, and formerly married.  Over 300 participant were single while around 100 are Married, and about 10 are previously married. While the majority of  older adult participants were single, almost 2/3rd are married. A significant population of young Adults were single. 
 

 The Relationship between Participants Marital status and Educational Status
```{r}
# tab_maritalstatus <- table(compliances_df$MaritalStatus)
# print(tab_maritalstatus)

# Relationship between age and educational level
compliances_age_education_df <- compliances_df_clean %>% 
  group_by(age_recode, mat_education) %>% 
  count(sort = TRUE) %>% 
  ungroup()

# Visualize age and marital status
compliances_age_education_df %>% 
  mutate(age_recode = factor(
    age_recode, 
    levels = compliances_age_education_df$age_recode %>% unique())) %>%
  ggplot(aes(
    x = age_recode, 
    y = n, 
    fill = mat_education)) +
  # geom_bar(stat = "identity", position = position_dodge()) +
  geom_col(position = position_stack(reverse = FALSE)) +
  # scale_fill_brewer(palette="Paired") +
  theme_minimal() +
  labs(
    x = "Age", y = "Population", fill = "Educational \nLevel",
    title = "Relationship Between Age and Educational Status"
  )


```
 Interpretation: Majority of the older adult had attained education within grade 8-12. While little of the participant has education above grade 12. This has significant implication to ARV medication adherence.


 Relationship Between Education and Marital status
```{r}
# Relationship between education and marital status

# Relationship between age and educational level
compliances_marriage_education_df <- compliances_df_clean %>% 
  group_by(maritalstatus_recode, mat_education) %>% 
  count(sort = TRUE) %>% 
  ungroup()

# Visualize age and marital status

compliances_marriage_education_df %>% 
  mutate(age_recode = factor(
    maritalstatus_recode, 
    levels = compliances_marriage_education_df$maritalstatus_recode %>% unique())) %>%
  ggplot(aes(
    x = maritalstatus_recode, 
    y = n, 
    fill = mat_education)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  # geom_col(position = position_stack(reverse = FALSE)) +
  geom_text(aes(label=n), vjust=-0.3, color="black",
            position = position_dodge(0.9), size=3.5)+
  scale_fill_brewer(palette="Paired")+
  theme_minimal() +
  labs(
    x = "Age", y = "Population", fill = "Educational \nLevel",
    title = "Relationship Between Marrital and Educational Status"
  )


```

Interpretation: Literacy level is found to be high across all marital status.This Barchart represents the distribution of education levels in the dataset. It indicates the number of individuals in each education category. For example, there are 421 individuals with education level "Grade 8-12", while post graduate educational level was the least.


EMPLOYMENT  AND ADHERENCE

```{r}
employment_df <- compliances_df_clean %>% 
  select(employ) %>% 
  mutate(id = row_number()) %>% 
  full_join(
    compliances_df3 %>% select(adherence) %>% mutate(id = row_number()),
    by = c("id")
  ) %>% 
  select(-id) %>% 
  group_by(employ, adherence) %>% 
  count(sort = TRUE) %>% 
  mutate(adherence = case_when(
    adherence %in% 0:4 ~ "Low",
    adherence %in% 5:6 ~ "Medium",
    adherence == 7 ~ "High"
    # TRUE ~ "High"
  ))
employment_plot <- employment_df %>%
  ggplot(aes(x = adherence, y = n, fill = employ  )) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_brewer(palette = "Paired") +
  theme_minimal() +
  labs(
    x = "Employment",
    y = "Population",
    fill = "Adherence Level",
    title = "Relationship Between Employment and Adherence Levels"
  )

# Display the bar plot
print(employment_plot)

```
Majority of the unemployed participants had low adherence to medication.



 LIMITATION OF DATASET:The dataset used for analysis  have a limited number of observations that can affect the generalizability of the findings to a larger population. Also there might be mistakes when handling errors and outliers that can impact the accuracy and reliability of the results. 

Additionally, there was inadequate knowledge to determine data  multicollinearity  Also, the test item was on a likert scale which should be assessed and validated, this was challenging. 



CONCLUSION: The analyzed results provide evidence that various factors, including Age, education, larrangement, matdisclosure, MaritalStatus, and soc_cap, have significant associations with adherence behavior. These findings can contribute to a better understanding of the factors influencing adherence and help guide interventions or strategies to improve adherence levels.



# REFERENCE
1 Jialie Fang, Ji-Wei Wang, Jiang Li, Hua Li & Chunhai Shao (2017) The correlates of social capital and adherence to healthy lifestyle in patients with coronary heart disease, Patient Preference and Adherence, , 1701-1707, DOI: 10.2147/PPA.S140787 To link to this article:  https://doi.org/10.2147/PPA.S140787
2 Moudatsou, M. M., Kritsotakis, G., Alegakis, A. K., Koutis, A., & Philalithis, A. E. (2014). Social capital and adherence to cervical and breast cancer screening guidelines: a cross-sectional study in rural Crete. Health & social care in the community, 22(4), 395–404. https://doi.org/10.1111/hsc.12096


