---
title: "Challenge 1 Instructions"
author: "Lauren Zichittella"
desription: "Reading in data and creating a post"
date: "02/18/2023"
format:
  html:
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
  - challenge_1
  - laurenzichittella
  - wild_bird_data
---
## Read in the Data, wild_bird_data.xlsx  

### Summary/Results  

This dataset was utilized by Nee et al to produce a figure to show the relationship
between wild bird weight and population size. 

It consists of two numeric variables, wet body weight in grams and population size
and 146 distinct observations. 

### Technical notes  
- Utilize read_xls  
- Skip import of first two rows  
  - row 1 includes header   
  - row 2 provides variable names I do not want to use  
- Assign new variable names: 
    - Wet body weight [g] = wet_weight_g  
    - Population size = pop_size  
- Output variable class       
- Sort by ascending values of wet_weight_g 

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)

library(readxl)

wild_bird_data2 <- read_xlsx( "_data/wild_bird_data.xlsx", skip = 2, col_names = c("wet_weight_g", "pop_size"))

sapply(wild_bird_data2, class)

arrange(wild_bird_data2,wet_weight_g)
```

## Provide baseline characterization of data 

### Summary/Results  
Both Variables are populated for all observations and represent a wide range of values. Distributions of weight and population size are positively skewed bu outliers. 

### Technical notes
Provide the following information:   
- Print of first 23 observations   
- Print of last 23 observations   
- Filter on presence of missing values for wet_weight_g or pop_size   
- Summarize numeric variables  

```{r}

slice(wild_bird_data2, 1:23)
slice(wild_bird_data2, 124:146)

filter(wild_bird_data2, is.na(wet_weight_g) | is.na(pop_size))

result='asis'
library(summarytools)

dfSummary(wild_bird_data2, varnumbers = FALSE, valid.col = FALSE)

hist(  wild_bird_data2$wet_weight_g
     , main="Distribution of Weight Captured in Wild Bird Datset"
     , xlab="Wet Weight in Grams"
     , col = "red")
hist(  wild_bird_data2$pop_size
     , main="Distribution of Population Size Captured in Wild Bird Datset"
     , xlab = "Population Size"
     , col = "red")

```

## Evaluate relationship between weight and population size graphically 

### Summary/Results  
Plotting weight by population size does not clearly point to any relationship between these variables.  
  
### Technical notes
Plot results on log10 scale  
  
```{r}
# Load packages
require(MASS) # to access Animals data sets
require(scales) # to access break formatting functions
library(ggplot2)
start_plot <- ggplot(wild_bird_data2, aes(x=pop_size, y=wet_weight_g)) + geom_point()

start_plot2 <- start_plot +
     scale_x_log10(breaks = trans_breaks("log10", function(x) 10^x),
              labels = trans_format("log10", math_format(10^.x))) +
     scale_y_log10(breaks = trans_breaks("log10", function(x) 10^x),
              labels = trans_format("log10", math_format(10^.x))) +
     theme_bw() 

start_plot2 +annotation_logticks() +
  ggtitle("Relationship between Weight and Population Size in Wild Birds") +
  xlab("Population Size")+
  ylab("Wet Weight (g)")

```
