---
title: "Challenge 2"
author: "Linda Humphrey"
description: "Reading in data and creating a post"
date: "03/01/2023"
format:
  html:
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
  - challenge_1: hotel_bookings.csv
  - my name: Linda Humphrey
  - dataset: hotel_bookings.csv
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)
library(readxl)
library(lubridate)
library(psych)
library(DT)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Challenge Overview

Today's challenge is to

1)  read in a data set, and describe the data using both words and any supporting information (e.g., tables, etc)
2)  provide summary statistics for different interesting groups within the data, and interpret those statistics

## Read in the Data

Read in one (or more) of the following data sets, available in the `posts/_data` folder, using the correct R package and command.

-   railroad\*.csv or StateCounty2012.xls ⭐
-   FAOstat\*.csv or birds.csv ⭐⭐⭐
-   hotel_bookings.csv ⭐⭐⭐⭐

```{r}
# Exploring hotel_bookings data
library(readr)
dataset<- read_csv("~/Desktop/Dacss601_spring2023/posts/_data/hotel_bookings.csv") 

#View data structure

str(dataset)
```


Add any comments or documentation as needed. More challenging data may require additional code chunks and documentation.

## Describe the data

Using a combination of words and results of R commands, can you provide a high level description of the data? Describe as efficiently as possible where/how the data was (likely) gathered, indicate the cases and variables (both the interpretation and any details you deem useful to the reader to fully understand your chosen data).
* Data gathered was an analysis of hotel bookings from 2015 to 2017

```{r}
library(readr)
dataset<- read_csv("~/Desktop/Dacss601_spring2023/posts/_data/hotel_bookings.csv") 
# Finding summary statistics for 'adults'

summary(dataset$adults)
```
*This data describes demand data for two different types of hotels, with 31 variables describing 40,060 observations and 79,330 observations.*

```{r}
library(readr)
dataset<- read_csv("~/Desktop/Dacss601_spring2023/posts/_data/hotel_bookings.csv") 
# Calculating the summary of the dataset

summary(dataset)

```
*The above data is calculating summary of the Hotel_bookings dataset.*
```{r}
library(readr)
dataset<- read_csv("~/Desktop/Dacss601_spring2023/posts/_data/hotel_bookings.csv") 
# Calculating null value count for all columns

colSums(is.na(dataset))

```
*The above data is calculating null value count for all columns.*
```{r}
library(readr)
dataset<- read_csv("~/Desktop/Dacss601_spring2023/posts/_data/hotel_bookings.csv") 
# Generating a table for the first 100 rows in the dataset.

knitr::kable(head(dataset,n = 100), "pandoc")

```
*The above table shows the first 100 rows in the dataset.*
```{r}
library(readr)
dataset<- read_csv("~/Desktop/Dacss601_spring2023/posts/_data/hotel_bookings.csv") 
# Finding summary statistics for 'children'

summary(dataset$children)
```
*The above shows the summary statistics of children column.*

## Provide Grouped Summary Statistics

Conduct some exploratory data analysis, using dplyr commands such as `group_by()`, `select()`, `filter()`, and `summarise()`. Find the central tendency (mean, median, mode) and dispersion (standard deviation, mix/max/quantile) for different subgroups within the data set.

```{r}
library(readr)
dataset<- read_csv("~/Desktop/Dacss601_spring2023/posts/_data/hotel_bookings.csv") 
# To show the distribution of the data
multi.hist(dataset[,sapply(dataset, is.numeric)])
```
*From the above histograms, August is the busiest month, with the most bookings in 2016 and the second half of the year.*

```{r}
library(readr)
dataset<- read_csv("~/Desktop/Dacss601_spring2023/posts/_data/hotel_bookings.csv")
# Room Summary
room_summary <- dataset %>%
  filter(is_canceled == 0) %>% 
  group_by(reserved_room_type) %>%
  summarize(room_count = n()) %>% 
  arrange(-room_count)

knitr::kable(room_summary)
```
*A is the most popular type of room, so the corporation should increase its number of type A rooms.*
```{r}
library(readr)
dataset<- read_csv("~/Desktop/Dacss601_spring2023/posts/_data/hotel_bookings.csv")
ggplot(dataset,aes(reserved_room_type,fill = (hotel))) +
  geom_bar(position = 'dodge') +
  ylab("Number Of Bookings") +
  xlab("Room Type") +
  ggtitle("Room Type Preferred") +
  labs(fill = 'Hotel Type')
```

*As the City Hotel generates the majority of bookings (66% of all reservations each year), which is far more than the resort hotel, the corporation may design strategies accordingly.
```{r}
library(readr)
dataset<- read_csv("~/Desktop/Dacss601_spring2023/posts/_data/hotel_bookings.csv")

# country with the most Guests
data_country <- dataset %>% group_by(country) %>%  summarise(booking_count = n()) %>% arrange(desc(booking_count))
top_n(data_country,10,booking_count) %>% 
  ggplot(.,aes(country, booking_count)) +
  geom_bar(stat = "identity", width = 0.25, fill ="blue")

```

*Summary of country with the most guest, we can see PRT has the highest rate of guests.*

### Explain and Interpret

Be sure to explain why you choose a specific group. Comment on the interpretation of any interesting differences between groups that you uncover. This section can be integrated with the exploratory data analysis, just be sure it is included.
*The hotels' ultimate goal is to boost their earnings, therefore they want to comprehend and concentrate on everything that can do so.The most important details are that the City Hotel receives the majority of reservations and generates the majority of money, and that PRT.A is the most popular type of room. This helps the customer plan for more guests, make the necessary preparations, and conduct effective marketing..*
