---
title: "Challenge 8 Instructions"
author: "Linda Humphrey"
description: "Joining Data"
date: "04/27/2022"
format:
  html:
    toc: true
    code-copy: true
    code-tools: true
categories:
  - challenge_8: snl
  - data: snl_actors.csv, snl_seasons.csv, snl_casts.csv
  - name: Linda Humphrey
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)
library(viridis)
library(patchwork)
library(hrbrthemes)
library(fmsb)
library(colormap)


knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Challenge Overview

Overview of SNL Dataset

## Read in data
Reading 3 differnt datasets of snl csv files
```{R}
#Load dataset

snl_seasons = read_csv("~/Desktop/601_Spring_2023/posts/_data/snl_seasons.csv", show_col_types = FALSE)

snl_casts = read_csv("~/Desktop/601_Spring_2023/posts/_data/snl_casts.csv", show_col_types = FALSE)

snl_actors = read_csv("~/Desktop/601_Spring_2023/posts/_data/snl_actors.csv", show_col_types = FALSE)

head(snl_seasons)
head(snl_casts)
head(snl_actors)
```

## Briefly Describe the data
The dataset snl_seasons contains 46 observations about the different seasons of "Saturday Night Live". The three datasets consist of actors, casts, and seasons, which contain details about each actor, type, and gender.

```{r}
library(summarytools)

print(summarytools::dfSummary(snl_seasons,
                        varnumbers = FALSE,
                        plain.ascii  = FALSE, 
                        style        = "grid", 
                        graph.magnif = 0.70, 
                        valid.col    = FALSE),
      method = 'render',
      table.classes = 'table-condensed')

print(summarytools::dfSummary(snl_casts,
                        varnumbers = FALSE,
                        plain.ascii  = FALSE, 
                        style        = "grid", 
                        graph.magnif = 0.70, 
                        valid.col    = FALSE),
      method = 'render',
      table.classes = 'table-condensed')

print(summarytools::dfSummary(snl_actors,
                        varnumbers = FALSE,
                        plain.ascii  = FALSE, 
                        style        = "grid", 
                        graph.magnif = 0.70, 
                        valid.col    = FALSE),
      method = 'render',
      table.classes = 'table-condensed')
```


## Tidy Data (as needed)
To tidy the dataset I rename all columns in each data set.
```{r}
# Renaming columns in seasons
snl_seasons <- snl_seasons %>%
  rename(
    Season = sid,
    Year = year,
    FirstEpisode = first_epid,
    LastEpisode = last_epid,
    TotEpisodes = n_episodes,
  )

# Renaming columns in casts
snl_casts <- snl_casts %>%
  rename(
    Actor = aid,
    Season = sid,
    Featured = featured,
    Anchor = update_anchor,
    Episodes = n_episodes,
    EpisodesProp = season_fraction
  )
# Renaming columns in actors
snl_actors <- snl_actors %>% 
  rename(
    Actor = aid,
    Type = type,
    Gender = gender
    )

# Mutate fields
snl_actors <- snl_actors %>%
  mutate(Gender = str_replace(Gender, "andy", "unknown"), Type = str_replace(Type, "unknown", "celebrity")
  )
```


## Join Data

Be sure to include a sanity check, and double-check that case count is correct!
```{R}
library(dplyr)
snl_joined = merge(x=snl_casts, y=snl_actors, by.y = "aid", by.x = "aid")

print(summarytools::dfSummary(snl_joined,
                        varnumbers = FALSE,
                        plain.ascii  = FALSE, 
                        style        = "grid", 
                        graph.magnif = 0.70, 
                        valid.col    = FALSE),
      method = 'render',
      table.classes = 'table-condensed')

```

