---
title: "Challenge 1"
author: "Linda Humphrey"
description: "Reading in data and creating a post"
date: "02/22/2023"
format:
  html:
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
  - challenge_1: railroad_2012_clean_county.csv
  - my name: Linda Humphrey
  - dataset: railroad_2012_clean_county.csv

---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Challenge Overview

Today's challenge is to

1)  read in a dataset, and

The data is about Total railroad employment by state and county

2)  describe the dataset using both words and any supporting information (e.g., tables, etc)

The data contains variables such as,calender year 2012, States, Counties and number of railroad employees in each of the states in accordance to the counties.  

## Read in the Data

Read in one (or more) of the following data sets, using the correct R package and command.

-   railroad_2012_clean_county.csv ⭐
-   birds.csv ⭐⭐
-   FAOstat\*.csv ⭐⭐
-   wild_bird_data.xlsx ⭐⭐⭐
-   StateCounty2012.xls ⭐⭐⭐⭐

Find the `_data` folder, located inside the `posts` folder. Then you can read in the data, using either one of the `readr` standard tidy read commands, or a specialized package such as `readxl`.

```{r}
library(readr)
dataset <- read_csv("~/Desktop/Dacss601_spring2023/posts/_data/railroad_2012_clean_county.csv") 
View(dataset)
#Filter dataset of total_employees >= 2
filter(dataset, total_employees >= 2)
```


```{r}
#Arrange dataset of state & total_employees
dataset <- read_csv("~/Desktop/Dacss601_spring2023/posts/_data/railroad_2012_clean_county.csv")
arrange(dataset, desc(total_employees))
```

Add any comments or documentation as needed. More challenging data sets may require additional code chunks and documentation.

## Describe the data

Using a combination of words and results of R commands, can you provide a high level description of the data? Describe as efficiently as possible where/how the data was (likely) gathered, indicate the cases and variables (both the interpretation and any details you deem useful to the reader to fully understand your chosen data).

```{r}
#| label: summary provides dataset of railroad total employess by state & county
dataset <- read_csv("~/Desktop/Dacss601_spring2023/posts/_data/railroad_2012_clean_county.csv")

dataset <- read_csv("~/Desktop/Dacss601_spring2023/posts/_data/railroad_2012_clean_county.csv")

print(is.data.frame(dataset))
print(ncol(dataset))
print(nrow(dataset))

# Get the max total_employees from dataset

total_employees <- max(dataset$total_employees)
print(total_employees)

# Get details of state with min total_employees from dataset

dataset <- read_csv("~/Desktop/Dacss601_spring2023/posts/_data/railroad_2012_clean_county.csv")


total_employees <- min(dataset$total_employees)
print(total_employees)

# Get details of state with Mean total_employees from dataset

dataset <- read_csv("~/Desktop/Dacss601_spring2023/posts/_data/railroad_2012_clean_county.csv")


total_employees.mean <- mean(dataset$total_employees)
print(total_employees.mean)

```
