---
title: "Challenge 7 Instructions"
author: "Linda Humphrey"
description: "Visualizing Multiple Dimensions"
date: "04/17/2022"
format:
  html:
    toc: true
    code-copy: true
    code-tools: true
categories:
  - challenge_7
  - hotel_bookings
  - australian_marriage
  - air_bnb
  - eggs
  - abc_poll
  - faostat
  - usa_households
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)
library(lubridate)
library(ggplot2)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Challenge Overview

Today's challenge is to:

1)  read in a data set, and describe the data set using both words and any supporting information (e.g., tables, etc)
2)  tidy data (as needed, including sanity checks)
3)  mutate variables as needed (including sanity checks)
4)  Recreate at least two graphs from previous exercises, but introduce at least one additional dimension that you omitted before using ggplot functionality (color, shape, line, facet, etc) The goal is not to create unneeded [chart ink (Tufte)](https://www.edwardtufte.com/tufte/), but to concisely capture variation in additional dimensions that were collapsed in your earlier 2 or 3 dimensional graphs.
   - Explain why you choose the specific graph type
5) If you haven't tried in previous weeks, work this week to make your graphs "publication" ready with titles, captions, and pretty axis labels and other viewer-friendly features

[R Graph Gallery](https://r-graph-gallery.com/) is a good starting point for thinking about what information is conveyed in standard graph types, and includes example R code. And anyone not familiar with Edward Tufte should check out his [fantastic books](https://www.edwardtufte.com/tufte/books_vdqi) and [courses on data visualizaton.](https://www.edwardtufte.com/tufte/courses)

(be sure to only include the category tags for the data you use!)

## Read in data

Read in one (or more) of the following datasets, using the correct R package and command.

  - eggs ⭐
  - abc_poll ⭐⭐
  - australian_marriage ⭐⭐
  - hotel_bookings ⭐⭐⭐
  - air_bnb  ⭐⭐⭐
  - us_hh ⭐⭐⭐⭐
  - faostat ⭐⭐⭐⭐⭐

### Briefly describe the data

The FAOSTAT_livestock.csv dataset is a comprehensive source of information on livestock production and trade, sourced from the FAO. The dataset includes data on various types of livestock, including cattle, sheep, goats, pigs, poultry, and other animals. For each type of livestock, the data includes information on production quantities, trade volumes, and population sizes, as well as related indicators such as feed consumption and meat yield.

The data is organized by country and year, with information spanning several decades in some cases. The dataset also includes aggregate data for regions and the world as a whole.

## Read data
```{R}
#Load data 
FAOSTAT_livestock <- read.csv("~/Desktop/601_Spring_2023/posts/_data/FAOSTAT_livestock.csv")
head(FAOSTAT_livestock)
```

## Tidy Data (as needed)


```{r}
# Group the data by country and year
FAOSTAT_livestock <- FAOSTAT_livestock %>%
  select("Area", "Year", "Livestock", "Indicator", "Value") %>%

```

Are there any variables that require mutation to be usable in your analysis stream? For example, do you need to calculate new values in order to graph them? Can string values be represented numerically? Do you need to turn any variables into factors and reorder for ease of graphics and visualization?

Document your work here.

```{r}

```

## Visualization with Multiple Dimensions


