---
title: "Final Project Assignment#1: Project & Data Description"
author: "Livia Charles"
description: "Project & Data Description"
date: "04/12/2023"
format:
  html:
    df-print: paged
    toc: true
    code-copy: true
    code-tools: true
    css: styles.css
categories:
  - final_Project_assignment_1
  - final_project_data_description
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)
library(stringr) 
library(rgdal)
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE) 
knitr::opts_knit$set(root.dir = "/Users/liviacharles/Documents/R Class Project") 
```

## Part 1. Introduction {#describe-the-data-sets}

1.  Dataset(s) Introduction:

The first data set is the Global River Obstruction Database v1.1 which was 
generated by a research team based in the UNC Chapel Hill Global Hydrology Lab.
The data set was published in December 2021 and the data was collected from 2018 through 2021. 
Each row in GROD represents a different river obstruction that occurs on one of
Earth's major rivers that are contained within the SWOT River Database. 
Information included in each row are the type of obstruction, it's geographic 
location, its distance to the nearest Swot River Database reach, and ID information. 

The second data set is the National Wildlife and Scenic Rivers Database produced by the U.S. Forest Service as a part of their effort to disseminate publicly available data in the  Geospatial Data Discovery site. The database was originally published in 2015 and updated in 2017.
The first column represents the name of the river, the second the county and state where the river is located, the fourth the boundary status of the river, and the fifth the classification of the river—either scenic or recreational. The third and sixth columns are river ID numbers and the seventh and eighth columns describe spatial attributes of the rivers. 

I will also use shapefile data from the National Wildlife and Scenic Rivers Database. However, in the introduction, I only worked with the csv data as the shapefile contains the same information and won't be useful until being used for map making. 

2.  What questions do you like to answer with this dataset(s)?

The GROD data set will allow me to gain an understanding of the most impacted 
large rivers in the world in terms of their obstruction of flow and thus 
indicate rivers or sections of river that could most benefit from dam removal 
or legislation (such as a rights of nature approach) that could help increase 
their ecological health.

The second data set will indicate which US rivers have higher protections. This will in turn provide me with insight into rivers' development restrictions  and free-flowing nature. For example, if a segment of a river is designated "wild" no residential structures can be located near the river, if it's recreational structures must be back 150ft, if it's scenic the structure must be back 250f. I will use this dataset to analyze which of the rivers in the previous database are most free-flowing. 

1.  read the dataset;

```{r}
#Read in the GROD data set 
grod <- read.csv("GROD_v1.1.csv", sep = ",")
#read in the National Wild and Scenic Rivers data set 
nwsr <- read.csv("National_Wild_and_Scenic_Rivers_(Feature_Layer).csv", sep = ",") 
nwsr.shp <- readOGR("./National_Wild_and_Scenic_Rivers_(Feature_Layer)/National_Wild_and_Scenic_Rivers_(Feature_Layer).shp")
```

2.  present the descriptive information of the dataset(s) using the functions in Challenges 1, 2, and 3;


    ```{r}
#look at the head of the data sets
head(grod)
head(nwsr)
#check the dimensions of the GROD data set 
dim(grod) 
dim(nwsr) 
dim(nwsr.shp)
#count the number of unique ids for river obstructions and sword reaches in the GROD data set 
length(unique(grod$grod_id))
length(unique(grod$sword_reach_id)) 
#count the number of unique rivers in the nwsr data set 
length(unique(nwsr$RIVER))
    ```

3.  conduct summary statistics of the dataset(s); especially show the basic statistics (min, max, mean, median, etc.) for the variables you are interested in.

```{r}
#summary statistics of relevant variables 
#Get counts of each type of dam within the GROD database
table(grod$type)
#get summary data for the distance to sword column in the GROD data set
summary(grod$distance_to_sword) 
#get summary information for multiple columns in the nwsr data set 
table(nwsr$BOUNDARYSTATUS)
table(nwsr$CLASSIFICATION)
table(nwsr[, c("BOUNDARYSTATUS", "CLASSIFICATION")])
#create a new column in the nwsr data set that identifies the state of each river 
nwsr$State <- sapply(str_split(nwsr$RIVER, ",\\s*"), `[`, 2)
table(nwsr$State)
```


## 3. The Tentative Plan for Visualization {#the-tentative-plan-for-visualization}

I plan to carry out an analysis that provides a spatial overview of the world's river obstructions and then specifically analyzes the occurrence, or lack thereof, of river obstructions on rivers designated as wild, scenic, or recreational in the U.S. A large part of this analysis will be a spatial visualization including maps of the river obstructions at global and regional scales. I will also create charts that show quantification of the different types of obstructions at global and regional scales. I will then focus my analysis on the wild, scenic, recreational rivers of the U.S. 

By showing the spatial patterns of river obstructions at global and regional scales and quantifying the number of different types of obstructions at different regional levels, it will provide the first indication as to which areas may be of highest priority for management that supports river health (e.g., dam removal). Tying such analyses to the wild, scenic, recreational river categorization will allow me to answer questions such as: Are wild, scenic, and recreational rivers more or less obstructed than other rivers or each other? Which of these rivers could benefit from dam removal? Which are the most flow obstructed of these rivers?

Much of the analysis will hinge on producing maps, so a lot of the processing and preparation of the data will be centered on readying it for mapping. The GROD data will also need to be merged with the wild/scenic/recreational river data by creating a buffer around the wild, scenic, recreational rivers, as well as intersecting it with the GROD data. The wild, scenic, and recreational river csv data will also need to be merged with its corresponding shapefile data. 


4.  (Optional) It is encouraged, **but optional**, to include a coding component of tidy data in this part.

```{r}

```

