---
title: "Challenge 1"
author: "Maanusri Balasubramanian"
description: "Reading in data and creating a post"
date: "05/03/2023"
format:
  html:
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
  - challenge_1
  - maanusri balasubramanian
  - railroads
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Challenge Overview

Today's challenge is to

1)  read in a dataset, and

2)  describe the dataset using both words and any supporting information (e.g., tables, etc)

## Read in the Data

Read in one (or more) of the following data sets, using the correct R package and command.

-   railroad_2012_clean_county.csv ⭐
-   birds.csv ⭐⭐
-   FAOstat\*.csv ⭐⭐
-   wild_bird_data.xlsx ⭐⭐⭐
-   StateCounty2012.xls ⭐⭐⭐⭐

Find the `_data` folder, located inside the `posts` folder. Then you can read in the data, using either one of the `readr` standard tidy read commands, or a specialized package such as `readxl`.

Add any comments or documentation as needed. More challenging data sets may require additional code chunks and documentation.

```{r}
# loading the data
rr <- read_csv("_data/railroad_2012_clean_county.csv")

# printing first 5 rows of the dataset
head(rr)

```

## Describe the data

Using a combination of words and results of R commands, can you provide a high level description of the data? Describe as efficiently as possible where/how the data was (likely) gathered, indicate the cases and variables (both the interpretation and any details you deem useful to the reader to fully understand your chosen data).

```{r}
#| label: summary

# description of the dataset
str(rr)

# number of rows and columns in the dataset
dim(rr)

# column names
colnames(rr)
```

From the above commands we can see that "railroad_2012_clean_county.csv" gives us the count of employees working in various counties in each state for the railroad in 2012. There are a total of 2930 entries. Each row gives us information about the number of employees in a county in the state. There are 3 columns, namely: state, county and total_employees.

```{r}
# Summarizing the data with summary
summary(rr)
```



```{r}
# Arranging entries wrt total employees
arrange(rr, `total_employees`)

# Arranging entries wrt total employees in the descending order
arrange(rr, desc(`total_employees`))
```
From the above result we know that the country 'COOK' in IL has the highest number of employees: 8207 and 1 is the minimum number of employees in any country (many counties have only 1 employee).

```{r}
# Grouping in terms of state to summarise
grouped_rr_state <- rr%>%
  group_by(state)%>%
  summarize(state_employees = sum(total_employees))%>%
  arrange(desc(`state_employees`))
grouped_rr_state

dim(grouped_rr_state)
```
From the above results we know that TX has the highest number of rail road employees: 19839 and AP has the least number of employees: 1.

And from the dimensions of grouped_rr_state, we know that there are 53 unique states in which rail road employees work.