---
title: "Challenge 1 "
author: "Matthew Weiner"
desription: "Reading in data and creating a post"
date: "3/6/2023"
format:
  html:
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
  - challenge_1
  - Matthew_Weiner
  - birds.csv
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Introduction

For this challenge I chose to analyze the dataset birds.csv

## First steps

When first examining this data I wanted to simply print out the results of the CSV file and also view the column names to get a general idea of the format of the file. 

```{r}
library(readr)
birds <- read_csv("_data/birds.csv")
head(birds)
colnames(birds)
```

These column names include things like "Year", "Item", "Value", "Area" so at this point I am suspecting that the data has to do with the sale of some kind of item, most likely some sort of bird (based on the title).

We can also use the following commands to view the dimensions of the data:

```{r}
dim(birds)
```

This shows us that there are 30977 rows in the data.

## Investigating Deeper

By using the following command, we are able to generate a table which shows us the distrbution of entries by country. The results of this show us that this file contains sales from multiple countries indicating to us that the data involves international sale of birds. 
```{r}
head(table(select(birds,Area)))
```

Likewise, we can use a very similar command to view the distrbution of items sold.
```{r}
table(select(birds,Item))
```

The results of this command show us that there are multiple types of birds being sold including chickens, ducks, geese, and guinea fowls, turkeys, and pigeons. 

Finally we can view the range of years in which these sales took place by using the following code block. 
```{r}
year <- select(birds,Year)
min(year)
max(year)
```

This shows us that the sales have run from 1961 until 2018.

## Results

Through the use of multiple R commands I was able to discover that this dataset is about the international trade of over 30000 birds from 1961 until 2018.
