---
title: "MattEckstein_hw1.qmd"
author: "Matt Eckstein"
desription: "Challenge 1 - Analysis of 2012 railroad data"
date: "03/01/2023"
format:
  html:
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
  - challenge_1
  - Matt Eckstein
  - railroad_2012_clean_county.csv
---

```{r}
#| label: setup
#| warning: false

library(tidyverse)

knitr::opts_chunk$set(echo = TRUE)
```



Loading the file and viewing the information in it:
```{r}
railroad_data <- read.csv("posts/_data/railroad_2012_clean_county.csv")

View(railroad_data)

summarize(railroad_data)

```


These data were likely gathered from a survey of occupations across geographies conducted by a federal agency such as the Bureau of Labor Statistics.

Each case is a county in the United States with at least one railroad worker. (The state and county columns are both essential for defining a case, since some county names occur in more than one state, and the state column is necessary for disambiguation.) The total_employees column indicates the number of railroad employees in the relevant county.
