---
title: "Global Hotel Bookings: 2015-2017"
author: "Michele Carlin"
desription: "Exploring, data wrangling, and summarizing hotel bookings data"
date: "02/28/2023"
format:
  html:
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
  - challenge_2
  - Michele Carlin
  - hotel_bookings
---

```{r}
#| label: setup
#| warning: false

library(tidyverse)

knitr::opts_chunk$set(echo = TRUE)
install.packages(summarytools)
library(summarytools)
install.packages(dplyr)
library(dplyr)
```
## Read in and view 'hotel bookings' dataset.
```{r}
library(readr)
hotel_bookings <- read_csv ("_data/hotel_bookings.csv")
View(hotel_bookings)

```
## List of variables in dataset.
```{r}
str(hotel_bookings)
```
## dfSummary
```{r}
view(dfSummary(hotel_bookings))
```
## Describing dataset
## The 'hotel_bookings' dataset contains 119,390 rows and 32 columns.  Each observation is that of a single hotel booking between 2015 and 2017 and provides information on type of resort, arrival date, number of nights/people, room type, country, market segment, etc.

#Create new variable for total number of guests.
```{r}
hotel_bookings <- hotel_bookings %>%
    mutate (guests = adults + children + babies)
select(hotel_bookings, "adults", "children", "babies", "guests")
```
## Create a new 'season' variable using 'case_when'.
```{r}
hotel_bookings <- hotel_bookings %>%
  mutate(season = case_when(
         arrival_date_month == "December" | arrival_date_month == "January" | arrival_date_month == "February" ~ "Winter",
        arrival_date_month == "March" | arrival_date_month == "April" | arrival_date_month == "May" ~ "Spring",
        arrival_date_month == "June" | arrival_date_month == "July" | arrival_date_month == "August" ~ "Summer",
       arrival_date_month == "September" | arrival_date_month == "October" | arrival_date_month == "November" ~ "Fall")
  )
table(select(hotel_bookings, season))
```
## Create crosstabs of Total number of guests by Season. Summer bookings tend to have higher guest totals possibly because families travel  more when kids are on summer break.
```{r}
xtabs(~ guests + season, hotel_bookings)
```

## Calculate min/max/mean average daily rate (adr) and group data by hotel type and arrival month.
```{r}
GrpByHotelType <- hotel_bookings %>%
	group_by(hotel, arrival_date_month) %>%
	summarise(minADR = min(adr), maxADR = max(adr), meanADR = mean(adr)) 
tail(GrpByHotelType)
```

## Graph adr by type of hotel and arrival month.
```{r}
ggplot(hotel_bookings, aes(fill=hotel, x=arrival_date_month, y=adr)) + 
  geom_bar(position="dodge", stat="identity")
```

## Sort data to find outlier(s). 
```{r}
hotel_bookings %>%
  arrange(desc(adr)) %>%
  select(hotel, adr)
```
## One hotel had an average daily rate of -6.4 and another of 5400 (the next highest adr is 510).Re-run bar chart with these two outliers excluded, summarize by season instead of month, and add title/axis labels.
```{r}
hotel_bookings %>%
  filter(adr > 0 & adr < 600) %>%
  ggplot(aes(fill=hotel, x=season, y=adr)) + 
  geom_bar(position="dodge", stat="identity") +
   labs(title = "Hotel Average Daily Rate by Type of Hotel and Season", y = "Avg Daily Rate", x = "Season")
```
## Graph adr by market segment and type of hotel.
```{r}
hotel_bookings %>%
  filter(adr > 0 & adr < 600) %>%
  ggplot(aes(fill=hotel, x=market_segment, y=adr)) + 
  geom_bar(position="dodge", stat="identity") +
     labs(title = "Hotel Average Daily Rate by Type of Hotel and Market Segment", y = "Avg Daily Rate", x = "Market Segment") +
  scale_x_discrete(guide = guide_axis(n.dodge=2))
```
## Aviation and Undefined don't appear to have any Resort Hotel data. Run crosstabs to verify.

## How do I leave empty space for teal column instead of red taking over width of both columns?
```{r}
xtabs(~ market_segment + hotel, hotel_bookings)
```
