---
title: "2003-2021 Average Debt of Various Household Expenses"
author: "Michele Carlin"
desription: "Imported, Tidied, and Summarized data"
date: "03/20/2023"
format:
  html:
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
  - challenge_4
  - Michele Carlin
  - debt
---

```{r}
#| label: setup
#| warning: false

library(tidyverse)
library(summarytools)
library(dbplyr)
library(readxl)

knitr::opts_chunk$set(echo = TRUE)
```

Read in and view summary of 'debt_in_trillions' dataset.

```{r}
sheet_names <- excel_sheets("_data/debt_in_trillions.xlsx")
sheet_names 
debt_trillions <- read_xlsx ("_data/debt_in_trillions.xlsx")
View(debt_trillions)
view(dfSummary(debt_trillions))
```

This dataset contains 74 rows and 8 columns.  Each row represents a particular year/quarter and provides the average debt in trillions for various types of household expenses (e.g., mortgage, auto loan, credit card, etc.). 

Pivot longer so that each row contains one observation.

```{r}
debt_trillions_long <- pivot_longer(debt_trillions, col = c(Mortgage, 'HE Revolving', 'Auto Loan', 'Credit Card', 'Student Loan', Other, Total),
                 names_to="debt_type",
                 values_to = "value")
View(debt_trillions_long)
view(dfSummary(debt_trillions_long))
```

Separate year and quarter into separate variables.
Summarize average debt by type of debt.

```{r}
debt_trillions_long %>%
  separate("Year and Quarter", c("year", "quarter"), ":") %>%
 group_by(year, debt_type) %>%
  summarise(Avg_Debt = mean(value),
            Min_Debt = min(value),
            Max_Debt = max(value))
```

Summarize total debt only by year.

```{r}
debt_trillions_long %>%
  separate("Year and Quarter", c("year", "quarter"), ":") %>%
  filter(debt_type == 'Total') %>%
  group_by(year, debt_type) %>%
  summarise(Avg_Debt = mean(value),
            Min_Debt = min(value),
            Max_Debt = max(value))
```

Read in and view summary of 'abc polls' dataset.

```{r}
abc_poll <- read_csv ("_data/abc_poll_2021.csv")
View(abc_poll)
view(dfSummary(abc_poll))
```

Remove 'skipped' answers from Q1_a variable.

```{r}
table(select(abc_poll, Q1_a))

abc_poll <- abc_poll%>%
  mutate(Q1_a = str_remove(Q1_a, "Skipped"))

table(select(abc_poll, Q1_a))

```


Specify the order of the ppinc7 values.

```{r}
ppinc7 <- factor('ppinc7', ordered = TRUE, levels = c("$150,000 or more", "$100,000 to $149,999", "$75,000 to $99,999", "$50,000 to $74,999", "$25,000 to $49,999", "$10,000 to $24,999", "Less than $10,000"))
ppinc7
table(ppinc7)
```
