---
title: "Challenge 1"
author: "Nanci Kopecky"
desription: "County Railroad Employees"
date: "02/21/23"
format:
  html:
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
  - challenge_1
  - railroads
  - faostat
  - wildbirds
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Challenge Overview

Today's challenge is to

1)  read in a dataset, and

2)  describe the dataset using both words and any supporting information (e.g., tables, etc)

## Read in the Data

Read in one (or more) of the following data sets, using the correct R package and command.

-   railroad_2012_clean_county.csv ⭐
-   birds.csv ⭐⭐
-   FAOstat\*.csv ⭐⭐
-   wild_bird_data.xlsx ⭐⭐⭐
-   StateCounty2012.xls ⭐⭐⭐⭐

Find the `_data` folder, located inside the `posts` folder. Then you can read in the data, using either one of the `readr` standard tidy read commands, or a specialized package such as `readxl`. ###Reading in Railroad data I had to set the working directory in order for this code to work.

```{r}
library(readr)
railroad<-read.csv(file = "_data/railroad_2012_clean_county.csv",
                header=TRUE,
                sep = ","
                )

```

Add any comments or documentation as needed. More challenging data sets may require additional code chunks and documentation.

My comment is that I was able to read in the data after I set the working directory. I selected Session, Set Working Directly, then Choose Directly.

## Describe the data

Using a combination of words and results of R commands, can you provide a high level description of the data? Describe as efficiently as possible where/how the data was (likely) gathered, indicate the cases and variables (both the interpretation and any details you deem useful to the reader to fully understand your chosen data).

```{r}
#| label: summary
```

Exploring some commands.

```{r}
head(railroad)
nrow(railroad)
ncol(railroad)
railroad$total_employees
mean(railroad$total_employees)
railroad %>%
  summarize(
    mean_employees=mean(total_employees),
    median_employees=median(total_employees),
    sd_employees=sd(total_employees))
```

The railroad data has 3 variables (State, County, Total Employees) and 2930 observations. The code below is exploring some commands.

The mean is 87 employees and the median is 21, wow, there must be some extreme values! And the standard deviation is 284.

### Visualize the Data

I was wondering about the extreme values for the railroad county employees. The box plot shows the outliers and the histogram shows the right tail.

```{r}
boxplot(railroad$total_employees)

```

```{r}
hist(railroad$total_employees)
```
