---
title: "Challenge 1"
author: "Nisarg Shah"
description: "Reading in data and creating a post"
date: "03/02/2023"
format:
  html:
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
  - challenge_1
  - railroads
  - faostat
  - wildbirds
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Read in the Data
-   birds.csv ⭐⭐

```{r}
#| label: reading data
birds <- read_csv("_data/birds.csv")
birds
```

Add any comments or documentation as needed. More challenging data sets may require additional code chunks and documentation.

## Describe the data

This dataset has 30977 rows and 14 columns. The different type of birds in the dataset are (Chickens, Ducks, Geese and Guinea Fowls, Pigeons other birds, Turkeys). The dataset shows the different area of the birds along with a code designated to the area. The dataset shows the Year, with the amount of units of that bird and the value of the bird as well. The years range from 1961-2018. Each point of data is marked with where the data might have come from such as Official sources or unofficial sources.

```{r}
#| label: first few rows of the data
head(birds)
```

```{r}
#| label: Dimensions of Data
dim(birds)
```

```{r}
#| label: summary
library(summarytools)
dfSummary(birds)
```