---
title: "Challenge 4"
author: "Nisarg Shah"
description: "More data wrangling: pivoting"
date: "08/18/2022"
format:
  html:
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
  - challenge_4
  - abc_poll
  - eggs
  - fed_rates
  - hotel_bookings
  - debt
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Challenge Overview

Today's challenge is to:

1)  read in a data set, and describe the data set using both words and any supporting information (e.g., tables, etc)
2)  tidy data (as needed, including sanity checks)
3)  identify variables that need to be mutated
4)  mutate variables and sanity check all mutations

## Read in data

Read in one (or more) of the following datasets, using the correct R package and command.

-   abc_poll.csv ⭐
-   poultry_tidy.xlsx or organiceggpoultry.xls⭐⭐
-   FedFundsRate.csv⭐⭐⭐
-   hotel_bookings.csv⭐⭐⭐⭐
-   debt_in_trillions.xlsx ⭐⭐⭐⭐⭐

```{r}
df <- read.csv("_data/abc_poll_2021.csv", fileEncoding = "UTF-8")
head(df)
```

### Briefly describe the data

This data appears to be a survey dataset, possibly related to public opinion or market research. Each row represents a respondent/participant and includes several variables that provide information about their demographic characteristics and opinions.

## Tidy Data (as needed)

Is your data already tidy, or is there work to be done? Be sure to anticipate your end result to provide a sanity check, and document your work here.

It appears that the data is already in a "tidy" format, with each row representing a separate observation and each column representing a separate variable.

```{r}
# summarize the data
summary_table <- summarytools::dfSummary(df,
    plain.ascii = FALSE,
    style = "grid",
    valid.col = FALSE
)

# view the summary table
summary_table
```

Any additional comments?

## Identify variables that need to be mutated

Are there any variables that require mutation to be usable in your analysis stream? For example, are all time variables correctly coded as dates? Are all string variables reduced and cleaned to sensible categories? Do you need to turn any variables into factors and reorder for ease of graphics and visualization?

Document your work here.

The variable "ppeduc5" is currently in character format, which may need to be converted to a factor if it is to be used in statistical modeling or visualization.

```{r}
```

Any additional comments?
