---
title: "railroad_2012_clean_county"
author: "PoChun Yang"
desription: "Analysing railroad clean county 2012"
date: "02/21/2023"
format:
  html:
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
  - challenge_1
  - railroads
  - PoChunYang
  - tidyverse
  - summarytools
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)
library(readxl)
library(summarytools)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Challenge Overview

Today's challenge is to

1)  read in a dataset, and

2)  describe the dataset using both words and any supporting information (e.g., tables, etc)

## Read in the Data

Read in one (or more) of the following data sets, using the correct R package and command.

-   railroad_2012_clean_county.csv ⭐
-   birds.csv ⭐⭐
-   FAOstat\*.csv ⭐⭐
-   wild_bird_data.xlsx ⭐⭐⭐
-   StateCounty2012.xls ⭐⭐⭐⭐

Find the `_data` folder, located inside the `posts` folder. Then you can read in the data, using either one of the `readr` standard tidy read commands, or a specialized package such as `readxl`.

```{r}
railroad <- read.csv("_data/railroad_2012_clean_county.csv")
# we want to skip two rows of the data
skip_data<-read.csv("_data/railroad_2012_clean_county.csv",skip = 2)
railroad
skip_data
```

Add any comments or documentation as needed. More challenging data sets may require additional code chunks and documentation.

## Describe the data

Using a combination of words and results of R commands, can you provide a high level description of the data? Describe as efficiently as possible where/how the data was (likely) gathered, indicate the cases and variables (both the interpretation and any details you deem useful to the reader to fully understand your chosen data).

We can get the state and county frequency shown out by using dfsummary command.
In the USA, WASHINGTON is the most frequent naming county. 
Moreover, the highest total number of employees is 8207.

```{r}
#| label: summary
#head() this command show the first element to six element 
head(railroad)
#dim() show the railroad's row and column
dim(railroad)
# colnames() show the column names "state","county","total_empolyee"
colnames(railroad)
#use the select to get the state and total_employees without county
select(railroad,"state","total_employees")
# we use filter to show the TX's data
filter(railroad,`state` == "TX")
# show the summary of the railroad_2012_clean_county.csv
dfSummary(railroad)
# the command is that I want to get each state of the all employees
railroad %>% 
  group_by(state) %>% 
  mutate(all=sum(total_employees))

```
