---
title: "Challenge 8"
author: "PoChun Yang"
description: "Joining Data"
date: "04/26/2023"
format:
  html:
    toc: true
    code-copy: true
    code-tools: true
categories:
  - challenge_8
  - railroads
  - snl
  - faostat
  - debt
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)
library(ggplot2)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Challenge Overview

Today's challenge is to:

1)  read in multiple data sets, and describe the data set using both words and any supporting information (e.g., tables, etc)
2)  tidy data (as needed, including sanity checks)
3)  mutate variables as needed (including sanity checks)
4)  join two or more data sets and analyze some aspect of the joined data

(be sure to only include the category tags for the data you use!)

## Read in data

Read in one (or more) of the following datasets, using the correct R package and command.

  - faostat ⭐⭐
  
In the read data, I read all th fao data here. There would be four of data which call egg, livestock, cattle, country. 
```{r}
egg<-read_csv("_data/FAOSTAT_egg_chicken.csv",show_col_types = FALSE)
head(egg,10)
dim(egg)
livestock<-read_csv("_data/FAOSTAT_livestock.csv",show_col_types = FALSE)
head(livestock,10)
dim(livestock)
cattle<-read_csv("_data/FAOSTAT_cattle_dairy.csv",show_col_types = FALSE)
head(cattle,10)
dim(cattle)
```

### Briefly describe the data

## Tidy Data (as needed)

The data is extremely tidy data. Therefor, I would not use any command to make a clean data. 
As the above table, I could know 14 of the columns with same column name. Therefore, I could join data easily.

## Join Data

Be sure to include a sanity check, and double-check that case count is correct!

The first joins cattle data and livestock. As I used the full_join command here, I got the  118565 rows. I want to prove my answer is correct so I have to sum the rows between those data. In the cattle data, there are 36449 rows. In the livestock data, there are 82116 rows. Therefore, I sum both of the data and got the 118565. This is how I check I am correct. 

```{r}
cattle_live<-full_join(cattle,livestock)
dim(cattle_live)
head(cattle_live)
```
As the command, I want to double check whether my idea is correct or not. Therefore, I sum the rows of egg and cattle data. 36449 + 38170 = 74619. 

```{r}
egg_cattle<-full_join(egg,cattle)
dim(egg_cattle)
head(egg_cattle)
```

