---
title: "Challenge 6"
author: "Poobigan Murugesan"
description: "Visualizing Time and Relationships"
date: "05/18/2023"
format:
  html:
    toc: true
    code-copy: true
    code-tools: true
categories:
  - challenge_6
  - hotel_bookings
  - poobigan murugesan
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)
library(ggplot2)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Challenge Overview

Today's challenge is to:

1)  read in a data set, and describe the data set using both words and any supporting information (e.g., tables, etc)
2)  tidy data (as needed, including sanity checks)
3)  mutate variables as needed (including sanity checks)
4)  create at least one graph including time (evolution)
   - try to make them "publication" ready (optional)
   - Explain why you choose the specific graph type
5)  Create at least one graph depicting part-whole or flow relationships
   - try to make them "publication" ready (optional)
   - Explain why you choose the specific graph type


## Read in data



  

```{r}
df<-read.csv("_data/hotel_bookings.csv")
```
Summary
```{r}
dim(df)
summary(df)
```


### Briefly describe the data

The dataset contains data on hotel bookings and comprises a total of 119,390 entries. It encompasses various information, including hotel type, cancellation status, lead time, arrival date (year, month, day), duration of stay, number of adults, children, and babies, meal preferences, country of origin, market segment, distribution channel, previous cancellations, reserved and assigned room types, booking changes, deposit type, days on the waiting list, customer type, average daily rate, required car parking spaces, and total number of special requests.

Unique countries in dataset (178)
```{r}
df$country%>%unique()
```

Unique hotels in dataset (2)
```{r}
df$hotel%>%unique()
df$arrival_date_year %>%unique()
```

## Tidy Data (as needed)

To observe the reservations trend over time, we merge all the relevant fields(date, month, year) into a single date field. However, we will keep the "arrival_month" field unchanged in order to visualize the monthly trend.
```{r}

df$date <- as.Date(paste(df$arrival_date_year, df$arrival_date_month, df$arrival_date_day_of_month, sep = "-"), format = "%Y-%B-%d")
df$date %>% head()
```






## Time Dependent Visualization
Number of reservations trend for City Hotel:
```{r}
df_plot2 <-df %>%
                  filter(hotel == "City Hotel")%>%
                  group_by(date)%>%
                  summarise(count = n())
  
ggplot(df_plot2, aes(x=date, y=count)) +
  geom_line()
```

## Visualizing Part-Whole Relationships

```{r}
df_plot <-df %>% 
                                  group_by(arrival_date_month, hotel) %>%
                                  summarise(count = n()) %>% 
                                  mutate(arrival_date_month = factor(arrival_date_month, levels = month.name)) %>%
                                  arrange(arrival_date_month)

ggplot(df_plot, aes(arrival_date_month, count, col=hotel, group=hotel)) + geom_line(size = 2) + geom_point(size = 3) + scale_color_manual(values=c("#CC6666", "#9999CC"))
```

The city hotel consistently has more reservations than the resort hotel. They also follow similar trends, peaking and dipping during the same months.
