---
title: "Challenge 2"
author: "Pooja Shah"
description: "Data wrangling: using group() and summarise()"
date: "04/26/2023"
format:
  html:
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
  - challenge_2
  - hotel_bookings
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)
library(summarytools)
library(dplyr)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Challenge Overview

Today's challenge is to

1)  read in a data set, and describe the data using both words and any supporting information (e.g., tables, etc)
2)  provide summary statistics for different interesting groups within the data, and interpret those statistics

## Read in the Data

Read in one (or more) of the following data sets, available in the `posts/_data` folder, using the correct R package and command.

-   railroad\*.csv or StateCounty2012.xls ⭐
-   FAOstat\*.csv or birds.csv ⭐⭐⭐
-   hotel_bookings.csv ⭐⭐⭐⭐

```{r}
#read and print the first few rows of data
booking <- read.csv("_data/hotel_bookings.csv")
head(booking)
```

Add any comments or documentation as needed. More challenging data may require additional code chunks and documentation.

## Describe the data

Using a combination of words and results of R commands, can you provide a high level description of the data? Describe as efficiently as possible where/how the data was (likely) gathered, indicate the cases and variables (both the interpretation and any details you deem useful to the reader to fully understand your chosen data).

```{r}
#| label: summary
dfSummary(booking)
```

## Provide Grouped Summary Statistics

Conduct some exploratory data analysis, using dplyr commands such as `group_by()`, `select()`, `filter()`, and `summarise()`. Find the central tendency (mean, median, mode) and dispersion (standard deviation, mix/max/quantile) for different subgroups within the data set.

```{r}
market <- booking %>%
  group_by(market_segment)
market %>%
  select(hotel, lead_time)
#summarise for market
summarise(market)

#trying different central tendency methods such as mean, median, mode
booking %>%
  summarise(mean_lead_time = mean(lead_time, na.rm = TRUE))
booking %>%
  summarise(median_stays_in_week_nights = median(stays_in_week_nights, na.rm = TRUE))
booking %>%
  summarise(mode_company = mode(company))

#trying different dispersion methods such as standard deviation, min, max, quantile of lead_time
booking %>%
  summarise(min = min(lead_time), max = max(lead_time), standardDeviation = sd(lead_time, na.rm=TRUE))

booking %>%
  summarise(quantile = quantile(lead_time, na.rm = TRUE))
```

### Explain and Interpret

Be sure to explain why you choose a specific group. Comment on the interpretation of any interesting differences between groups that you uncover. This section can be integrated with the exploratory data analysis, just be sure it is included.

Using summarise we were able to see the distinct values of market_segment.
Quantile gave us the 0th(min), 25th, 50th, 75th and 100th(max) percentile of data.
