---
title: "Challenge 2"
author: "Prachiti Parkar"
description: "Data wrangling and Data Sumarization to find statistics in data"
date: "03/08/2023"
format:
  html:
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
  - challenge_2
  - faostat
  - faostat_livestock
  - Prachiti Parkar
---

```{r}
#| label: setup
#| warning: false
#| message: false
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Read in the Data
I have selected faostat_livestock dataset to do my analysis on.

```{r}
# Importing the faostat_livestock dataset and reading them
library(readr)
data <-read_csv("_data/FAOSTAT_livestock.csv")
#Lets see the first 3 rows of the dataset
head(data, 3)
```

## Describe the data

Using R commands I am exploring the data and coming up with the description from the data.

There are total 82117 rows and 14 columns
The data ranges from year 1961 to 2018

### Checking the variation in data for each column

```{r}
#| label: summary
# Check the variation in data for each column
summary(data)
```

## Provide Grouped Summary Statistics

Conducting  exploratory data analysis, using dplyr commands such as `group_by()`, `select()`, `filter()`, and `summarise()`.


```{r}
#install.packages("dplyr")
library(dplyr)
```

Grouping by the Area and finding the mean_value
```{r}
data %>% 
  group_by(Area) %>% 
  summarise(mean_value = mean(Value))
```

Table output by selecting Columns "Area", "Year", and "Value".
```{r}
data %>% 
  select(Area, Year, Value)
```

I wanted to check out how any data rows / points are available for Afghanistan and Ireland in the year of 1962 for the item Asses. There are 2 datapoints available.
```{r}
data %>% 
  filter(Area == "Afghanistan" | Area == "Ireland",Item == "Asses", Year == 1962)
```


## Table Value Statistics
Find the central tendency (mean, median, mode) and dispersion (standard deviation, mix/max/quantile) for Value within the data set.
```{r}
#deleting the rows with Null values
data2 <- na.omit(data)
# Find the central tendency (mean, median, mode) and dispersion (standard deviation, mix/max/quantile) for different subgroups within the data set.
data2%>%
  group_by(Year)%>%
  summarize(mean.Value = mean(Value, na.rm = TRUE), median.Value = median(Value, na.rm = TRUE), min.Value = min(Value, na.rm = TRUE), max.Value = max(Value, na.rm = TRUE), sd.Value = sd(Value, na.rm = TRUE), var.Value = var(Value, na.rm = TRUE), IQR.Value = IQR(Value, na.rm = TRUE))
```

Some Conclusion from seeing the data summary

1. The Data is Collected from the year 1961 - 2018
2. The Element column just has one value Stocks and the Element Code also has only 1 value - 5111
3. The Area Code varies from 1 - 5504
4. The Value Column has a lot of variance

Some insights:
There is no need of both Year Code and Year, either can be used to save space

