---
title: "Challenge 6"
author: "Pradhakshya Dhanakumar"
desription: "Worked with Hotel Bookings Data "
date: "05/05/2023"
format:
  html:
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
  - Challenge 6
  - Pradhakshya Dhanakumar
  - Hotel Bookings
  - ggplot2
---

```{r}
#| label: setup
#| warning: false

library(tidyverse)
library(dplyr)
knitr::opts_chunk$set(echo = TRUE)
```

## Reading Data
```{r}
df <- read.csv("_data/hotel_bookings.csv")
```

```{r}
head(df)
```

## Describe Data
```{r}
colnames(df)
```

```{r}
dim(df)
```


## Tidying the data

```{r}
df<- na.omit(df)
```


## Time dependent Visualization
```{r}
library(ggplot2)
```

The resulting plot shows the total number of bookings for each month and hotel combination in the year 2016 using the filter, group by, ggplot fucntions. The bars are color-coded by hotel - City or Resort Hotel.

```{r}
df %>%
  filter(arrival_date_year == 2015) %>%
  group_by(arrival_date_month, hotel) %>%
  summarise(total_bookings = n()) %>%
  ggplot(aes(x = arrival_date_month, y = total_bookings, fill = hotel)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(x = "Month", y = "Total Bookings", title = "Monthly bookings by hotel") +
  scale_fill_manual(values = c("#4e79a7", "#f28e2c"), 
                    name = "Hotel", 
                    labels = c("City Hotel", "Resort Hotel"))
```

## Visualizing Part-Whole Relationships

This will create a stacked bar chart showing the total number of bookings for each month, with the cancelled bookings shown in red and the non-cancelled bookings shown in green. The legend will show the range of cancelled bookings, with more cancelled bookings being represented by darker shades of red.

```{r}
df %>% 
  group_by(arrival_date_month) %>% 
  summarize(
    total_bookings = n(),
    cancelled_bookings = sum(is_canceled)
  ) %>% 
  ggplot(aes(x = arrival_date_month, y = total_bookings, fill = cancelled_bookings)) + 
  geom_col() + 
  scale_fill_gradient(low = "green", high = "red") +
  labs(title = "Total Bookings and Cancelled Bookings by Month", 
       x = "Month of Arrival", y = "Total Bookings") +
  theme_minimal()
```

