---
title: "Challenge 3"
author: "Pradhakshya Dhanakumar"
desription: "Worked with Animal Weights Dataset"
date: "03/08/2023"
format:
  html:
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
  - Challenge 3
  - Pradhakshya Dhanakumar
  - Animal Weights
---

```{r}
#| label: setup
#| warning: false

library(tidyverse)

knitr::opts_chunk$set(echo = TRUE)
```

## Reading Data
Read the data from a .csv file
```{r}
data <- read_csv("_data/animal_weight.csv")
print(data)
```
## Shape of Data

To find total rows and columns in the dataset
```{r}
nrow(data)
ncol(data)
```
Data Summary
```{r}
summary(data)
```

We can see that the animal_weight data shows the count of different cattle (Buffaloes, Chicken, Turkeys etc) across different regions. There are 9 rows and 17 columns in the given dataset. The 17 columns in this dataset make it difficult to understand or work on the data further. So we can use pivot_longer() function to restructure the data making it more efficient.


There are 9 rows (observations) and (17-1) columns (variables). Hence, we will be needing 9*16 = 144 rows and 3 columns of data(region, cattle type and weight) in the restructured data.

## Pivot Data
```{r}
data_longer<-pivot_longer(data, col=-`IPCC Area`,
                                    names_to = "Livestock",
                                    values_to = "Weight")
print(data_longer)

```

Dimensions of restructured data
```{r}
dim(data_longer)
```

Now we can see that on using pivot_longer(), we get a dataset with 144 rows and 3 columns as expected.

