---
title: "Challenge 1"
author: "Pradhakshya Dhanakumar"
desription: "Worked with snl_seasons data"
date: "02/26/2023"
format:
  html:
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
  - Challenge 1
  - Pradhakshya Dhanakumar
  - SNL_SEASONS DATA
---

```{r}
#| label: setup
#| warning: false

library(tidyverse)

knitr::opts_chunk$set(echo = TRUE)
```
## Reading Data
Read the data from a .csv file
```{r}
data <- read_csv("_data/snl_seasons.csv")
print(data)
```

Top 10 rows of the data
```{r}
head(data,10)
```
Bottom 2 rows of data
```{r}
tail(data,2)
```


Selecting specific columns from data and displaying top 6 rows
```{r}
actors = select(data, "year", "n_episodes")
head(actors)
```

## Data Description
Data type of each column
```{r}
as.data.frame(sapply(data, class))
```

Dimension of data
```{r}
dim(data)
````

Printing column names
```{r}
colnames(data)
```
Total rows in dataset
```{r}
nrow(data)
```
Total columns in dataset

````{r}
ncol(data)
```
## Data Summary
```{r}
summary(data)
```




The dataset used has 46 rows and 5 columns . Each row has information on which year the season was released along with when the first and last episodes were released and the total number of episodes in that particular season. All of the columns have the data type as 'numeric'. We can see the summarized info for each column using the summary() command. Most of the seasons have an average of 20 episodes.
