---
title: "Challenge 6"
author: "Pranav Komaravolu"
description: "Visualizing Time and Relationships"
date: "05/19/2023"
format:
  html:
    toc: true
    code-copy: true
    code-tools: true
categories:
  - challenge_6
  - hotel_bookings
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)
library(ggplot2)
library(dplyr)
options(dplyr.summarise.inform = FALSE)
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

<!--
## Challenge Overview

Today's challenge is to:

1)  read in a data set, and describe the data set using both words and any supporting information (e.g., tables, etc)
2)  tidy data (as needed, including sanity checks)
3)  mutate variables as needed (including sanity checks)
4)  create at least one graph including time (evolution)
   - try to make them "publication" ready (optional)
   - Explain why you choose the specific graph type
5)  Create at least one graph depicting part-whole or flow relationships
   - try to make them "publication" ready (optional)
   - Explain why you choose the specific graph type

[R Graph Gallery](https://r-graph-gallery.com/) is a good starting point for thinking about what information is conveyed in standard graph types, and includes example R code.

(be sure to only include the category tags for the data you use!)
-->
## Data Description

Reading the data
```{r}
data <- read_csv("_data/hotel_bookings.csv", show_col_types = FALSE)
head(data)
```
Columns in the dataset:
```{r}
colnames(data)
```
The dimensions of the dataset are as follows:
```{r}
dim(data)
```
There are 32 columns and 119390 rows in the dataset.

## Tidying the Data
I plan on visualizing the number of people who stayed in a hotel during each year so the dataset can be assumed to be tidy and can be used for visualization.

## Time Dependent Visualization
First I want to analyse at which of these years the hotels had a maximum number of people staying at the hotel. To do that first I will have to filter out the people who cancelled from the dataset, then we can group the data on hotel and arrival_date_year and this data could be summarised to obtain the number of bookings in each of the hotels during different years. The command is as follows:
```{r}
yearly_data <- data%>%
  filter(is_canceled == 0)%>%
  group_by(hotel, arrival_date_year)%>%
  summarise(yearly_count = n())
head(yearly_data)
```
This data can be efficiently depicted using a histogram as it involves frequencies.
```{r}
ggplot(yearly_data)+
  geom_bar(aes(x = arrival_date_year, y = yearly_count, fill= hotel), stat = "identity", position = "dodge")
```
Therefore from the above bar plot both the hotels seem to perform well in the year 2016, hosting more number of guests than any other year.

## Visualizing Part-Whole Relationships
Since there are some cancelled bookings in the dataset we can plot the total bookings while depicting a part with canceled bookings. To get such a visualization the following command can be run:
```{r}
total_yearly_data <- data%>%
                      group_by(hotel, arrival_date_year) %>%
                      summarise(yearly_counts = n(),
                                canceled_bookings = sum(is_canceled))
head(total_yearly_data)
```
Now the plot depicting the different parts of the data is as follows:
```{r}
ggplot(total_yearly_data)+
  geom_col(aes(x = arrival_date_year, y = yearly_counts, fill = canceled_bookings))
```

