---
title: "Final Project Assignment#1: Pranav Bharadwaj Komaravolu"
author: "Pranav Bharadwaj Komaravolu"
description: "Project & Data Description"
date: "04/18/2023"
format:
  html:
    toc: true
    code-fold: false
    code-copy: true
    code-tools: true
categories:
  - Final Project Assignment#1
  - Credit score evaluation
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)
library(readr)
library(mosaic)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Introduction
1. The problem I chose to address through my final project is to identify how the credit score is impacted by various socioeconomic factors. Credits refer to the power of borrowing money with a promise of repayment in the future. It supports two main actors in the economy 1) consumers and 2) banks. This ensures rapid progress/advancements in the well being of the people/governments. Credit score refers to a rating or an assessment offered by banks/agencies based on the societal status and analytics on certain economic trends and individual's behavioral patterns. Based on this value banks provide credits which thus shall contribute to the well being of the consumers. Being a new user of the credit card I was curious on how the credit system works and what parameters/attributes have more impact in determining the credit score of an individual. Each row in the datasets contains various attributes such as name, credit_accounts, delays/defaults in loan repayment and so on.
  
  2. The questions I will be addressing through this project are:
      *  What are the parameters that impact the credit score/assessment the most and what are their correlations?
      *  Which model works best to classify individuals into groups of "Good", "Standard" and "Low"?

## Dataset

For this task I have identified the "Credit score classification" dataset on kaggle. The number of available datasets for this task are very small in number and out of those datasets the current choice seemed more promising. 

1.    reading the dataset:
```{r}
data <- read_csv(" 601_Spring_2023 project/PranavKomaravolu_FinalProjectData/train.csv")
head(data)
```

2. The datset is very vast and its dimensions are as follows:
```{r}
dim(data)
```
There are a 100000 rows and 28 columns in the dataset, from the above head of the dataset we can also observe some impurity/information gap.

The different columns in the dataset are as follows:
```{r}
names(data)
```

The number of unique customers can be obtained as follows:

From the above table we observe "Customer_ID" to be a unique attribute assigned to each customer so if we can compute the number of unique customer ids we can get the total number of customers observed by the sample. And also we can simultaneously obtain different labels that are assigned to the induviduals as their corresponding credit score/evaluation.
```{r}
  data %>%
  select(Customer_ID, Credit_Score) %>%
  summarize_all(n_distinct)
```

So there are about 12500 customers and each customer is assigned one of the three labels for the Credit_Score.

The additional summary statistics for different columns are as follows:
```{r}
summary(data)
```

So each of these columns indicate a specific parameter that impacts the target attribute "Credit_Score". The question I am trying to address is how each of these impoact or affect the credit score and what weights can we assign to each of these attributes during credit evaluation.

## Visualization Plan
Firstly, I will tidy the data identify empty/NA fields and try to populate such rows using various strategies. To visualize and analyse the problem I would like to use a variety of plots which focus on how the target attribute is affected by the varying factors. One such plot being a correlation plot. Following which I would like to try out different classification algorithms like decision trees, logistic regression and asses their performance using the ROC curve and would also like to evaluate the variations caused due to change in empty fields handling strategies.  



