---
title: "Default of credit card clients dataset - UCI Machine Learning Repository"
author: "Pranav Komaravolu"
desription: "Credit card default prediction dataset collected in Taiwan"
date: "02/22/2023"
format:
  html:
    toc: true
    code-fold: false
    code-copy: true
    code-tools: true
categories:
  - challenge_1
  - Pranav Bharadwaj Komaravolu
  - UCI_Credit_Card.csv
---

```{r}
#| label: setup
#| warning: false

library(tidyverse)
library(readr)

knitr::opts_chunk$set(echo = TRUE)
```
<!--## Instructions

This document provides yaml header inforamtion you will need to replicate each week to submit your homework or other blog posts. Please observe the following conventions:

-   Save your own copy of this template as a blog post in the `posts` folder, naming it `FirstLast_hwX.qmd`
-   Edit the yaml header to change your author name - use the same name each week
-   include a *description* that is reader friendly
-   update the *category* list to indicate the type of submission, the data used, the main packages or techniques, your name, or any thing else to make your document easy to find
-   edit as a normal qmd/rmd file


```{r}
x <- c(2,3,4,5)
mean(x)
```
-->
## Sources
The dataset was obtained from the University of California Irvine (UCI) Machine Learning Repository. With the dataset labeled "Default of credit card clients dataset". The dataset can be accessed <a href="https://archive.ics.uci.edu/ml/datasets/default+of+credit+card+clients">here</a>.

## Reading the dataset
To code to read the dataset is as follows:
```{r}
data <- read_csv('_data/UCI_Credit_Card.csv')
head(data)
```
The head of the table is shown above.

The dataset with skipped header is as follows:
```{r}
data2 <- read_csv('_data/UCI_Credit_Card.csv', skip=1) # the header is skipped
head(data2)
```



## Dataset Description
The number of rows in the dataset can be obtained as follows:
```{r}
nrow(data)
```
The number of fields or columns can be obtained as follows:
```{r}
ncol(data)
```
The different columns are as follows:
```{r}
names(data)
```
The data types of different columns are as follows: 
```{r}
as.data.frame(sapply(data, class))
```

Since there are some categorical attributes we can obtain the number of unique elements in each column as follows:
```{r}
getUniqueValues <- function(column_label) {
  data %>% select(all_of(column_label)) %>% n_distinct(.)
}
for (i in names(data)) {
  print(i)
  print(getUniqueValues(i))
}
```

The dataset comprises of 25 different columns, 30000 rows. Each of these columns depict a particular attribute of the data here a person and the column "default.payment.next.month" depicts whether the person defaulted the credit payment or not. All the attributes are preprocessed so that ML algorithms can efficiently train on the data as we can observe that even some of the categorical values are converted to integer type. From the above cell we can observe some attributes such as "SEX", "EDUCATION", "MARRIAGE" and the "default.payment.next.month" have very less number of unique values suggesting that these attributes are categorical.

## Dataset summary
To summarize the dataset we can use the following code:
```{r}
summary(data)
```



<!--## Rendering your post

When you click the **Render** button a document will be generated that includes both content and the output of embedded code. 

:::{.callout-warning}
Be sure that you have moved your `*.qmd` file into the `posts` folder **BEFORE** you render it, so that all files are stored in the correct location.
:::

:::{.callout-important}
Only render a single file - don't try to render the whole website!
:::

:::{.callout-note}
## Pilot Student Blogs

We are piloting a workflow including individual student websites with direted and limited pull requests back to course blogs. Please let us know if you would like to participate.
:::

## Reading in data files

The easiest data source to use - at least initially - is to choose something easily accessible, either from our `_data` folder provided, or from an online source that is publicly available.

:::{.callout-tip}
## Using Other Data
If you would like to use a source that you have access to and it is small enough and you don't mind making it public, you can copy it into the `_data` file and include in your *commit* and *pull request*.
:::

:::{.callout-tip}
## Using Private Data

If you would like to use a proprietary source of data, that should be possible using the same process outlined above. There may initially be a few issues. We hope to have this feature working smoothly soon!
:::-->