---
title: "Challenge 4"
author: "Priyanka Perumalla"
description: "More data wrangling: pivoting"
date: "05/15/2023"
format:
  html:
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
  - challenge_4
  - Priyanka Perumalla
  - hotel_bookings
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Challenge Overview

Today's challenge is to:

1)  read in a data set, and describe the data set using both words and any supporting information (e.g., tables, etc)
2)  tidy data (as needed, including sanity checks)
3)  identify variables that need to be mutated
4)  mutate variables and sanity check all mutations

## Read in data

Read in one (or more) of the following datasets, using the correct R package and command.

-   abc_poll.csv ⭐
-   poultry_tidy.xlsx or organiceggpoultry.xls⭐⭐
-   FedFundsRate.csv⭐⭐⭐
-   hotel_bookings.csv⭐⭐⭐⭐
-   debt_in_trillions.xlsx ⭐⭐⭐⭐⭐

```{r}

hotel_df <- read.csv('_data/hotel_bookings.csv')
head(hotel_df)
```

### Briefly describe the data

```{r}
dim(hotel_df)
```

```{r}
colnames(hotel_df)
```

```{r}
str(hotel_df)
```


```{r}
#| label: summary
library(summarytools)
dfSummary(hotel_df)
```

The data contains information of hotel bookings in different hotels. Every booking entry is a row having information on which hotel the booking corresponds to, when it happened, what is the start date and what is the end date, who made the booking, for who all the booking was made, the channel used for booking, the number of days for which it was made, how much money was paid, all the preferences that were entered along with the booking etc. 


## Tidy Data (as needed)

Is your data already tidy, or is there work to be done? Be sure to anticipate your end result to provide a sanity check, and document your work here.

```{r}
sum(is.na(hotel_df$adr))
```

```{r}
sum(is.na(hotel_df$arrival_date_month))
```

```{r}
sum(is.na(hotel_df$arrival_date_year))
```

```{r}
sum(is.na(hotel_df$`hotel`))
```

```{r}
sum(is.na(hotel_df$`booking_changes`))
```

```{r}
sum(is.na(hotel_df$`children`))
```

```{r}
sum(is.na(hotel_df$`meal`))
```

```{r}
sum(is.na(hotel_df$`agent`))
```

```{r}
sum(is.na(hotel_df$`stays_in_weekend_nights`))
```

Surprisingly, the data looks very clean. There are almost no null values or missing entries. The 'children' column seems to have 4 NA values but unless my analysis revolves around this column, I prefer not to delete the date as the rest of the columns for these rows has other valuable information. 


## Identify variables that need to be mutated

Are there any variables that require mutation to be usable in your analysis stream? For example, are all time variables correctly coded as dates? Are all string variables reduced and cleaned to sensible categories? Do you need to turn any variables into factors and reorder for ease of graphics and visualization?

Document your work here.


Most databases have a single date time stamp instead of seperate columns for date, month, year and time. To allow such analysis I would like to combine all date related columns into one. ( Year, Month, and Day).In general any hotel data would have the number of total days in a week the hotel is busy instead of having weekdays and weekend days separately (unless interested in such an analysis). So I want to create a new column for it combining both.

Next, we will analyze the data and attempt to identify the correlation between Unemployment and Inflation rates based on date.

```{r}
hotel_df$arrival_datestamp <- ymd(paste(hotel_df$arrival_date_year, hotel_df$arrival_date_month, hotel_df$arrival_date_day_of_month,  sep = "-"))
summary(hotel_df$arrival_datestamp)
```
```{r}
hotel_df$total_occupied_days <- sum(hotel_df$stays_in_week_nights, hotel_df$stays_in_weekend_nights)
summary(hotel_df$total_occupied_days)
```


Next, we will analyze the data and attempt to group data based on arrival date year to see how data looks
```{r}
summarydf <- hotel_df %>%
group_by(arrival_date_year) %>%
print(summarydf)
```
Analyzing how data varies month wise by hotel for the variable stays in weekend nights.
```{r}
monthVsWeekendbookingVsHotel <- hotel_df  %>% select(arrival_date_month,stays_in_weekend_nights, hotel)
monthVsWeekendbookingVsHotel
```
