---
title: "Challenge 6"
author: "Priyanka Perumalla"
description: "Visualizing Time and Relationships"
date: "05/15/2023"
format:
  html:
    toc: true
    code-copy: true
    code-tools: true
categories:
  - challenge_6
  - Priyanka Perumalla
  - hotel_bookings
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)
library(ggplot2)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Challenge Overview

Today's challenge is to:

1)  read in a data set, and describe the data set using both words and any supporting information (e.g., tables, etc)
2)  tidy data (as needed, including sanity checks)
3)  mutate variables as needed (including sanity checks)
4)  create at least one graph including time (evolution)
   - try to make them "publication" ready (optional)
   - Explain why you choose the specific graph type
5)  Create at least one graph depicting part-whole or flow relationships
   - try to make them "publication" ready (optional)
   - Explain why you choose the specific graph type

[R Graph Gallery](https://r-graph-gallery.com/) is a good starting point for thinking about what information is conveyed in standard graph types, and includes example R code.

(be sure to only include the category tags for the data you use!)

## Read in data

Showing first 5 columns in the data
```{r}
df <- read.csv("_data/hotel_bookings.csv")
head(df, n = 5)[,1:5]
summary(df)
```

### Briefly describe the data
The data provided has information about hotel bookings. There are a total of 119,390 entries in the dataset. The dataset includes information such as the type of hotel, whether the booking was canceled, lead time, arrival year, month and day, number of nights stayed, the number of adults, children and babies, meal type, country of origin, market segment, distribution channel, previous cancellations, reserved and assigned room type, booking changes, deposit type, days on the waiting list, customer type, average daily rate, required car parking spaces, and total number of special requests.

No of countries present in the data:
```{r}
unique_country <- unique(df$country)
cat(paste(unique_country, collapse = ", "))
```

## Tidy Data (as needed)
Creating a new column for total no of bookings combining all the age group booking to visualize no of bookings happened in 2015 for 5 countries: USA, BEL, GBR, DEU, FRA over the span of every month bookings. It helps to see on which seasons or depending on month the no of bookings are high in a country. Aggregating month and year to form the new date.
```{r}
# create new date column adding all the bookings
df$total_bookings<-rowSums(df[, c("adults", "children","babies")], na.rm = TRUE)
df$date <- as.Date(paste(df$arrival_date_year, df$arrival_date_month, df$arrival_date_day_of_month, sep = "-"), format = "%Y-%B-%d")
head(df)
df1<-df
```

## Time Dependent Visualization
```{r}
#summing all the bookings of a month w.r.t country
df <- df %>% filter(arrival_date_year == 2015)%>% group_by(arrival_date_month, country) %>% summarise(monthly_bookings =sum(total_bookings))

df<-df%>% filter(country %in% c("USA", "BEL", "GBR", "DEU", "FRA"))
ggplot(df, aes(x = arrival_date_month, y = monthly_bookings, fill = country)) + 
  geom_bar(stat = "identity", position = "dodge") + 
  labs(x = "Month of Booking", y = "Total Bookings", title = paste("Total Bookings in a country in 2015")) + 
  theme_bw()+
  scale_fill_manual(values = c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2"))
```

## Visualizing Part-Whole Relationships
```{r}
ggplot(df, aes(x = arrival_date_month, y = monthly_bookings, fill = country)) +
  geom_col() +
  labs(title = "Visualizing Part-Whole Relationships Country and Month",
       y = "Monthly Bookings",
       x = "Month") +
  theme_bw()+
  scale_fill_manual(values = c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2"))
```

