---
title: "Challenge 7_PriyankaThatikonda"
author: "Meredith Rolfe"
description: "Visualizing Multiple Dimensions"
date: "08/24/2022"
format:
  html:
    toc: true
    code-copy: true
    code-tools: true
categories:
  - challenge_7
  - hotel_bookings
  - australian_marriage
  - air_bnb
  - eggs
  - abc_poll
  - faostat
  - usa_households
editor: 
  markdown: 
    wrap: 72
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)
library(ggplot2)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Challenge Overview

Today's challenge is to:

1)  read in a data set, and describe the data set using both words and
    any supporting information (e.g., tables, etc)
2)  tidy data (as needed, including sanity checks)
3)  mutate variables as needed (including sanity checks)
4)  Recreate at least two graphs from previous exercises, but introduce
    at least one additional dimension that you omitted before using
    ggplot functionality (color, shape, line, facet, etc) The goal is
    not to create unneeded [chart ink
    (Tufte)](https://www.edwardtufte.com/tufte/), but to concisely
    capture variation in additional dimensions that were collapsed in
    your earlier 2 or 3 dimensional graphs.

-   Explain why you choose the specific graph type

5)  If you haven't tried in previous weeks, work this week to make your
    graphs "publication" ready with titles, captions, and pretty axis
    labels and other viewer-friendly features

[R Graph Gallery](https://r-graph-gallery.com/) is a good starting point
for thinking about what information is conveyed in standard graph types,
and includes example R code. And anyone not familiar with Edward Tufte
should check out his [fantastic
books](https://www.edwardtufte.com/tufte/books_vdqi) and [courses on
data visualizaton.](https://www.edwardtufte.com/tufte/courses)

(be sure to only include the category tags for the data you use!)

## Read in data

Read in one (or more) of the following datasets, using the correct R
package and command.

-   eggs ⭐
-   abc_poll ⭐⭐
-   australian_marriage ⭐⭐
-   hotel_bookings ⭐⭐⭐
-   air_bnb ⭐⭐⭐
-   us_hh ⭐⭐⭐⭐
-   faostat ⭐⭐⭐⭐⭐

```{r}
 data <- read.csv("_data/AB_NYC_2019.csv")
 head(data)
 dim(data)
 summary(data)
```

### Briefly describe the data

The dataset contains information about approximately 49,000 rental properties that were listed on Airbnb in New York City during the year 2019. Each entry represents a unique rental unit and includes 16 variables that offer a range of details. These details include the unit's ID, name, location, host ID and name, room type, price, minimum nights required for booking, number of reviews, date of the last review, average monthly reviews, calculated count of host listings on Airbnb, and availability.

## Tidy Data (as needed)

Is your data already tidy, or is there work to be done? Be sure to anticipate your end result to provide a sanity check, and document your work here.

```{r}
data <- data %>% 
  na.omit() %>% 
  mutate(high_price = ifelse(price > 500, "High", "Low"))
print(data)
```

Are there any variables that require mutation to be usable in your
analysis stream? For example, do you need to calculate new values in
order to graph them? Can string values be represented numerically? Do
you need to turn any variables into factors and reorder for ease of
graphics and visualization?

Document your work here.

```{r}

```

## Visualization with Multiple Dimensions

```{r}
ggplot(data, aes(x = price)) +
  geom_histogram(binwidth = 40, color = "black", fill = "black") +
  labs(title = "Price Distribution", x = "Price", y = "Frequency") +
  theme_minimal()

ggplot(data, aes(x = price, y = reviews_per_month)) +
  geom_point(alpha = 0.46, color = "pink") +
  labs(title = "Monthly Price vs Reviews", x = "Price", y = "Reviews per Month") +
  theme_minimal()


```
