---
title: "Challenge 1"
author: "Priyanka Perumalla"
description: "Getting acquainted with the properties of the dataset"
date: "05/15/2023"
format:
  html:
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
  - challenge_1
  - railroads
  - faostat
  - wildbirds
---

```{r}
#| label: setup
#| warning: false

library(tidyverse)

knitr::opts_chunk$set(echo = TRUE, warning =FALSE, message=FALSE)
```

## Challenge Overview

Today's challenge is to

1)  read in a dataset, and

2)  describe the dataset using both words and any supporting information (e.g., tables, etc)

## Read in the Data

Read in one (or more) of the following data sets, using the correct R package and command.

-   railroad_2012_clean_county.csv ⭐
-   birds.csv ⭐⭐
-   FAOstat\*.csv ⭐⭐
-   wild_bird_data.xlsx ⭐⭐⭐
-   StateCounty2012.xls ⭐⭐⭐⭐

Find the `_data` folder, located inside the `posts` folder. Then you can read in the data, using either one of the `readr` standard tidy read commands, or a specialized package such as `readxl`.

The birds.csv file is read in using read_csv().

```{r}
df <- read_csv('/Users/priyankaperumalla/Desktop/daccs/601_Spring_2023/posts/_data/FAOSTAT_egg_chicken.csv', show_col_types = FALSE)
head(df)
```

## Describe the data

Using a combination of words and results of R commands, can you provide a high level description of the data? Describe as efficiently as possible where/how the data was (likely) gathered, indicate the cases and variables (both the interpretation and any details you deem useful to the reader to fully understand your chosen data).

Description : The dataset 'FAOSTAT_egg_chicken.csv' contains information about the livestock data produced/consumed by countries during various timelines (number of years) ranging from 1961 to 2018.


```{r}
#| label: summary
head(df)
```

Displaying the summary of the dataset.

```{r}
summary(df)
```

Displaying the dimensions of the dataset,

```{r}
dim(df)
```

Printing all columns

```{r}
colnames(df)
```

Printing the unique Years

```{r}
unique_years <- df%>% select(Year)%>% n_distinct(.)
unique_years
```

The dataset contains info of 58 unique years i.e 1961-2018

Printing the unique areas
```{r}
unique_areas <- df%>% select(Area)%>% n_distinct(.)
unique_areas
```

The dataset contains info about 245 unique areas

Printing the unique Domains (eg; Primary Live Stock etc) 
The data is all from a single domain
```{r}
unique_areas <- df%>% select(Domain)%>% n_distinct(.)
unique_areas
```