---
title: "Eggs Sold Graph Visualizations"
author: "Sahan Prasad Podduturi Reddy"
description: "Visualizing Multiple Dimensions"
date: "05/11/2023"
format:
  html:
    toc: true
    code-copy: true
    code-tools: true
categories:
  - challenge_7
  - eggs
  - Sahan Prasad Podduturi Reddy
---

## Introduction

I was trying to analyze the 'hotel_bookings.csv' dataset. This dataset contains information about quantity of eggs sold between 2004-2013. It lists 4 different package types - large_half_dozen, extra_large_half_dozen, large_dozen, extra_large_dozen . We first start by importing the necessary libraries and setting the working directory to point to the location where the spreadsheet is located. Then we read in the csv file.

```{r}
#| label: Setup
#| warning: false
#| message: false

library(tidyverse)
library(ggplot2)
library(readxl)
library(lubridate)
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
setwd("D:/MyDocs/Class Slides/DACSS601/601_Spring_2023/posts/_data")
eggs <- read.csv("eggs_tidy.csv")
head(eggs)
```

## Read and Tidy Data

We first read in the dataset into our dataframe and sort months by chronological order. Then we pivot the data such that package types are all in a single column while the Quantities are in a separate column. This makes it easier for us to meaningfully visualize our data. We also append a month_number column to our dataframe so that we can plot line graphs because they need continuous data.

```{r}
#| label: ReadFile
#| warning: false
#| message: false

month_order <- c("January", "February", "March", "April", "May", "June", "July", "August", "September", "October", "November", "December")
eggs$month <- factor(eggs$month, levels = month_order)
eggs_one <- eggs[order(eggs$month),] %>% group_by(month) %>% summarise("large_half_dozen" = sum(large_half_dozen), "large_dozen" = sum(large_dozen), "extra_large_half_dozen" = sum(extra_large_half_dozen), "extra_large_dozen" = sum(extra_large_dozen)) %>% pivot_longer(cols = c(large_half_dozen, extra_large_half_dozen, large_dozen, extra_large_dozen), names_to = "size", values_to = "quantity") %>% ungroup() %>% mutate(month_number = match(month, month.name))
head(eggs_one)
```

## Visualization with Multiple Dimensions

We first plot a grouped bar chart which indicates the quantities sold per month over all year values in our dataframe. We use the position="dodge" attribute to plot bars side-by-side so that we can directly compare quantities sold for different packages types. We notice that the quantity sold for each package type doesn't fluctuate much over the months of the year. We also plot a geom_point() graph which confirms our observations. "Dozen" size packages are much more in demand than "Half-Dozen" size packages. There is a slight increase in egg sales around June and then the sales remain the same till the end of the Year. We also do another visualization of the Quantity of Eggs sold over the Year by plotting graphs for different years in the dataset. On doing so, we notice how the current rate of Eggs sold was reached through the years.

```{r}
#| label: Visualization
#| warning: false
#| message: false

eggs_one %>% 
    ggplot(aes(y=`quantity`, x=month, fill=size)) +
    geom_bar(position="dodge", stat="identity") +
    labs(title = "Quantity of Eggs Sold 2004-2013", x = "Month", y = "Quantity")

eggs_one %>% 
    ggplot(aes(y=`quantity`, x=month, color=size, shape=size)) +
    geom_point() +
    theme_bw()
    labs(title = "Quantity of Eggs Sold 2004-2013", x = "Month", y = "Quantity")

eggs_two <- eggs[order(eggs$month),] %>% group_by(month, year) %>% mutate(total = `large_half_dozen`+`large_dozen`+`extra_large_half_dozen`+`extra_large_dozen`) %>% mutate(month_number = match(month, month.name)) %>% select(month, year, month_number, total)
  eggs_two
  
  eggs_two %>% 
    ggplot(aes(y=`total`, x=month_number)) +
    geom_line() +
    facet_wrap(vars(year), scales = "free")
    labs(title = "Quantity of Eggs Sold 2004-2013", x = "Month", y = "Quantity")
```