---
title: "Challenge 1"
author: "Sai Pranav Kurly"
desription: "Reading in data and creating a post"
date: "03/01/2023"
format:
  html:
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
  - challenge_1
  - railroads
  - faostat
  - wildbirds
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Challenge Overview

Today's challenge is to

1)  read in a dataset, and

2)  describe the dataset using both words and any supporting information (e.g., tables, etc)

## Read in the Data

We can read the railroad_2012_clean_county.csv dataset using the read_csv function. Below also shows the first few rows and the summary of the dataset.

```{r}
#Read the dataset
dataframe <- read_csv("_data/railroad_2012_clean_county.csv")

#Display the first few rows of the dataset
head(dataframe)

#Summarise the dataset
dataframe %>%
  summary(dataframe)
```

## Describe the data

We see that the dataset is that of the employees in the railroad across the United States.There are 3 columns STATE, COUNTY, and TOTAL EMPLOYEES and a total of 2930 rows. 

```{r}
#Grouping the data by state in descending order of employees
dataframe%>%
  group_by(state)%>%
  summarise(total=sum(total_employees))%>%
  arrange(desc(total)) %>%
slice(1:10)
```
Above shows the top 10 states where the most employees are present. We see from above that Texas has the highest number of employees. 

```{r}
#Filter state AE from the dataset 
filter(dataframe, state == "TX")
```
 Above shows us a filtered veiw of TX.

```{r}
#Grouping the data by state in ascending order of employees
dataframe%>%
  group_by(state)%>%
  summarise(total=sum(total_employees))%>%
  arrange(total) %>%
slice(1:10)
```

We see from above that AP has the lowest number of employees.  