---
title: "Challenge 1"
author: "Sai Venkatesh"
description: "Reading in data and creating a post"
date: "04/12/2023"
format:
  html:
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
- challenge_1
- railroads
- faostat
- wildbirds
---
  
```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Challenge Overview

Today's challenge is to

1)  read in a dataset, and

2)  describe the dataset using both words and any supporting information (e.g., tables, etc)

## Read in the Data

We are going to load the railroad data.

```{r}


railroad <- read.csv('_data/railroad_2012_clean_county.csv')

print("Lets load the data and see the dimensions and columns of the data.")

# The Dimensions 
dim(railroad)

# The Column Names 
colnames(railroad)

```

From the above, we can see that the Railroad data has 2930 rows and 3 columns.
The 3 column names are state, county and total_employees.


```{r}

print("The top rows of the data are :- ")
head(railroad)
```


## Describe the data

Using a combination of words and results of R commands, can you provide a high level description of the data? Describe as efficiently as possible where/how the data was (likely) gathered, indicate the cases and variables (both the interpretation and any details you deem useful to the reader to fully understand your chosen data).

The data seems to represent the railroad employees and the distribution of the employees by state and county.


```{r}

# Number of Employees Per State
print("The total number of employees in the states ordered by the count:-")
railroad %>%
  group_by(state) %>%
  summarize(state_total = sum(total_employees))  %>%
  select(state, state_total)  %>%
  arrange(desc(state_total))

```


```{r}

# Number of Employees > 1000
print("The counties with employees greater than 1000 ordered by the count:-")
railroad %>%
  filter(total_employees > 1000)  %>%
  select(state, county, total_employees) %>%
  arrange(desc(total_employees))

```

We can see that Texas state has the most employees and Cook County has the most employees.