---
title: "Exploring the Birds dataset"
author: "Saksham Kumar"
description: "Reading in data and creating a post"
date: "03/20/2023"
format:
  html:
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
  - challenge_1
  - birds
  - Saksham Kumar
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Challenge Overview

Today's challenge is to

1)  read in a dataset, and

2)  describe the dataset using both words and any supporting information (e.g., tables, etc)

## Read in the Data

Read in one (or more) of the following data sets, using the correct R package and command.

-   railroad_2012_clean_county.csv ⭐
-   birds.csv ⭐⭐
-   FAOstat\*.csv ⭐⭐
-   wild_bird_data.xlsx ⭐⭐⭐
-   StateCounty2012.xls ⭐⭐⭐⭐

Find the `_data` folder, located inside the `posts` folder. Then you can read in the data, using either one of the `readr` standard tidy read commands, or a specialized package such as `readxl`.

```{r}
birds <- read_csv('_data/birds.csv', show_col_types = FALSE)
```

```{r}
birds
```

Add any comments or documentation as needed. More challenging data sets may require additional code chunks and documentation.

## Describe the data

Using a combination of words and results of R commands, can you provide a high level description of the data? Describe as efficiently as possible where/how the data was (likely) gathered, indicate the cases and variables (both the interpretation and any details you deem useful to the reader to fully understand your chosen data).

```{r}
dim(birds)
```

The data has 30977 rows and 14 columns (variables) 

```{r}
colnames(birds)
```
The columns (variables) in our data are listed above.

```{r}
#| label: summary
spec(birds)
```
This helps us understand the datatype of the variables we have.

```{r}
birds%>%select(Item)%>%n_distinct()
birds%>%select(Item)%>%unique()
```

The dataset contains 5 unique values in the variable 'Items'. These seem to poultry birds from a first glance.

```{r}
birds%>%select(Area)%>%n_distinct()
birds%>%select(Area)%>%unique()
```
The dataset contains 248 unique values in the variable 'Area' i.e. it contains data from across 248 countries

```{r}
library(ggplot2)
ggplot(data = birds, aes(x = Item)) + 
  geom_bar()
```

This graph helps us understand the distribution of birds in our dataset. As we can see, chicken is the most popular bird in our dataset.