---
title: "Railroad Employees Challenge 1"
author: "Sue-Ellen Duffy"
desription: "Railroad Employees"
date: "02/23/2023"
format:
  html:
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
  - challenge_1
  - Sue-Ellen Duffy
  - Railroad Employee Dataset
---

```{r}
#| label: setup
#| warning: false

library(tidyverse)
library(readxl)
library(summarytools)

knitr::opts_chunk$set(echo = TRUE)
```

## Reading in the Data

I analyzed the "railroad_2012_county_clean.csv" data for Challenge 1. This data describes the Total Number of Railroad Employees by County and State in the United States in 2012. Upon first glance the data contains 3 columns and 2,930 rows. The columns are: state, county, and total_employees

```{r}
#Read in data and rename railroad_2012_clean_county as data
data <- rename(read_csv("_data/railroad_2012_clean_county.csv"))
#Preview data 
data
```

## Summary of Data

Running the dfsummary(data) function shows us:

-   The data is complete: there are no missing data.

-   Top ten states ranked with the most counties. Texas has the most counties of any other state, accounting for 7.5% of all counties in the United States.

-   There are multiples of county names. We see in the following graph the top 10 county names that are used in the United States. (There are 31 Washington county names in this data plot, that's far more than I thought there were in the United States!)

```{r}
dfSummary(data)
```

```{r}
#How many states are represented in the data?
data %>%
  select(state) %>%
  n_distinct(.)
```

There are only 50 recognized states, so we need to dig a little deeper to find out what the three additional 'states' represent.

```{r}
#Show unique state data
unique(data$state)
```

AE, AP, and DC are the three non-states cases. AE and AP are military addresses. DC is Washington DC.
