---
title: "Challenge 4"
author: "Shantanu Patil"
description: "More data wrangling: pivoting"
date: "03/26/2023"
format:
  html:
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
  - challenge_4
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)
library(lubridate)
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Challenge Overview

Today's challenge is to:

1)  read in a data set, and describe the data set using both words and any supporting information (e.g., tables, etc)
2)  tidy data (as needed, including sanity checks)
3)  identify variables that need to be mutated
4)  mutate variables and sanity check all mutations

## Read in data



```{r}
## Read in data
poul <- read_csv("_data/poultry_tidy.csv")

## Sample a few data points
poul_sample <- poul[sample(nrow(poul), 10), ]
print(poul_sample)

## Calculate product counts
product_counts <- table(select(poul, Product))
print(product_counts)

```

### Briefly describe the data

Given the Product, Year, and Month, this dataset describes the price of poultry meat. We can see that there are five different types of meat.

## Tidy Data (as needed)
The data already tidy.


## Identify variables that need to be mutated

The 'Product' and 'Month' categories must be converted to <dbl> (double) data types so that we can perform numerical operations on them more easily. For example, converting the Month category to a numeric data type would allow us to perform month-to-month comparisons while accounting for the new year's rollover.   We should also add a 'Date' column based on the 'Month' and'Year' columns. This will simplify data visualization and analysis.

```{r}
# Assign product IDs
poul_mut <- poul %>%
  mutate(Product_ID = recode(Product, "B/S Breast" = 1, 
                             "Bone-in Breast" = 2, 
                             "Thighs" = 3, 
                             "Whole" = 4, 
                             "Whole Legs" = 5))

# Convert month names to month numbers
poul_mut <- poul_mut %>%
  mutate(Month_num = recode(Month, "January" = 1, "February" = 2, "March" = 3, 
                            "April" = 4, "May" = 5, "June" = 6, "July" = 7, 
                            "August" = 8, "September" = 9, "October" = 10, 
                            "November" = 11, "December" = 12))

# Create Date column based on Year and Month_num
poul_mut <- poul_mut %>%
  mutate(Date = make_date(Year, Month_num))

# Display a sample of the modified data
poul_mut[sample(nrow(poul_mut), 10), ]
```

