---
title: "Challenge 7"
author: "Shantanu Patil"
description: "Visualizing Multiple Dimensions"
date: "05/13/2023"
format:
  html:
    toc: true
    code-copy: true
    code-tools: true
categories:
  - challenge_7
  - eggs

---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)
library(ggplot2)
library(lubridate)


knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```


## Read in data


```{r}
eggs_data <- read_csv('_data/eggs_tidy.csv')
head(eggs_data)
```

### Briefly describe the data
Given the month and year, this dataset describes the average price for a carton of eggs. Carton sizes are as follows: big half dozen, large dozen, extra large half dozen, and extra large dozen.

## Tidy Data (as needed)

I'll rearrange the dataset utilizing pivot functionality to aid in visualization. Following that, divide the egg carton type into two columns: one for the size and one for the number (half or full dozen). This allows us to plot each attribute independently. The "between extra_large and half_dozen will be removed to maintain only one" in the name, separating size and quantity. These will be divided into two columns. Furthermore, lubridate will be used to transform the date into the right format.

```{r}
eggs <- eggs_data %>%
    pivot_longer(cols = c(`large_half_dozen`, `large_dozen`, `extra_large_half_dozen`, `extra_large_dozen`), values_to = "Price ($)") %>%
    mutate(name = str_replace(name, "extra_large", "Extra Large"), name = str_replace(name, "half_dozen", "Half Dozen"), name = str_replace(name, "dozen", "Dozen"), name = str_replace(name, "large", "Large")) %>%
    separate(name, into = c("Size", "Quantity"), sep = "_") %>%
    mutate(Date = ym(paste(`year`, `month`, sep = "-")))
head(eggs)



```



## Visualization with Multiple Dimensions


```{r}
ggplot(eggs, aes(Date, `Price ($)`, col = Size)) +
    ggtitle('Cost Evolution for Different Types of Egg Cartons') +
    geom_line() +
    facet_wrap(~Quantity, scales = 'free_y')


```
The above graph type was chosen to illustrate the price variation for extra large and large eggs over the course of time. I divided the graph based on the carton size since the prices vary accordingly. The graph reveals a decrease in prices for large eggs sold by the dozen and an increase for those sold by the half dozen.

```{r}
ggplot(eggs, aes(Date, `Price ($)`, fill = Quantity)) +
    ggtitle('Price Variations Based on Egg Size') +
    geom_bar(position = "stack", stat = "identity") +
    facet_wrap(~Size)



```
In the second graph, I replaced the grouping variable with the egg size and then stacked the prices of both quantities. This visual representation allows us to observe the cumulative price changes depending on the size of the eggs. It's evident from this visualization that large eggs are consistently cheaper than extra large ones and that both egg sizes follow a similar trend over time.

