---
title: "Basic R commands and exploring Iris dataset"
author: "Shantanu Patil"
desription: "Exploring and Analysing Iris dataset"
date: "02/23/2022"
format:
  html:
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
  - week1 post
  - hw1
  - shantanu patil
  - dataset
---


```{r}
#| label: setup
#| warning: false

library(tidyverse)
library(readr)
knitr::opts_chunk$set(echo = TRUE)
```

## Testing c function for integers and strings and seq functions


```{r}
x <- c(2,3,4,5,8, 6, 7, 5, 3, 0, 9)
mean(x)
names = c("Om", "Damian", "Sunny", "Oguz", "Jake", "Jain")
names
y = seq(1, 14, 1.5)
y


```
## Reading the iris dataset

The iris dataset is a built-in dataset in R. It can be loaded using the command data(iris)


## Exploring the iris data 
I decided to explore the iris data for my first post of week 1.

Iris is a flowering plant, and the researchers have measured and digitally documented the varied characteristics of the many iris blossoms.

The Iris data set has 150 rows and 5 colums.There are 3 Species of Iris plant present in the dataset.


The Colums are the various dimensions of the petals which are Sepal.Length, Sepal.Width, Petal.Length, Petal.Width,  Species.



```{r}

data(iris)

names(iris)

dim(iris)

str(iris)

```

## Summary of iris data 
Using the summary function in R gives us the various statistics of the values of each column in the dataset.


```{r}

summary(iris)

```







