---
title: "Challenge 5 "
author: "Shantanu Patil"
description: "Introduction to Visualization"
date: "03/26/2023"
format:
  html:
    toc: true
    code-copy: true
    code-tools: true
categories:
  - challenge_5
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)
library(ggplot2)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Read in data

```{r}
# reading the ceral dataset 
cereal <- read_csv("_data/cereal.csv")
head(cereal,4)
```

### Briefly describe the data
This data set contains four columns:

* `Cereal <chr>`: Cereal Name
* `Sodium <dbl>`: Amount of sodium per serving of the cereal
* `Sugar <dbl>`: Amount of sugar per serving of cereal
* `Type <chr>`: Cereal Type (Child or Adult)

## Tidy Data (as needed)
Just making a few changes to the dataset.

```{r}
cereal <- cereal %>%
  mutate(Type = recode(Type, "A" = "Adult", "C" = "Child"))

# Display first 5 rows of the modified data
head(cereal, 5)

```

## Univariate Visualizations
```{r}
# Create a histogram for the distribution of Sodium based on Type
sodium_plot <- ggplot(cereal, aes(Sodium)) +
  geom_histogram(binwidth = 100) +
  facet_grid(vars(Type)) +
  labs(title = "Distribution of Sodium based on Type")

# Display the Sodium histogram
print(sodium_plot)

# Create a histogram for the distribution of Sugar based on Type
sugar_plot <- ggplot(cereal, aes(Sugar)) +
  geom_histogram(binwidth = 5) +
  facet_grid(vars(Type)) +
  labs(title = "Distribution of Sugar based on Type")

# Display the Sugar histogram
print(sugar_plot)

```
Because we are dealing with univariate, continuous data, I chose histograms. I wanted to compare things by type, so I used a facet grid.

## Bivariate Visualization(s)
```{r}
# Create a boxplot for the amount of Sugar in Cereal by Type
sugar_by_type_plot <- ggplot(cereal, aes(Type, Sugar)) +
  geom_boxplot() +
  labs(title = "Amount of Sugar in Cereal by Type")

# Display the boxplot
print(sugar_by_type_plot)

```
We can see that, on average, children's cereal contains more sugar than adult cereal. Because the data is so sparse, I went with a box plot.
