---
title: "Exploring and Analysing the Birds Dataset"
author: "Shantanu Patil"
desription: "Reading in data and creating a post"
date: "02/23/2023"
format:
  html:
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
  - challenge_1
  - birds
  - hw1
  - wildbirds
  - shantanu patil
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Challenge Overview

Today's challenge is to

1)  read in a dataset, and

2)  describe the dataset using both words and any supporting information (e.g., tables, etc)

## Reading in the bird data

I loaded the library readr using command library(readr)

The birds.csv file has 30977 Rows and 14 Columns.
I have used head function to load the column headers and first 3 rows.


```{r}

library(readr)
birds_data <- read.csv(file = "_data/birds.csv")

head(birds_data, 3)

# a data set created with only numeric values skip header
bird_data2 <- read_csv(file = "_data/birds.csv", skip=1)
head(bird_data2, 1)


```
## Describe the data
We can see that the bird data is made of 14 columns of which 8 are of character data type and the remaining are of int data type.
To find out what are the column names we can use colnames function.
The data gathered has information about Domain, Area, Element, Item, Year, Unit, Value, Flag, Flag.Description.
```{r}

str(birds_data)

colnames(birds_data)
```


## Finding Out the start and end date from when the data was collected.
We can see that the data was collected from 1961 to 2018.

```{r}
max(birds_data$Year)
min(birds_data$Year)
```





