---
title: "Challenge 2: Finding the Cattle Dairy Value statistics for each country from 1961 - 2018"
author: "Shantanu Patil"
description: "Data wrangling and Data Sumarization to find statistics in data"
date: "03/01/2023"
format:
  html:
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
  - challenge_2
  - faostat
  - faostat_cattle_diary
  - Shantanu Patil
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Read in the Data
I have selected faostat_cattle_diary dataset to do my analysis on.

```{r}
# Importing the faostat_cattle_diary dataset and reading them
library(readr)
data <-read_csv("_data/FAOSTAT_cattle_dairy.csv")
#Lets see the first 3 rows of the dataset
head(data, 3)
```

## Describe the data

Using R commands I am exploring the data and coming up with the description from the data.
The dataset contains 14 column, 8 of which have character datatype and 6 have integer data type.
Here are the Columns

chr (8): Domain Code, Domain, Area, Element, Item, Unit, Flag, Flag Description

dbl (6): Area Code, Element Code, Item Code, Year Code, Year, Value

Here is the size of the Dataset ---  Rows: 36449 Columns: 14  


### Checking the variation in data for each column

```{r}
#| label: summary
# Check the variation in data for each column
summary(data)

```

## Provide Grouped Summary Statistics

Conducting  exploratory data analysis, using dplyr commands such as `group_by()`, `select()`, `filter()`, and `summarise()`.

Remove columns with "Item Code" values as it has constant value 882

```{r}
#install.packages("dplyr")
library(dplyr)
#remove columns with "Item Code" values as it has contant value 882
data<-data%>%
  select(-contains("Item Code"))

```

Grouping by the Area and finding the mean_value
```{r}
data %>% 
  group_by(Area) %>% 
  summarise(mean_value = mean(Value))
```

Table output by selecting Columns "Area", "Year", and "Value".
```{r}
data %>% 
  select(Area, Year, Value)
```

I wanted to check out how any data rows / points are available for Antigua and Barbuda in the year of 1975. There are 3 datapoints available.
```{r}
data %>% 
  filter(Area == "Antigua and Barbuda" , Year == 1975)
```


## Table Value Statistics
Find the central tendency (mean, median, mode) and dispersion (standard deviation, mix/max/quantile) for Value within the data set.
```{r}
#deleting the rows with Null values
data2 <- na.omit(data)
# Find the central tendency (mean, median, mode) and dispersion (standard deviation, mix/max/quantile) for different subgroups within the data set.
data2%>%
  group_by(Area)%>%
  summarize(mean.Value = mean(Value, na.rm = TRUE), median.Value = median(Value, na.rm = TRUE), min.Value = min(Value, na.rm = TRUE), max.Value = max(Value, na.rm = TRUE), sd.Value = sd(Value, na.rm = TRUE), var.Value = var(Value, na.rm = TRUE), IQR.Value = IQR(Value, na.rm = TRUE))

```
I wanted to check out how any data rows / points are available for Antigua and Barbuda in the year of 1975. There are 3 datapoints available.
Some Conclusion from seeing the data summary

1. The Data is Collected from the year 1961 - 2018
2. The Item Code column just has one value 882
3. The Area Code varies from 1 - 5504
4. The Element Code varies frojm 5318 - 5510
5. The Value Column has a lot of variance

I deleted the Item Code Column.

I choose Area (Country) as a group because I wanted to see how does the Value of output of cattle and diary farms vary based on teh country they are located in. I think every country has a set of climatic condition, thus the value in tropical countries might be far greater than countries near north pole, I wanted to test this hypothesis.

I found out the relationship is not linear, on seeing the data I realized that some countries might be the around the same  distance from equator or have same climatic conditions but still the mean value out put varies. Ex Afghanistan and Africa may have somewhat same climatic conditions but as afganisatn is located i the mountains the mean Value is much lower compared to Africa.


