---
title: "Challenge 8"
author: "Shaunak Padhye"
description: "Joining Data"
date: "05/20/2023"
format:
  html:
    toc: true
    code-copy: true
    code-tools: true
categories:
  - challenge_8
  - Shaunak Padhye
  - snl
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)
library(ggplot2)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Read in data

For this challenge I will be using the following dataset.

  - snl ⭐⭐⭐⭐⭐

```{r}
snl_casts <- read_csv("_data/snl_casts.csv", skip = 1, col_names = c("aid", "sid", "featured", "cast_first_epid", "cast_last_epid", "update_anchor", "cast_n_epid", "season_fraction"))
  
snl_actors <- read_csv("_data/snl_actors.csv")
  
snl_seasons <- read_csv("_data/snl_seasons.csv", skip = 1, col_names = c("sid", "year", "seasons_first_epid", "seasons_last_epid", "seasons_n_epid"))
```

### Briefly describe the data

This dataset contains three different dataframes:

snl_actors: The actors dataframe contains details of the cast of people who have appeared in SNL from 1975 to 2020.

snl_casts: The casts dataframe contains details of the cast for various seasons of SNL.

snl_seasons: The seasons dataframe contains details of each season from 1975 to 2020.

## Tidy Data (as needed)

Lets look what dataframes we have.

```{r}
head(snl_actors)
```

We can remove the url column for our analysis.
```{r}
snl_actors <- snl_actors %>% select(-url)
str(snl_actors)
```
Lets look at the casts dataframe:

```{r}
head(snl_casts)
```
```{r}
summary(snl_casts)
```

Finally we take a look at the seasons data.

```{r}
head(snl_seasons)
```
```{r}
summary(snl_seasons)
```

## Join Data

Now we will join the dataframes:

```{r}
snl_castsseasons <- left_join(snl_casts, snl_seasons, by = "sid")

snl_data <- left_join(snl_castsseasons, snl_actors, "aid")

glimpse(snl_data)
```

## Data Exploration

We can see the distribution of gender for the cast member who have been Weekend Update Anchors.
```{r}
snl_data %>%
  filter(update_anchor == TRUE) %>%
  ggplot(aes(x = gender, fill = gender)) +
  geom_bar() +
  xlab("Gender") +
  ylab("Count") +
  ggtitle("Gender Distribution for Weekend Update") +
  scale_fill_brewer(palette = "Set1")
```
We can see from the above graph that SNL has cast more Male cast members over female cast members as the Weekend Update Anchor.


It will be interesting to see the gender distribution of cast members over the years.
```{r}
snl_data %>%
  group_by(year, gender) %>%
  summarise(count = n()) %>%
  ggplot(aes(x = year, y = count, fill = gender)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(x = "Year", y = "Count", fill = "Gender") +
  ggtitle("Gender Distribution by Year") +
  scale_fill_brewer(palette = "Set1") +
  theme_minimal()
```
The percentage of female cast members has increased over the years. However, SNL still tends to cast more male actors than female.
