---
title: "Rail Road Employee Observations"
author: "Steven McKeen"
description: "Learning more about rail road stations across North America"
date: "02/28/2022"
format:
  html:
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
  - challenge_1
  - railroads
  - faostat
  - Steven
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Challenge Overview

Today's challenge is to

1)  read in a dataset, and

2)  describe the dataset using both words and any supporting information (e.g., tables, etc)

## Read in the Data

Read in one (or more) of the following data sets, using the correct R package and command.

-   railroad_2012_clean_county.csv ⭐
-   birds.csv ⭐⭐
-   FAOstat\*.csv ⭐⭐
-   wild_bird_data.xlsx ⭐⭐⭐
-   StateCounty2012.xls ⭐⭐⭐⭐

Find the `_data` folder, located inside the `posts` folder. Then you can read in the data, using either one of the `readr` standard tidy read commands, or a specialized package such as `readxl`.

```{r}
setwd("~/GitHub/601_Spring_2023/")
library(readr)
railroad_2012_clean_county <- read_csv("posts/_data/railroad_2012_clean_county.csv")
View(railroad_2012_clean_county)

```

Add any comments or documentation as needed. More challenging data sets may require additional code chunks and documentation.

## Describing the data

The Railroads data set provides a unique break down by county of employees that work at the railroad. In total across all the states that are accounted for there are 255,432 employees.  That averages to 87 employees across the 2930 county that were included in the census.

```{r}
#| label: summary
total_employees <- select(railroad_2012_clean_county, total_employees)
min(total_employees)
max(total_employees)
all_workers <- sum(total_employees)
all_workers
all_workers/2930


```
