---
title: "Understanding the FAOSTAT Country Codes" 
author: "Sue-Ellen Duffy" 
desription: "Demonstrated some of the differences in FAOSTAT Country Codes" 
date: "02/24/2023" 
format:
  html:
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
  - challenge_2
  - Sue-Ellen Duffy
  - FAOSTAT
editor: 
  markdown: 
    wrap: 72
---

```{r}
#| label: setup
#| warning: false
library(tidyverse)
library(summarytools)
library(readxl)
knitr::opts_chunk$set(echo = TRUE)
```

## FAOSTAT data

This data set is simply the Country Profiles for the Food and
Agriculture Organization Corporate Statistical Database (FAOSTAT).

In this set of data, each column is describing the same data, but vary
in who is describing or using that data. The United Nations Terminology
Database, the Statistics Division of the United Nations Secretariat, and
the International Organization for Standardization each have different
ways of coding the same countries, so this database helps us understand
which country or region or group of countries is being described in
other FAO data sets.

```{r}
#Read in data and rename FAOSTAT_country_groups as groups
data <- read_csv("_data/FAOSTAT_country_groups.csv")
data
```

## Summarize the data

Looking through this quick summary we see that there are 277 Countries
in this data, that IS02 is missing data, that the M49 Code column is
characterized as being characters (though they are all numeric) and
while there are other data here, there are two charts of interests -
Country Group and Country.

IS02 is "missing data" because of the code NA which is their 2-alpha
code for Nambia.

M49 Code column is characterized as being characters even though they
are all numeric.

The Country Group tibble shows us the 10 Country Groups containing the
most Countries.

The Country tibble shows us the 10 Countries that were categorized into
different Country Groups the most.

```{r}
dfSummary(data)
```

## How are the codes different?

Take the United States of America for example. When filtering for
"United States of America" we come out with 8 different rows of data.
The Country Code, M49 Code, IS02 Code and IS03 Codes while unique to
their specifics are unchanged for the 8 rows. The difference here, and
why we get 8 different rows for the United States of America, is that
their Country Group Code and Country Group are different. The Country
Group Code is simply the number associated with the Country Group. It
appears that Country Group is a categorical code, listing the USA as
being part of the Americas, High-income economies, North and Central
America, Annex I countries, etc.

The Country Group would allow for quick and categorical data analysis,
such as analyzing the countries by economics (high-income economies and
low-income economies) or by region (Northern and Central America to the
Americas.

```{r}
USA <- filter(data, `Country` == "United States of America")
USA
```
## The "World"

The one Country Group that contains all of the countries is "World",
which consists of 277 Countries. If the goal was to analyze all the
countries at once, the filter should be set to "World".

```{r}
world <- filter(data, `Country Group` == "World")
world
```

## Anything Else Interesting?

The M49 Codes with the suffix ".01" are characterized as being
"unspecified (population)". I'm not entirely sure what that means, so it
could be interesting to understand this further. Here is one example:

```{r}
data %>%
  filter(`M49 Code` == "155.01") %>%
select("Country Group", "Country", "M49 Code")
```
