---
title: "Challenge 8 "
author: "Sue-Ellen Duffy"
description: "Joining Data"
date: "05/22/2023"
format:
  html:
    toc: true
    code-copy: true
    code-tools: true
categories:
  - challenge_8
  - snl
  - Sue-Ellen_Duffy

---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)
library(ggplot2)
library(dplyr)
library(tidyr)
library(readr)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```


# Reading in the data
This dataset will look at SNL actor/actress gender over time. 

```{r}
actors <- read_csv("_data/snl_actors.csv")
casts <- read_csv("_data/snl_casts.csv")
seasons <- read_csv("_data/snl_seasons.csv")
```

```{r}
tibble(actors, 10)
```

```{r}
tibble(casts, 10)
```

```{r}
tibble(seasons, 10)
```


In order to join these datasets I will have to join actors and casts by "aid" or actor ID and then join that with seasons through "sid" or season ID. 

```{r}
unique(actors$ "type")
```
As a way to filter out the potential skewing of data through guests starring or crew (though those would also be interesting to look at later) I will filter these out of our datasets.
```{r}
cast_actors <- actors %>%
  left_join(casts, by = "aid") 
```

```{r}
all_data_prep <- cast_actors %>%
  left_join(seasons, by = "sid")
```

```{r}
all_data <- all_data_prep %>%
  group_by (sid, year, gender, featured) %>% 
  summarize(count = n())

all_data
```

# Visualization

Overtime SNL is adding more female actresses, but there is no equilibirium yet. T
```{r}
ggplot(all_data, aes(sid, count, col = gender))+ geom_line() + 
     coord_cartesian(ylim = c(0, 15))

all_data%>%
  filter(gender != "andy") %>%
ggplot(., aes(featured, stat = "count")) + geom_bar()


all_data%>%
  filter(gender != "andy") %>%
ggplot(., aes(x = year, y = featured, stat = "count", )) + geom_violin()  + facet_wrap (~gender,)
```