---
title: "Challenge 2"
author: "Teresa Lardo"
description: "Grouping a dataset on hotel bookings and providing summary statistics"
date: "02/22/2023"
format:
  html:
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
  - challenge_2
  - Teresa Lardo
  - hotel_bookings
editor: 
  markdown: 
    wrap: 72
---

## Read in the Data

For this challenge, I'll read in the csv file on hotel bookings.

```{r}
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
library(readr)
hotel_bookings <- read_csv("_data/hotel_bookings.csv", show_col_types = FALSE)
```
#### First 6 Rows
```{r}
head(hotel_bookings)
```

#### Last 6 Rows

```{r}
tail(hotel_bookings)
```


## Describe the data

#### Dimensions
```{r}
#Find out the dimensions of the tibble
dim(hotel_bookings)
```

#### Unique Hotels

```{r}
#Find out which hotels are listed in the hotel column
unique(hotel_bookings$hotel)
```


The cases in this dataset are just under 120,000 individual bookings at two hotels
(Resort Hotel and City Hotel) over the time period July 2015 through
August 2017. There are 32 variables in this dataset, including:

\- meal packages,

\- how far in advance the booking was created (lead time),

\- average daily rate,

\- whether a booking was canceled or not,

\- type of room reserved/assigned, and

\- how many adults, children, and babies are associated with each
booking.

Because each variable ought to have a column, we can look at all the
variables in this dataset by using the `colnames()` function.

```{r}
colnames(hotel_bookings)
```

The "is_canceled" and "reservation_status" columns clue us in to the
fact that the cases are bookings *made* but not necessarily actual 
*stays* at the hotels, as some customers canceled or no-showed. This
indicates that the data was most likely gathered from the booking
systems of the hotels themselves.

```{r}
# Show unique values under is_canceled column
unique(hotel_bookings$is_canceled)

# Show unique values under reservation_status column
unique(hotel_bookings$reservation_status)
```


## Grouped Summary Statistics

Because this dataset has so many variables, there are many potential
ways to group data. We could group our data by hotel, by bookings that
turned into actual *hotel stays* (where reservation_status is 'Check-Out'), by bookings that include children or
babies (where the children or babies column is not '0'), by group
bookings (where market_segment is 'groups'), by bookings with a full
deposit (where deposit_type is 'Non Refund'), by bookings for which the
reserved and assign room types differ (where reserved_room_type is not
the same as the value in assigned_room_type), by bookings with different
types of meal packages, etc.

### When are the hotels in demand?

Let's look at average lead time (number of days between booking and
anticipated arrival date) to get clues into when during the year each
hotel is more likely to be booked far in advanced. Let's also filter out bookings that were cancelled. 

```{r}
library(dplyr)
hotel_bookings %>%
  filter(reservation_status == "Check-Out") %>%
  group_by(arrival_date_month, hotel) %>%
  select(lead_time) %>%
  summarize_all(mean) %>%
  ungroup() %>%
  arrange(desc(lead_time))
```

Reservations for September at Resort Hotel have the highest average lead
time, with July reservations at City Hotel following behind. The lowest average lead times for City Hotel are in January, February, and Dcember. The lowest average lead times for Resort Hotel are in November, February, and January.

## Central Tendency and Dispersion

Despite the 32 different variables in this dataset, only 13 columns can
be directly used to get means, medians, and modes. Other columns may
appear to be numeric but answer Yes/No questions with 0 or 1, or list
numbers referring to dates (such as week of the year or day of the
month), which would not benefit from being averaged.

### Hotel Stays

I will filter for bookings that actually turned into hotel stays (where
the reservation status is "Check-Out" as opposed to "No show" or
"Canceled").

```{r}
#Create a vector containing the indices of the columns for summary statistics
stat_cols <- c(3, 8:12, 18, 19, 22, 26, 28:30)

#Load summarytools package to us descr()
library(summarytools)

hotel_stays <- hotel_bookings %>% 
  filter(reservation_status == "Check-Out")

descr(hotel_stays[stat_cols])

```



### Cancelled Bookings

To contrast with the summary statistics of the bookings that turned into
hotel stays, let's filter for only *cancelled* bookings at the two
hotels.

```{r}
canceled_bookings <- hotel_bookings %>%
  filter(is_canceled == 1) 

descr(canceled_bookings[stat_cols])
```


I notice the mean & median values for **ADR** is different in the summary stats for hotel stays vs. cancelled bookings. The max ADR value in cancelled bookings is 5400 - possibly a reason that reservation was cancelled. 

I also notice that the required car parking spaces variable seems to only have values of zero for the canceled bookings subset. This suggests that information about parking spaces is not collected until the check-in process when customers physically arrive to the hotels. 