---
title: "Thrishul Challenge 3"
author: "Meredith Rolfe"
desription: ""
date: "08/16/2022"
format:
  html:
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
  - challenge_2
  - railroads
  - faostat
  - hotel_bookings
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Challenge Overview

Today's challenge is to

1)  read in a data set, and describe the data using both words and any supporting information (e.g., tables, etc)
2)  provide summary statistics for different interesting groups within the data, and interpret those statistics

## Read in the Data

Read in one (or more) of the following data sets, available in the `posts/_data` folder, using the correct R package and command.

-   railroad\*.csv or StateCounty2012.xls ⭐
-   FAOstat\*.csv or birds.csv ⭐⭐⭐
-   hotel_bookings.csv ⭐⭐⭐⭐

```{r}
df<- read.csv(file = "_data/hotel_bookings.csv")
```

Add any comments or documentation as needed. More challenging data may require additional code chunks and documentation.

## Describe the data

Using a combination of words and results of R commands, can you provide a high level description of the data? Describe as efficiently as possible where/how the data was (likely) gathered, indicate the cases and variables (both the interpretation and any details you deem useful to the reader to fully understand your chosen data).

```{r}
#| label: summary
str(data)
num_observations <- nrow(df)
num_variables <- ncol(df)
cat("Dataset Size: ", num_observations, " observations, ", num_variables, " variables\n")


# histogram of is_canceled
hist(df$is_canceled)

#scatterplot of stays_in_weekend_nights, stays_in_week_nights
plot(df$stays_in_weekend_nights, df$stays_in_week_nights)

# boxplot of arrival_date_week_number
boxplot(df$arrival_date_week_number, )
summary(data)


```

## Provide Grouped Summary Statistics

Conduct some exploratory data analysis, using dplyr commands such as `group_by()`, `select()`, `filter()`, and `summarise()`. Find the central tendency (mean, median, mode) and dispersion (standard deviation, mix/max/quantile) for different subgroups within the data set.

```{r}
# Load necessary libraries
library(dplyr)
library(ggplot2)

# Read data
hotel_data <- df
```

```{r}
data <- select(df,-where(is.character))
cor(data)

```




### Group by City hotel and mean, median, and mode of lead_time

```{r}

# Group by hotel type and calculate mean, median, and mode of lead_time
hotel_lead_time_city <- hotel_data %>%
  group_by(hotel = 'City Hotel') %>%
  summarise(
    mean_lead_time = mean(lead_time),
    median_lead_time = median(lead_time),
    mode_lead_time = lead_time[which.max(table(lead_time))]
  )
print(hotel_lead_time_city)
```

### Group by Resort hotel and mean, median, and mode of lead_time

```{r}
# Group by hotel type and calculate mean, median, and mode of lead_time
hotel_lead_time_resort <- hotel_data %>%
  group_by(hotel = 'Resort Hotel') %>%
  summarise(
    mean_lead_time = mean(lead_time),
    median_lead_time = median(lead_time),
    mode_lead_time = lead_time[which.max(table(lead_time))]
  )
print(hotel_lead_time_resort)
```

### only reservations from Portugal and standard deviation of adr

```{r}
# Filter for only reservations from Portugal and calculate standard deviation of adr
portugal_adr_sd <- hotel_data %>%
  filter(country == "PRT") %>%
  summarise(adr_sd = sd(adr))
print(portugal_adr_sd)
```

### min, max, and 25th, 50th, and 75th percentiles of total_of_special_requests

```{r}
# Calculate min, max, and 25th, 50th, and 75th percentiles of total_of_special_requests
special_req_summary <- hotel_data %>%
  summarise(
    min_special_req = min(total_of_special_requests),
    max_special_req = max(total_of_special_requests),
    q25_special_req = quantile(total_of_special_requests, 0.25),
    q50_special_req = quantile(total_of_special_requests, 0.5),
    q75_special_req = quantile(total_of_special_requests, 0.75))
print(special_req_summary)


```
