---
title: "Challenge 3"
author: "Meredith Rolfe"
desription: "Tidy Data: Pivoting"
date: "08/17/2022"
format:
  html:
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
  - challenge_3
  - animal_weights
  - eggs
  - australian_marriage
  - usa_households
  - sce_labor
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Challenge Overview

Today's challenge is to:

1.  read in a data set, and describe the data set using both words and any supporting information (e.g., tables, etc)
2.  identify what needs to be done to tidy the current data
3.  anticipate the shape of pivoted data
4.  pivot the data into tidy format using `pivot_longer`

## Read in data

Read in one (or more) of the following datasets, using the correct R package and command.

-   animal_weights.csv ⭐
-   eggs_tidy.csv ⭐⭐ or organiceggpoultry.xls ⭐⭐⭐
-   australian_marriage\*.xls ⭐⭐⭐
-   USA Households\*.xlsx ⭐⭐⭐⭐
-   sce_labor_chart_data_public.xlsx 🌟🌟🌟🌟🌟

```{r}
aus_marriage <- readxl::read_excel("_data/australian_marriage_law_postal_survey_2017_-_response_final.xls", sheet="Table 2", skip=7)

aus_marriage <- aus_marriage[1:172,] #remove footnotes

aus_marriage$division <- NA #initialize division variable

aus_marriage$division[1:47] <- "New South Wales"
aus_marriage$division[51:87] <- "Victoria"
aus_marriage$division[91:120] <- "Queensland"
aus_marriage$division[124:134] <- "South Australia"
aus_marriage$division[157:161] <- "Tasmania"
aus_marriage$division[165:166] <- "Northern Territory"
aus_marriage$division[170:171] <- "Australian Capital Territory"

aus_marriage <- aus_marriage %>%
  filter(!is.na(division)) %>% #remove any invalid divisions
    rename(x1 = 1, x2 = 2, x3 = 3, x4 = 4, x5 = 5, x6 = 6, x7 = 7, x8 = 8, x9 = 9, x10 = 10, x11 = 11, x12 = 12, x13 = 13, x14 = 14, x15 = 15, x16 = 16) %>%
  mutate(x1 = gsub("\\(c)", "", x1)) %>%
  mutate(x1 = gsub("\\(d)", "", x1)) %>%
  mutate(x1 = gsub("\\(e)", "", x1)) #eliminate footnote letters in district name

head(aus_marriage)

```

### Briefly describe the data

Describe the data, and be sure to comment on why you are planning to pivot it to make it "tidy"

## Anticipate the End Result

The first step in pivoting the data is to try to come up with a concrete vision of what the end product *should* look like - that way you will know whether or not your pivoting was successful.

One easy way to do this is to think about the dimensions of your current data (tibble, dataframe, or matrix), and then calculate what the dimensions of the pivoted data should be.

Suppose you have a dataset with $n$ rows and $k$ variables. In our example, 3 of the variables are used to identify a case, so you will be pivoting $k-3$ variables into a longer format where the $k-3$ variable names will move into the `names_to` variable and the current values in each of those columns will move into the `values_to` variable. Therefore, we would expect $n * (k-3)$ rows in the pivoted dataframe!

### Example: find current and future data dimensions

Lets see if this works with a simple example.

```{r}
#| tbl-cap: Example

df<-tibble(country = rep(c("Mexico", "USA", "France"),2),
           year = rep(c(1980,1990), 3), 
           trade = rep(c("NAFTA", "NAFTA", "EU"),2),
           outgoing = rnorm(6, mean=1000, sd=500),
           incoming = rlogis(6, location=1000, 
                             scale = 400))
df

#existing rows/cases
nrow(df)

#existing columns/cases
ncol(df)

#expected rows/cases
nrow(df) * (ncol(df)-3)

# expected columns 
3 + 2
```

Or simple example has $n = 6$ rows and $k - 3 = 2$ variables being pivoted, so we expect a new dataframe to have $n * 2 = 12$ rows x $3 + 2 = 5$ columns.

### Challenge: Describe the final dimensions

This results in a new dataframe with 5 columns, where two columns identify the case (division and district) and the remaining three columns represent the pivoted variables (count, percent, and response type).

To calculate the total number of columns in the new dataframe, the text uses the formula k = 2 + 2 + 1, where k represents the total number of columns. The first two terms, 2 + 2, represent the number of columns used to identify the case, and the last term, 1, represents the response type variable. Applying this formula yields a total of 5 columns.

However, the text does not provide enough information to determine the exact number of columns in the resulting dataframe. Instead, we can infer that the dataframe will have at least 5 columns, based on the description of the data transformation process.

To determine the exact number of columns in the new dataframe, we would need additional information about the original dataframe, such as the number of rows, the names and types of the original columns, and the specific method used to pivot the data. Without this information, we cannot calculate the exact number of columns in the resulting dataframe.

```{r}
aus_marriage2 <- aus_marriage %>%
    select(x1, x2, x3, x4, x5, division) %>%
  rename(district = x1, yes1count=x2, yes1percent=x3, no2count=x4, no2percent=x5)

#existing rows/cases
nrow(aus_marriage2)

```

```{r}
#existing columns/cases
ncol(aus_marriage2)
#expected rows/cases
nrow(aus_marriage2) * (ncol(aus_marriage2)-2)
# expected columns 
2 + 2 + 1
```


## Pivot the Data

Now we will pivot the data, and compare our pivoted data dimensions to the dimensions calculated above as a "sanity" check.

### Example

```{r}
#| tbl-cap: Pivoted Example

df<-pivot_longer(df, col = c(outgoing, incoming),
                 names_to="trade_direction",
                 values_to = "trade_value")
df
```

Yes, once it is pivoted long, our resulting data are $12x5$ - exactly what we expected!

### Challenge: Pivot the Chosen Data
The dataset contains information on responses to a single type of question (yes or no) in a specific district within a particular division in Australia. Each row in the dataset represents a unique observation of this type of response in the district, and each column represents a distinct variable of interest. The dataset satisfies the principles of tidy data, meaning that each cell in the dataset contains a single value, and there are no duplicate or missing values. In total, the tidy dataset contains 268 rows and 5 columns.

```{r}
aus_marriage_tidy <- aus_marriage2 %>%
    pivot_longer(-c('district', 'division'), names_to=c("resp", ".value"), names_sep = "\\d")

aus_marriage_tidy

```


When performing data pivoting in R, it can be useful to rename the new columns created by the pivot operation based on a consistent pattern. This helps R identify which columns correspond to which values in the resulting table. By renaming the columns in this way, we can make it easier to work with the pivoted data and apply additional data manipulations or analysis. Essentially, renaming the columns in a consistent way after pivoting allows for clearer identification and organization of the resulting data.

