---
title: "Final Project - Assignment 1"
author: "Thrishul"
desription: "To acquire a better understanding of what sales could look like for the next two years."
date: "05/07/2023"
format:
  html:
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
  - Final Project 1
---
# Introduction
This data science project aims to explore and analyze a comprehensive dataset of job titles, salaries, employment types, company sizes and locations within the field of data science. With a focus on understanding current trends and patterns, this project seeks to provide insights into the current state of the data science job market. Through analyzing the distribution of job titles, seniority levels, salaries, and employment types, we can gain an understanding of the most common roles, skillsets, and experience required in the field. Additionally, examining the location and size of companies can provide insights into which regions are leading in the data science industry and what size of companies are more likely to hire data scientists. This project aims to provide valuable insights to job seekers, employers, and researchers in the data science field.

## Dataset Description

work_year: This column indicates the year in which the salary was paid to the employee. It allows us to analyze trends in salaries over time and compare salaries between different years.

experience_level: This column indicates the experience level of the employee in the job during the year. It allows us to analyze how experience level affects salaries and identify common experience levels for different job titles.

employment_type: This column indicates the type of employment for the role, whether it is Contract, Freelance, Full-Time, or Part-Time. It allows us to analyze the prevalence of different employment types in the data science field.

job_title: This column indicates the role worked in during the year. It allows us to analyze the most common job titles in the data science field and identify trends in job titles over time.

salary: This column indicates the total gross salary amount paid to the employee in the specified currency. It allows us to analyze salary ranges, identify outliers, and compare salaries between different job titles and experience levels.

salary_currency: This column indicates the currency of the salary paid as an ISO 4217 currency code. It allows us to convert salaries to a common currency for analysis and comparison.

salaryinusd: This column indicates the salary in USD. It allows us to compare salaries in a common currency and analyze the impact of currency exchange rates on salaries.

employee_residence: This column indicates the employee's primary country of residence during the work year as an ISO 3166 country code. It allows us to analyze the geographic distribution of employees and identify common countries of residence for different job titles and experience levels.

remote_ratio: This column indicates the overall amount of work done remotely by the employee during the year. It allows us to analyze the prevalence of remote work in the data science field and identify common remote work ratios for different job titles and experience levels.

company_location: This column indicates the country of the employer's main office or contracting branch. It allows us to analyze the geographic distribution of companies and identify common countries where data science jobs are located.

company_size: This column indicates the median number of people that worked for the company during the year. It allows us to analyze the size distribution of companies and identify common company sizes for different job titles and experience levels.




## Loading Necessary Packages


```{r}
pacman :: p_load(pacman,stats, dplyr, knitr, ggplot2, plotly, psych, gridExtra,
                 waffle, emojifont,tidyr, tidytext, wordcloud, GGally, viridis,
                tidyverse, rnaturalearth, rnaturalearthdata) 

options(warn=-1) # warnings will be suppressed and will not be displayed in the console or any output
```
# Data Reading

```{r}
df <- read.table("_data/ds_salaries.csv", sep = ",", header = T)

# Nr. of instances
cat("Number of instances:", nrow(df))

# View data
head(df, 10)

```





```{r}
# Inspect data
check <- function(data) {
  l <- list()
  columns <- names(data)
  for (col in columns) {
    instances <- sum(!is.na(data[[col]]))
    dtypes <- class(data[[col]])
    unique <- length(unique(data[[col]]))
    sum_null <- sum(is.na(data[[col]]))
    duplicates <- sum(duplicated(data))
    l[[length(l) + 1]] <- c(col, dtypes, instances, unique, sum_null, duplicates)
  }
  data_check <- as.data.frame(do.call(rbind, l))
  names(data_check) <- c("column", "dtype", "instances", "unique", "sum_null", "duplicates")
  return(data_check)
}

check(df)

```

# Inspecting the Data
```{r}
str(df)
describe(df)

```

# Conclusion

The Data Science Job Salaries Dataset contains information on salaries paid to employees working in the data science field. The dataset contains 11 columns that provide information on the work year, experience level, employment type, job title, salary amount, salary currency, salary in USD, employee residence, remote work ratio, company location, and company size.

With this dataset, I can analyze and explore different aspects of data science job salaries such as:

Trends in salaries over time by analyzing the work year column.

The impact of experience level on salaries by analyzing the experience_level column.

The prevalence of different types of employment in the data science field by analyzing the employment_type column.

The most common job titles in the data science field by analyzing the job_title column.

Salary ranges and outliers in the data science field by analyzing the salary column.

The geographic distribution of employees and companies in the data science field by analyzing the employee_residence and company_location columns.

The impact of remote work on salaries and prevalence of remote work in the data science field by analyzing the remote_ratio column.

The size distribution of companies in the data science field by analyzing the company_size column.

Overall, this dataset allows me to gain insights into different aspects of data science job salaries and explore trends and patterns in the data.



















