---
title: "Challenge 6"
author: "Thrishul"
desription: "Visualizing Time and Relationships"
date: "04/05/2023"
format:
  html:
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
  - challenge_6
  - fed rates
  - debt
---

```{r}
library(tidyverse)
library(ggplot2)
library(readxl)
library(lubridate)
library(here)

source(here("posts","umass_colors.R"))
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

# Challenge Overview

Today’s challenge is to:

create at least one graph including time (evolution)
try to make them “publication” ready (optional)
Explain why you choose the specific graph type
Create at least one graph depicting part-whole or flow relationships
try to make them “publication” ready (optional)
Explain why you choose the specific graph type


## Dataset Description
The dataset spans from July 1954 to March 2017 and comprises daily macroeconomic indicators associated with the effective federal funds rate. This rate refers to the interest rate at which banks lend money to each other to fulfill mandated reserve requirements. In addition to the date column, the dataset contains seven variables, four of which pertain to the federal funds rate (target, upper target, lower target, and effective). The remaining three variables represent macroeconomic indicators such as inflation, GDP change, and unemployment rate.
```{r}
fed_rates_vars<-here("posts","_data","FedFundsRate.csv") %>% 
  read_csv(n_max = 1,
           col_names = NULL)%>%
  select(-c(X1:X3))%>%
  unlist(.)

names(fed_rates_vars) <-c("fed_target", "fed_target_upper",
                         "fed_target_lower", "fed_effective",
                         "gdp_ch", "unemploy", "inflation")
      
fed_rates_orig<-here("posts","_data","FedFundsRate.csv") %>% 
  read_csv(skip=1,
           col_names = c("Year", "Month", "Day", 
                         names(fed_rates_vars)))

fed_rates<-fed_rates_orig%>%
  mutate(date = make_date(Year, Month, Day))%>%
  select(-c(Year, Month, Day))

fed_rates <- fed_rates%>%
  pivot_longer(cols=-date, 
               names_to = "variable",
               values_to = "value")
```

Next, we aimed to visualize the evolution of the macroeconomic indicators and federal funds rate over time, while carefully handling the issue of missing data.

```{r}
fed_rates%>%
  filter(str_starts(variable, "fed"))%>%
ggplot(., aes(x=date, y=value, color=variable))+
  geom_point(size=0)+
  geom_line()+
  scale_y_continuous(labels = scales::label_percent(scale = 1))
```

By analyzing the plotted data, it becomes apparent how closely the effective rate follows the target rate, and how the Federal Reserve adjusted its approach to target rate setting in response to the 2009 financial crisis. To gain additional insights, we explored the relationship between the effective rate and one of the other macroeconomic indicators.

```{r}
fed_rates%>%
  filter(variable%in%c("fed_effective", "gdp_ch", 
                       "unemploy", "inflation"))%>%
ggplot(., aes(x=date, y=value, color=variable))+
  geom_point(size=0)+
  geom_line()+
  facet_grid(rows = vars(variable))
```

```{r}
year_unemploy <- fed_rates %>%
  pivot_wider(names_from = variable, values_from = value) %>%
  mutate(year=year(date)) %>%
  group_by(year) %>%
  summarise(median_rate=median(unemploy)/100) %>%
  ungroup()
year_unemploy
```

```{r}
year_unemploy %>%
  ggplot(aes(year,median_rate))+
  geom_line()
```

```{r}
year_unemploy %>%
  filter(year<=1981) %>%
  ggplot(aes(year,median_rate))+
  geom_line()+
  scale_y_continuous(labels=scales::percent_format(),limits=c(0,.1))+
  scale_x_continuous(breaks=seq(1955,1980,5))
```

```{r}
  labs(x="year",y="median unemployment rate")
```