---
title: "Challenge 1"
author: "Tim Shores"
desription: "Reading in data and creating a post"
date: "02/25/2023"
format:
  html:
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
  - challenge_1
  - wildbirds
---
```{r}
#| label: setup
#| warning: false
#| message: false

my_packages <- c("dplyr", "magrittr", "readxl", "summarytools") # create vector of packages
invisible(lapply(my_packages, require, character.only = TRUE)) # load multiple packages

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## My content

Challenge 1 pits us unfortunate students up against two tasks:

1)  read in a dataset, and

2)  describe the dataset using both words and any supporting information (e.g., tables, etc)

```{r}
wbirds <- read_xlsx("../posts/_data/wild_bird_data.xlsx",skip=1) # read in data
showing <- 42
```

I chose to read in wild_bird_data.xlsx, because wild birds are *exciting*. The wild bird file includes two rows of headings. I skipped the first row to tidy up my tibble, but it tells us that the data source is "Figure 1 of Nee et al." That led me to this May 1991 Nature article: [The relationship between abundance and body size in British birds](https://www.nature.com/articles/351312a0)

The wild bird data set includes **`r nrow(wbirds)`** observations under **`r ncol(wbirds)`** variables. Here are the top **`r showing`** observations of British birds -- not just *any* wild birds:

```{r}
print(wbirds, n=showing)
wb_col1_null <- wbirds %>% select(names(wbirds)[1]) %>% n_distinct(.)
wb_col2_null <- wbirds %>% select(names(wbirds)[2]) %>% n_distinct(.)
```

The *`r names(wbirds)[1]`* variable has `r nrow(wbirds) - wb_col1_null` null values.

The *`r names(wbirds)[2]`* variable has `r nrow(wbirds) - wb_col2_null` null values.

Here's the summary:

```{r}
summary(wbirds)
```