---
title: "Challenge_6"
author: "Xinyang Mao"
description: "Visualizing Time and Relationships"
date: "04/19/2023"
format:
  html:
    toc: true
    code-copy: true
    code-tools: true
categories:
  - challenge_6
  - hotel_bookings
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)
library(ggplot2)
library(lubridate)
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Challenge Overview

Today's challenge is to:

1)  read in a data set, and describe the data set using both words and any supporting information (e.g., tables, etc)
2)  tidy data (as needed, including sanity checks)
3)  mutate variables as needed (including sanity checks)
4)  create at least one graph including time (evolution)
   - try to make them "publication" ready (optional)
   - Explain why you choose the specific graph type
5)  Create at least one graph depicting part-whole or flow relationships
   - try to make them "publication" ready (optional)
   - Explain why you choose the specific graph type

[R Graph Gallery](https://r-graph-gallery.com/) is a good starting point for thinking about what information is conveyed in standard graph types, and includes example R code.

(be sure to only include the category tags for the data you use!)

## Read in data

Read in one (or more) of the following datasets, using the correct R package and command.
  - hotel_bookings ⭐⭐⭐⭐

```{r}
hotel_raw <- read.csv("_data/hotel_bookings.csv")
print(summarytools::dfSummary(hotel_raw,
                        varnumbers = FALSE,
                        plain.ascii  = FALSE, 
                        style        = "grid", 
                        graph.magnif = 0.70, 
                        valid.col    = FALSE),
      method = 'render',
      table.classes = 'table-condensed')
```

We can see there are lots of attributes in these dataset.
In this challenge I mainly focus on the number of guests and time.
```{r}
str(hotel_raw)
```

## Tidy Data

Firstly,I transfer string type of month to int for combine year and month,thus more convenient to observe the relationship between the total number of guests of each month and time.
```{r}
hotel_raw<- na.omit(hotel_raw)
hotel <- hotel_raw%>%
  mutate(Month = match(arrival_date_month,month.name))
head(hotel$Month)
```
Combine year and month used to display the graph later.
```{r}

hotel1 <- hotel%>%
  mutate(date = str_c(arrival_date_year,Month,sep = "-"))
```
check the result,it's what I want!
```{r}
head(hotel1$date)
```
Calculate the total number of guests for each day.
```{r}
hotel_month <- hotel1%>%
  mutate(total_guests = rowSums(select(.,adults,children,babies),na.rm = TRUE))
summary(hotel_month$total_guests)

```
## Time Dependent Visualization
According to the `date` column I processed previously,I grouped the rows by month and use `summarise()` to calculate the total guests number of each month from 2015 to 2017.
```{r}
hotel_month<-hotel_month%>%
  group_by(date) %>%
  summarise(month_total = sum(total_guests)) %>%
  ungroup() %>%
  mutate(month_total = as.integer(month_total))
hotel_month
```
Then draw a gragh for observing,we can see it was floated a lot,and the least crowded months are November 2015,Jan of both 2016 and 2017.
```{r}
ggplot(hotel_month, aes(x = date, y = month_total,group = 1)) +
  geom_line() +
  theme(axis.text.x = element_text(size = 5 ))
```

## Visualizing Part-Whole Relationships
In this part I'd like exploring the relationship between months and different type of guests.
Firstly I I calculated the total monthly amount of different kinds of guests separately.
```{r}
hotel_type<-hotel1%>%
  group_by(Month) %>%
  summarise(month_child = sum(children),month_babies = sum(babies),month_adults = sum(adults)) %>%
  ungroup() %>%
  mutate(month_child = as.integer(month_child),month_babies = as.integer(month_babies),month_adults = as.integer(month_adults),Month = month(Month))


hotel_type
```
Then I drawed lines for each kind of guests:adults,child and babies.
I find that there are many more adult guests than children and babies!And almost all year round at a very high level except from November to January it may means people will travel less in winter.
For children,although the number of children is very low,but what is interesting is that during the summer holidays (June to September) there is a significant increase.It can be seen that children usually go out in the summer holidays.
As for babies, the number of hotel reservations has remained at a very low number.
```{r}
ggplot(hotel_type, aes(Month,month_child,group = 1,col = "child")) +
  geom_line() +
  geom_line(aes(Month,month_babies, group = 1,col = "babies")) +
  geom_line(aes(Month,month_adults, group = 1,col ="adults")) +
  scale_x_continuous(breaks = seq(1,12,1)) +
  scale_color_manual(values = c("red","blue","green"))
  theme(axis.text.x = element_text(size = 2 ))
```






