---
title: "Challenge_5"
author: "Xinyang Mao"
description: "Introduction to Visualization"
date: "04/04/2023"
format:
  html:
    toc: true
    code-copy: true
    code-tools: true
categories:
  - challenge_5
  - air_bnb
 
---

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)
library(ggplot2)

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Challenge Overview

Today's challenge is to:

1)  read in a data set, and describe the data set using both words and any supporting information (e.g., tables, etc)
2)  tidy data (as needed, including sanity checks)
3)  mutate variables as needed (including sanity checks)
4)  create at least two univariate visualizations
   - try to make them "publication" ready
   - Explain why you choose the specific graph type
5)  Create at least one bivariate visualization
   - try to make them "publication" ready
   - Explain why you choose the specific graph type

[R Graph Gallery](https://r-graph-gallery.com/) is a good starting point for thinking about what information is conveyed in standard graph types, and includes example R code.

(be sure to only include the category tags for the data you use!)

## Read in data

Read in one (or more) of the following datasets, using the correct R package and command.

-   AB_NYC_2019.csv ⭐⭐⭐
In this challenge I chose the airbnb dataset.
```{r}
airbnbData <- read.csv("_data/AB_NYC_2019.csv")
head(airbnbData)
```

### Briefly describe the data
```{r}
print(summarytools::dfSummary(airbnbData,
                        varnumbers = FALSE,
                        plain.ascii  = FALSE, 
                        style        = "grid", 
                        graph.magnif = 0.70, 
                        valid.col    = FALSE),
      method = 'render',
      table.classes = 'table-condensed')
```
We can see there are many attributes in this dataset,we can choose several of them to analysis.
```{r}
str(airbnbData)
```

## Univariate Visualizations

Firstly,I'm curious about the number of reviews.
We can use `ggplot()` and `geom_histogram()` to show it's basic character.We can see that the number of reviews mainly concerntrate between 0 and 500.So we can try to tidy them.
```{r}
ggplot(airbnbData,aes(number_of_reviews)) +
  geom_histogram()

```


```{r}
range(airbnbData$number_of_reviews)

```
`filter()` can help to filter the outliers.
```{r}
airbnb_filtered <- airbnbData %>%
  filter(number_of_reviews >0 & number_of_reviews < 500)
airbnb_filtered %>%
  ggplot(aes(number_of_reviews))+
  geom_histogram()
```


## Bivariate Visualization(s)
I also wonder what's the relationship between `number_of_reviews` and `price`.We can observe the `price` attribute firstly.

```{r}
ggplot(airbnb_filtered,aes(price))+
  geom_histogram()
```
```{r}
range(airbnb_filtered$price)
```
Also,we can filter the outliers of `price`,to keep the all of valid value in the range of 0~3000.
```{r}
airbnb_filtered <- airbnb_filtered %>%
  filter(price >0 & price < 3000)
airbnb_filtered %>%
  ggplot(aes(price))+
  geom_histogram()
```
Now we can explore the relationship between `number_of_reviews` and `price`,let's have a look.
We can use `geom_point()` to draw the Scatter chart.
```{r}
airbnb_filtered %>%
  ggplot(aes(number_of_reviews,price))+
  geom_point()
```
It seems that the number of evaluations is negatively correlated with the price.
