---
title: "Challenge1"
author: "Xinyang Mao"
desription: "Reading in data and describing"
date: "02/22/2023"
format:
  html:
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
  - hw1
  - challenge1
  - my name
  - dataset
  - ggplot2
---
## Import package
```{r}
#| label: setup
#| warning: false

library(tidyverse)

knitr::opts_chunk$set(echo = TRUE)
```


## Read dataset
```{r}

data <- read_csv("_data/railroad_2012_clean_county.csv")
```
## Shows the first 6 rows of the data frame
This function defaults to printing the first 6 row.
```{r}
head(data)
```
## Shows the dimensions of the data frame by row and column
We can see there are 2930 rows and 3 colums in this dataset.
```{r}
dim(data)
```
## Shows the structure of the data frame
The structure of the data set also tells us the number of rowsand columns, but it provides even more information. It tells us the column names, the class of each column (what kind of data is stored in it), and the first few observations of each variable.
```{r}
str(data)
```
## Shows the name of each column in the data frame
This function prints a vector of the column names.
```{r}
colnames(data)
```
## Provides summary statistics on the columns of the data frame
The summary provides descriptive statistics including the min, max, mean, median, and quartiles of each column. For example, we can see in this data set that the average number of total employees is 87.18.
```{r}
summary(data)
```
## Shows a spreadsheet-like display of the entire data frame
This window provides vertical and horizontal scroll bars to browse the entire data set. It looks exactly like an Excel spreadsheet.
```{r}
View(data)
```

